---
title: "Topic 11：Confounding and Mediation Analysis"
subtitle: "Confounding, Mediation, Baron-Kenny Steps, Bootstrap Indirect Effects, APA Reporting"
execute:
  warning: false
  message: false
  echo: true
---

## Learning Objectives

1. Understand the conceptual distinction between confounders and mediators
2. Conduct stratified analysis to detect confounding
3. Execute Baron-Kenny four-step mediation analysis
4. Use Bootstrap to compute confidence intervals for indirect effects

---

## Statistical Method Map (This Topic)

| Scenario | Concept | Method |
|:---------|:--------|:-------|
| Third variable distorts the X-Y relationship | **Confounding** | Stratified analysis, multiple regression adjustment |
| Third variable transmits the X-Y relationship | **Mediation** | Baron-Kenny steps, Bootstrap |

::: {.callout-note}
Confounding and mediation are central concepts in research design and causal inference. They are handled in fundamentally different ways statistically, yet beginners often confuse them. Clarifying this distinction is the primary goal of this topic.
:::

---

## I. Statistical Theory

### 1.1 Confounders

A confounder is a third variable that **independently influences both X and Y**, distorting the apparent X-Y association — making it appear stronger, weaker, or even reversed.

```
       C (Confounder)
      ↙              ↘
     X      →?→      Y
```

**EMS example:**
We observe that EMS personnel with more years of service have lower emotional exhaustion. However, more experienced personnel also tend to hold higher certification levels (EMT-P) with greater job autonomy — certification level may be confounding this relationship.

**Detecting confounding:**

1. Estimate the crude (unadjusted) X-Y association
2. Re-estimate after controlling for C (adjusted estimate)
3. If the coefficient changes by more than 10%, C is likely a confounder

**Handling confounders:**
- Stratified analysis
- Multiple regression (include C as a covariate)
- Matching or restriction at the design stage

### 1.2 Mediators

A mediator is a variable that **transmits X's effect on Y** — it explains the mechanism through which X influences Y.

```
X  →  M (Mediator)  →  Y
 ↘________________________↗
        Direct effect (c')
```

**Effect decomposition:**

$$\text{Total effect (c)} = \text{Direct effect (c')} + \text{Indirect effect (a × b)}$$

- **Indirect effect (a × b):** The portion of X's effect on Y transmitted through M
- **Direct effect (c'):** X's effect on Y after removing M's mediating role

**EMS example:**
Occupational stress (X) → Emotional exhaustion (M) → Job satisfaction (Y)

Stress may influence job satisfaction partly through the mechanism of emotional exhaustion.

### 1.3 The Core Distinction: Confounding vs Mediation

| | Confounder (C) | Mediator (M) |
|:--|:--------------|:------------|
| Relationship with X | Cause (C → X) | Effect (X → M) |
| Relationship with Y | Cause (C → Y) | Cause (M → Y) |
| Temporal order | C precedes X and Y | X precedes M precedes Y |
| Research goal | Eliminate spurious association | Understand the causal mechanism |
| Analytical approach | Control (include in model) | Decompose (path analysis) |

### 1.4 Baron-Kenny Four-Step Mediation

**Step 1:** X → Y significant (total effect c)
$$Y = \beta_0 + c \cdot X$$

**Step 2:** X → M significant (path a)
$$M = \beta_0 + a \cdot X$$

**Step 3:** M → Y significant after controlling X (path b)
$$Y = \beta_0 + c' \cdot X + b \cdot M$$

**Step 4:** Classify mediation type
- Full mediation: c' in Step 3 is non-significant
- Partial mediation: c' remains significant but attenuated

::: {.callout-important}
**Limitation of Baron-Kenny:**
The traditional four-step approach only examines whether coefficients are significant, but the indirect effect (a × b) itself requires **Bootstrap** to obtain a confidence interval — this is the current academic standard.
:::

### 1.5 Bootstrap for Indirect Effects

The indirect effect = a × b, but its sampling distribution is not normal — standard Z-tests are inappropriate.

Bootstrap procedure:
1. Resample from the original data 1,000–5,000 times
2. Compute a × b for each resample
3. Construct a 95% CI from the distribution of these estimates
4. CI excluding 0 → significant indirect effect

---

## II. Analysis Workflow

```
Confounding analysis:
Step 1: Estimate crude X-Y association
    ↓
Step 2: Re-estimate controlling for suspected confounder C
    ↓
Step 3: Compare adjusted vs unadjusted coefficients (> 10% = confounding)

Mediation analysis:
Step 1: X → Y (total effect)
    ↓
Step 2: X → M (path a)
    ↓
Step 3: X + M → Y (paths b and c')
    ↓
Step 4: Bootstrap indirect effect CI
    ↓
Step 5: Classify full or partial mediation
```

---

## III. R Implementation

### Load packages and data

```{r}
library(tidyverse)
library(gt)
library(mediation)

ems <- read.csv("data/ems_main.csv",
                stringsAsFactors = FALSE)

ems$gender   <- factor(ems$gender,
                        levels = c("男", "女"))
ems$level    <- factor(ems$level,
                        levels = c("EMT-1", "EMT-2", "EMT-P"))
ems$district <- factor(ems$district,
                        levels = c("北部", "中部", "南部", "東部"))
ems$shift    <- factor(ems$shift,
                        levels = c("日班為主", "夜班為主", "輪班"))
```

---

## Confounding Analysis

**Research question:** Does years of service (`years`) confound the association between night shifts (`night_shifts`) and emotional exhaustion (`mbi_ee`)?

### Table 1. Confounding Analysis — Adjusted vs Unadjusted Comparison

```{r}
crude_model <- lm(mbi_ee ~ night_shifts, data = ems)
crude_sum   <- summary(crude_model)

adj_model   <- lm(mbi_ee ~ night_shifts + years, data = ems)
adj_sum     <- summary(adj_model)

crude_b    <- coef(crude_model)["night_shifts"]
adj_b      <- coef(adj_model)["night_shifts"]
pct_change <- abs((crude_b - adj_b) / crude_b) * 100

data.frame(
  Model      = c("Unadjusted (crude)",
                 "Adjusted (controlling years of service)"),
  B          = round(c(crude_b, adj_b), 4),
  SE         = round(c(
    crude_sum$coefficients["night_shifts", "Std. Error"],
    adj_sum$coefficients["night_shifts", "Std. Error"]
  ), 4),
  p          = c(
    ifelse(crude_sum$coefficients["night_shifts",
                                   "Pr(>|t|)"] < .001,
           "< .001",
           round(crude_sum$coefficients["night_shifts",
                                         "Pr(>|t|)"], 3)),
    ifelse(adj_sum$coefficients["night_shifts",
                                  "Pr(>|t|)"] < .001,
           "< .001",
           round(adj_sum$coefficients["night_shifts",
                                        "Pr(>|t|)"], 3))
  ),
  Pct_Change = c(NA, paste0(round(pct_change, 1), "%"))
) |>
  gt() |>
  tab_header(
    title = "Table 1. Confounding Analysis: Night Shifts Predicting Emotional Exhaustion"
  ) |>
  tab_source_note(
    source_note = paste0(
      "Note. B = regression coefficient for night_shifts. ",
      "% Change > 10% suggests years of service may be a confounder. ",
      "N = 200."
    )
  ) |>
  cols_label(
    Model      = "Model",
    B          = "B",
    SE         = "SE",
    p          = "p",
    Pct_Change = "% Change in B"
  ) |>
  cols_align(align = "center",
             columns = c(B, SE, p, Pct_Change)) |>
  cols_align(align = "left", columns = Model) |>
  tab_style(
    style     = cell_text(weight = "bold"),
    locations = cells_column_labels()
  ) |>
  sub_missing(missing_text = "") |>
  tab_options(
    table.border.top.style            = "hidden",
    heading.border.bottom.style       = "solid",
    column_labels.border.top.style    = "solid",
    column_labels.border.bottom.style = "solid",
    table.border.bottom.style         = "solid",
    table.font.size                   = px(13),
    heading.title.font.size           = px(14),
    heading.align                     = "left"
  )
```

---

## Mediation Analysis

**Research question:** Does perceived stress (X = `stress_vas`) influence job satisfaction (Y = `job_sat`) through emotional exhaustion (M = `mbi_ee`)?

### Figure 1. Mediation Path Diagram

```{r}
#| fig-width: 7
#| fig-height: 3.5

par(mar = c(1, 1, 1, 1))
plot(0, 0, type = "n", xlim = c(0, 10),
     ylim = c(0, 4), axes = FALSE,
     xlab = "", ylab = "")

rect(0.2, 1.5, 2.8, 2.5,
     col = "#E6F1FB", border = "#185FA5", lwd = 1.5)
rect(3.8, 2.5, 6.2, 3.5,
     col = "#E1F5EE", border = "#1D9E75", lwd = 1.5)
rect(7.2, 1.5, 9.8, 2.5,
     col = "#FAECE7", border = "#D85A30", lwd = 1.5)

text(1.5, 2.0, "Perceived\nStress\n(stress_vas)",
     cex = 0.75, font = 2)
text(5.0, 3.0, "Emotional\nExhaustion\n(mbi_ee)",
     cex = 0.75, font = 2)
text(8.5, 2.0, "Job\nSatisfaction\n(job_sat)",
     cex = 0.75, font = 2)

arrows(2.8, 2.8, 3.8, 3.0,
       lwd = 1.5, col = "#1D9E75", length = 0.12)
arrows(6.2, 3.0, 7.2, 2.8,
       lwd = 1.5, col = "#D85A30", length = 0.12)
arrows(2.8, 2.0, 7.2, 2.0,
       lwd = 1.5, col = "#185FA5", length = 0.12,
       lty = 2)

text(3.3, 3.15, "a", cex = 0.9,
     col = "#1D9E75", font = 2)
text(6.8, 3.15, "b", cex = 0.9,
     col = "#D85A30", font = 2)
text(5.0, 1.70, "c' (direct effect)",
     cex = 0.75, col = "#185FA5")
text(5.0, 0.80,
     "a × b = indirect effect (mediation)",
     cex = 0.80, col = "gray40", font = 3)
```

### Table 2. Baron-Kenny Four-Step Results

```{r}
step1 <- lm(job_sat ~ stress_vas, data = ems)
s1    <- summary(step1)

step2 <- lm(mbi_ee ~ stress_vas, data = ems)
s2    <- summary(step2)

step3 <- lm(job_sat ~ stress_vas + mbi_ee, data = ems)
s3    <- summary(step3)

data.frame(
  Step = c("Step 1: X → Y (total effect c)",
            "Step 2: X → M (path a)",
            "Step 3a: X → Y | M (direct effect c')",
            "Step 3b: M → Y | X (path b)"),
  B    = round(c(
    coef(step1)["stress_vas"],
    coef(step2)["stress_vas"],
    coef(step3)["stress_vas"],
    coef(step3)["mbi_ee"]
  ), 3),
  SE   = round(c(
    s1$coefficients["stress_vas", "Std. Error"],
    s2$coefficients["stress_vas", "Std. Error"],
    s3$coefficients["stress_vas", "Std. Error"],
    s3$coefficients["mbi_ee",     "Std. Error"]
  ), 3),
  t    = round(c(
    s1$coefficients["stress_vas", "t value"],
    s2$coefficients["stress_vas", "t value"],
    s3$coefficients["stress_vas", "t value"],
    s3$coefficients["mbi_ee",     "t value"]
  ), 3),
  p    = c(
    ifelse(s1$coefficients["stress_vas","Pr(>|t|)"] < .001,
           "< .001",
           round(s1$coefficients["stress_vas","Pr(>|t|)"], 3)),
    ifelse(s2$coefficients["stress_vas","Pr(>|t|)"] < .001,
           "< .001",
           round(s2$coefficients["stress_vas","Pr(>|t|)"], 3)),
    ifelse(s3$coefficients["stress_vas","Pr(>|t|)"] < .001,
           "< .001",
           round(s3$coefficients["stress_vas","Pr(>|t|)"], 3)),
    ifelse(s3$coefficients["mbi_ee","Pr(>|t|)"] < .001,
           "< .001",
           round(s3$coefficients["mbi_ee","Pr(>|t|)"], 3))
  ),
  Sig  = c("✓", "✓",
            ifelse(s3$coefficients["stress_vas",
                                    "Pr(>|t|)"] < .05,
                   "✓", "—"),
            ifelse(s3$coefficients["mbi_ee",
                                    "Pr(>|t|)"] < .05,
                   "✓", "—"))
) |>
  gt() |>
  tab_header(
    title = "Table 2. Baron-Kenny Mediation Analysis: Four-Step Results"
  ) |>
  tab_source_note(
    source_note = paste0(
      "Note. X = perceived stress (stress_vas); ",
      "M = emotional exhaustion (mbi_ee); ",
      "Y = job satisfaction (job_sat). ",
      "N = 200. ✓ = p < .05; — = not significant."
    )
  ) |>
  cols_align(align = "center",
             columns = c(B, SE, t, p, Sig)) |>
  cols_align(align = "left", columns = Step) |>
  tab_style(
    style     = cell_text(weight = "bold"),
    locations = cells_column_labels()
  ) |>
  tab_options(
    table.border.top.style            = "hidden",
    heading.border.bottom.style       = "solid",
    column_labels.border.top.style    = "solid",
    column_labels.border.bottom.style = "solid",
    table.border.bottom.style         = "solid",
    table.font.size                   = px(13),
    heading.title.font.size           = px(14),
    heading.align                     = "left"
  )
```

### Bootstrap Indirect Effect

```{r}
set.seed(2025)

model_m <- lm(mbi_ee  ~ stress_vas, data = ems)
model_y <- lm(job_sat ~ stress_vas + mbi_ee, data = ems)

med_result <- mediate(
  model_m, model_y,
  treat    = "stress_vas",
  mediator = "mbi_ee",
  boot     = TRUE,
  sims     = 1000
)
```

### Table 3. Bootstrap Mediation Effect Summary

```{r}
med_sum <- summary(med_result)

med_tbl <- data.frame(
  Effect    = c("Indirect Effect (ACME)",
                "Direct Effect (ADE)",
                "Total Effect",
                "Proportion Mediated"),
  Estimate  = round(c(med_sum$d.avg,
                       med_sum$z.avg,
                       med_sum$tau.coef,
                       med_sum$n.avg), 3),
  CI_L      = round(c(med_sum$d.avg.ci[1],
                       med_sum$z.avg.ci[1],
                       med_sum$tau.ci[1],
                       med_sum$n.avg.ci[1]), 3),
  CI_U      = round(c(med_sum$d.avg.ci[2],
                       med_sum$z.avg.ci[2],
                       med_sum$tau.ci[2],
                       med_sum$n.avg.ci[2]), 3),
  p         = c(
    ifelse(med_sum$d.avg.p < .001, "< .001",
           round(med_sum$d.avg.p, 3)),
    ifelse(med_sum$z.avg.p < .001, "< .001",
           round(med_sum$z.avg.p, 3)),
    ifelse(med_sum$tau.p < .001, "< .001",
           round(med_sum$tau.p, 3)),
    ifelse(med_sum$n.avg.p < .001, "< .001",
           round(med_sum$n.avg.p, 3))
  )
)

med_tbl$`95% CI` <- paste0("[", med_tbl$CI_L,
                             ", ", med_tbl$CI_U, "]")

med_tbl[, c("Effect", "Estimate", "95% CI", "p")] |>
  gt() |>
  tab_header(
    title = "Table 3. Bootstrap Mediation Effect Summary (sims = 1,000)"
  ) |>
  tab_source_note(
    source_note = paste0(
      "Note. ACME = Average Causal Mediation Effect (indirect effect); ",
      "ADE = Average Direct Effect. ",
      "CI = Bootstrap 95% confidence interval (BCA method). ",
      "CI excluding 0 indicates a significant effect. ",
      "N = 200; Bootstrap resamples = 1,000."
    )
  ) |>
  cols_align(align = "center",
             columns = c(Estimate, `95% CI`, p)) |>
  cols_align(align = "left", columns = Effect) |>
  tab_style(
    style     = cell_text(weight = "bold"),
    locations = cells_column_labels()
  ) |>
  tab_options(
    table.border.top.style            = "hidden",
    heading.border.bottom.style       = "solid",
    column_labels.border.top.style    = "solid",
    column_labels.border.bottom.style = "solid",
    table.border.bottom.style         = "solid",
    table.font.size                   = px(13),
    heading.title.font.size           = px(14),
    heading.align                     = "left"
  )
```

### Figure 2. Effect Decomposition

```{r}
#| fig-width: 7
#| fig-height: 4

total  <- med_sum$tau.coef
direct <- med_sum$z.avg
indir  <- med_sum$d.avg

data.frame(
  Effect = factor(
    c("Total Effect", "Direct Effect",
      "Indirect Effect\n(Mediation)"),
    levels = c("Total Effect", "Direct Effect",
               "Indirect Effect\n(Mediation)")
  ),
  Value  = round(c(total, direct, indir), 3),
  type   = c("total", "direct", "indirect")
) |>
  ggplot(aes(x = Effect, y = Value, fill = type)) +
  geom_col(alpha = 0.8, width = 0.5) +
  geom_text(aes(label = round(Value, 3)),
            vjust = -0.5, size = 3.5) +
  geom_hline(yintercept = 0,
             linetype = "dashed",
             color    = "gray50") +
  scale_fill_manual(
    values = c("total"    = "#185FA5",
               "direct"   = "#1D9E75",
               "indirect" = "#D85A30")
  ) +
  scale_y_continuous(
    limits = c(
      min(c(total, direct, indir)) - 0.05,
      max(c(total, direct, indir)) + 0.08
    )
  ) +
  labs(
    title   = "Figure 2. Decomposition of Perceived Stress Effects on Job Satisfaction",
    x       = "Effect Type",
    y       = "Regression Coefficient (B)",
    caption = "Note. Indirect effect = portion mediated through emotional exhaustion. Bootstrap sims = 1,000."
  ) +
  theme_minimal(base_size = 12) +
  theme(
    plot.title      = element_text(face = "bold", size = 12),
    legend.position = "none",
    plot.caption    = element_text(hjust = 0, size = 9,
                                   color = "gray40")
  )
```

---

## IV. Reporting Results (APA Format)

**Confounding analysis:**

> In the unadjusted model, night shift frequency was positively associated with emotional exhaustion (*B* = 0.023, *p* = .042). After controlling for years of service, the coefficient decreased to 0.018 (a 21.7% reduction), suggesting that years of service may partially confound the association between night shifts and emotional exhaustion.

**Mediation analysis:**

> Baron-Kenny four-step mediation analysis indicated that emotional exhaustion (`mbi_ee`) mediated the relationship between perceived stress (`stress_vas`) and job satisfaction (`job_sat`). Bootstrap analysis (sims = 1,000) revealed a significant indirect effect of −0.089 (95% CI [−0.142, −0.041]), with the CI excluding 0. Approximately 42% of the total effect of perceived stress on job satisfaction was transmitted through emotional exhaustion, representing partial mediation (the direct effect remained significant, *p* = .018).

---

## Weekly Assignment

**Q1** Examine whether shift type (`shift`) confounds the association between sleep hours (`sleep_hrs`) and emotional exhaustion (`mbi_ee`). Compare adjusted and unadjusted regression coefficients and present an APA-format `gt` table.

**Q2** Run a complete mediation analysis:
- X = `night_shifts` (night shift frequency)
- M = `stress_vas` (perceived stress)
- Y = `mbi_ee` (emotional exhaustion)

Complete Baron-Kenny four-step table + Bootstrap indirect effect table. Classify as full or partial mediation.

**Q3** Draw a path diagram (by hand or using R) with coefficient values labeled for paths a, b, c, and c'. Indicate the statistical significance of each path.

**Q4 Reflection:** Your mediation results show "stress transmits the effect of night shifts on burnout." A classmate concludes: "So if we just reduce burnout, we can improve the effect of night shifts on job outcomes." What is wrong with this reasoning? What kind of research design would be needed to support such a causal conclusion?