---
title: "Topic 7：One-Way ANOVA"
subtitle: "F-statistic, Post-Hoc Comparisons, Effect Size, APA Reporting"
execute:
  warning: false
  message: false
  echo: true
---

## Learning Objectives

1. Understand the statistical logic of ANOVA and the meaning of the F-statistic
2. Correctly execute one-way ANOVA and post-hoc comparisons
3. Calculate and interpret effect size (η²)
4. Produce APA-format result tables and reporting sentences

---

## Statistical Method Map (This Topic)

| X (Predictor) | Y (Outcome) | Method | Nonparametric Alternative |
|:-------------|:-----------|:-------|:------------------------|
| Categorical (2 groups) | Continuous | t-test (Topic 6) | Mann-Whitney U |
| **Categorical (≥ 3 groups)** | **Continuous** | **One-way ANOVA** | **Kruskal-Wallis (Topic 8)** |

::: {.callout-note}
**Why not run multiple t-tests instead?**
With 3 groups, you would need 3 pairwise t-tests. At α = .05 each, the familywise Type I error rate inflates to 1 − (0.95)³ = 14.3%. ANOVA tests all groups simultaneously in a single test, controlling the overall error rate at α = .05.
:::

---

## I. Statistical Theory

### 1.1 The Core Logic of ANOVA

ANOVA partitions the total variability in the data (Total SS) into two sources:

$$SS_{\text{Total}} = SS_{\text{Between}} + SS_{\text{Within}}$$

- **Between-group variance (SS_Between):** Differences between group means and the grand mean → reflects the effect of X on Y
- **Within-group variance (SS_Within):** Differences between individual observations and their group mean → reflects individual variability and measurement error

**Intuition:** If between-group differences are much larger than within-group differences, there is good reason to believe the group means are genuinely different.

### 1.2 The F-Statistic

$$F = \frac{MS_{\text{Between}}}{MS_{\text{Within}}} = \frac{SS_{\text{Between}} / (k-1)}{SS_{\text{Within}} / (N-k)}$$

- k = number of groups; N = total sample size
- MS = Mean Square = SS ÷ degrees of freedom
- Larger F → between-group differences are large relative to within-group noise → stronger evidence against H₀

**H₀:** All group population means are equal (μ₁ = μ₂ = μ₃ = ...)

::: {.callout-important}
**ANOVA's F-test only tells you "at least one group differs." It does not tell you which groups differ.**
Post-hoc comparisons are required to locate the specific differences.
:::

### 1.3 Assumptions

| Assumption | Description | How to Check |
|:-----------|:-----------|:------------|
| Normality | Y approximately normal within each group | Shapiro-Wilk; Q-Q plot |
| Homogeneity of variance | Equal variances across groups | Levene's test |
| Independence | Observations are mutually independent | Design-level judgment |

::: {.callout-tip}
**If Levene's test p < .05 (unequal variances):**
Use Welch's ANOVA (`oneway.test()`), analogous to the Welch correction for independent t-tests. Follow with Games-Howell post-hoc comparisons.
:::

### 1.4 Post-Hoc Tests

After a significant ANOVA, post-hoc comparisons identify which specific pairs of groups differ:

| Method | When to Use | Conservativeness |
|:-------|:-----------|:----------------|
| Tukey HSD | All pairwise comparisons, equal group sizes | Moderate |
| Bonferroni | Few planned comparisons | More conservative |
| Scheffé | All possible contrasts, unequal group sizes | Most conservative |
| Games-Howell | Unequal variances | Moderate |

**Course recommendation:** Default to Tukey HSD; use Games-Howell when variances are unequal.

### 1.5 Effect Size: η² (Eta-Squared)

$$\eta^2 = \frac{SS_{\text{Between}}}{SS_{\text{Total}}}$$

Represents the proportion of Y's total variance explained by group membership.

| η² range | Strength |
|:---------|:--------|
| .01–.05 | Small effect |
| .06–.13 | Medium effect |
| ≥ .14 | Large effect |

::: {.callout-note}
η² tends to be slightly inflated with larger samples. A less biased version is **ω² (Omega-squared)**, available from the `effectsize` package.
:::

---

## II. Analysis Workflow

```
Step 1: Descriptive statistics + boxplot (n, Mean, SD per group)
    ↓
Step 2: Assumption checks
    ├── Normality: Shapiro-Wilk
    └── Homogeneity of variance: Levene's test
    ↓
Step 3: Run ANOVA
    ├── Equal variances → aov() + TukeyHSD()
    └── Unequal variances → oneway.test() + Games-Howell
    ↓
Step 4: Compute effect size (η²)
    ↓
Step 5: APA-format report
```

---

## III. R Implementation

### Load packages and data

```{r}
library(tidyverse)
library(gt)
library(car)
library(rstatix)

ems <- read.csv("data/ems_main.csv",
                stringsAsFactors = FALSE)

ems$gender   <- factor(ems$gender,
                        levels = c("男", "女"))
ems$level    <- factor(ems$level,
                        levels = c("EMT-1", "EMT-2", "EMT-P"))
ems$district <- factor(ems$district,
                        levels = c("北部", "中部", "南部", "東部"))
ems$shift    <- factor(ems$shift,
                        levels = c("日班為主", "夜班為主", "輪班"))
```

**Research question:** Do the three EMT certification levels (EMT-1, EMT-2, EMT-P) differ significantly in mean on-scene response time?

### Figure 1. Distribution by Group

```{r}
#| fig-width: 7
#| fig-height: 4.5

ggplot(ems, aes(x = level, y = response_time,
                fill = level)) +
  geom_boxplot(alpha = 0.6,
               outlier.shape = 21,
               outlier.fill  = "white",
               outlier.size  = 2) +
  geom_jitter(width = 0.15, alpha = 0.25, size = 1.2) +
  stat_summary(fun  = mean, geom = "point",
               shape = 23, size = 3,
               fill = "white", color = "black") +
  scale_fill_manual(
    values = c("EMT-1" = "#F5C4B3",
               "EMT-2" = "#9FE1CB",
               "EMT-P" = "#85B7EB")
  ) +
  labs(
    title   = "Figure 1. Response Time Distribution by EMT Level",
    x       = "EMT Certification Level",
    y       = "Response Time (minutes)",
    caption = "Note. Center line = median; diamond = mean; box = IQR; points = individual observations. N = 200."
  ) +
  theme_minimal(base_size = 12) +
  theme(
    plot.title      = element_text(face = "bold", size = 12),
    legend.position = "none",
    plot.caption    = element_text(hjust = 0, size = 9,
                                   color = "gray40")
  )
```

### Table 1. Descriptive Statistics by Group

```{r}
ems |>
  dplyr::group_by(level) |>
  dplyr::summarise(
    n      = n(),
    Mean   = round(mean(response_time), 2),
    SD     = round(sd(response_time), 2),
    Median = round(median(response_time), 2),
    Q1     = round(quantile(response_time, 0.25), 2),
    Q3     = round(quantile(response_time, 0.75), 2),
    .groups = "drop"
  ) |>
  dplyr::rename(Level = level) |>
  gt() |>
  tab_header(
    title = "Table 1. Response Time Descriptive Statistics by EMT Level"
  ) |>
  tab_spanner(label = "Central Tendency",
              columns = c(Mean, Median)) |>
  tab_spanner(label = "Quartiles",
              columns = c(Q1, Q3)) |>
  tab_source_note(
    source_note = "Note. Units: minutes."
  ) |>
  cols_align(align = "center", columns = -Level) |>
  cols_align(align = "left",   columns = Level) |>
  tab_style(
    style     = cell_text(weight = "bold"),
    locations = cells_column_labels()
  ) |>
  tab_style(
    style     = cell_text(weight = "bold"),
    locations = cells_column_spanners()
  ) |>
  tab_options(
    table.border.top.style            = "hidden",
    heading.border.bottom.style       = "solid",
    column_labels.border.top.style    = "solid",
    column_labels.border.bottom.style = "solid",
    table.border.bottom.style         = "solid",
    table.font.size                   = px(13),
    heading.title.font.size           = px(14),
    heading.align                     = "left"
  )
```

### Table 2. Shapiro-Wilk Normality Test by Group

```{r}
sw_results <- ems |>
  dplyr::group_by(level) |>
  dplyr::summarise(
    W        = round(shapiro.test(response_time)$statistic, 3),
    p        = round(shapiro.test(response_time)$p.value, 3),
    Decision = ifelse(shapiro.test(response_time)$p.value > .05,
                      "Fail to reject normality",
                      "Reject normality"),
    .groups  = "drop"
  ) |>
  dplyr::rename(Group = level)

sw_results |>
  gt() |>
  tab_header(
    title = "Table 2. Shapiro-Wilk Normality Test by EMT Level"
  ) |>
  tab_source_note(
    source_note = "Note. p > .05 indicates failure to reject the normality assumption."
  ) |>
  cols_align(align = "center", columns = c(W, p)) |>
  cols_align(align = "left",
             columns = c(Group, Decision)) |>
  tab_style(
    style     = cell_text(weight = "bold"),
    locations = cells_column_labels()
  ) |>
  tab_options(
    table.border.top.style            = "hidden",
    heading.border.bottom.style       = "solid",
    column_labels.border.top.style    = "solid",
    column_labels.border.bottom.style = "solid",
    table.border.bottom.style         = "solid",
    table.font.size                   = px(13),
    heading.title.font.size           = px(14),
    heading.align                     = "left"
  )
```

### Table 3. Levene's Test for Homogeneity of Variance

```{r}
levene_result <- leveneTest(response_time ~ level,
                             data = ems)

data.frame(
  Test     = "Levene's Test",
  F        = round(levene_result$`F value`[1], 3),
  df1      = levene_result$Df[1],
  df2      = levene_result$Df[2],
  p        = round(levene_result$`Pr(>F)`[1], 3),
  Decision = ifelse(levene_result$`Pr(>F)`[1] > .05,
                    "Variances equal (standard ANOVA appropriate)",
                    "Variances unequal (Welch ANOVA recommended)")
) |>
  gt() |>
  tab_header(
    title = "Table 3. Levene's Test for Homogeneity of Variance"
  ) |>
  tab_source_note(
    source_note = "Note. p > .05 indicates homogeneous variances."
  ) |>
  cols_align(align = "center",
             columns = c(F, df1, df2, p)) |>
  cols_align(align = "left",
             columns = c(Test, Decision)) |>
  tab_style(
    style     = cell_text(weight = "bold"),
    locations = cells_column_labels()
  ) |>
  tab_options(
    table.border.top.style            = "hidden",
    heading.border.bottom.style       = "solid",
    column_labels.border.top.style    = "solid",
    column_labels.border.bottom.style = "solid",
    table.border.bottom.style         = "solid",
    table.font.size                   = px(13),
    heading.title.font.size           = px(14),
    heading.align                     = "left"
  )
```

### Table 4. One-Way ANOVA Summary Table

```{r}
aov_result  <- aov(response_time ~ level, data = ems)
aov_summary <- summary(aov_result)

ss_between <- aov_summary[[1]]$`Sum Sq`[1]
ss_within  <- aov_summary[[1]]$`Sum Sq`[2]
ss_total   <- ss_between + ss_within
df_between <- aov_summary[[1]]$Df[1]
df_within  <- aov_summary[[1]]$Df[2]
ms_between <- aov_summary[[1]]$`Mean Sq`[1]
ms_within  <- aov_summary[[1]]$`Mean Sq`[2]
f_val      <- aov_summary[[1]]$`F value`[1]
p_val      <- aov_summary[[1]]$`Pr(>F)`[1]
eta2       <- ss_between / ss_total

data.frame(
  Source = c("Between groups (Level)",
             "Within groups (Error)",
             "Total"),
  SS     = round(c(ss_between, ss_within, ss_total), 2),
  df     = c(df_between, df_within,
              df_between + df_within),
  MS     = c(round(ms_between, 2),
              round(ms_within, 2), NA),
  F      = c(round(f_val, 3), NA, NA),
  p      = c(ifelse(p_val < .001, "< .001",
                    round(p_val, 3)), NA, NA),
  eta2   = c(round(eta2, 3), NA, NA)
) |>
  gt() |>
  tab_header(
    title = "Table 4. One-Way ANOVA Summary: Response Time by EMT Level"
  ) |>
  tab_source_note(
    source_note = paste0(
      "Note. SS = sum of squares; MS = mean square; ",
      "η² = eta-squared effect size. ",
      "η² ≥ .14 = large effect (Cohen, 1988)."
    )
  ) |>
  cols_label(
    Source = "Source", SS = "SS", df = "df",
    MS = "MS", F = "F", p = "p", eta2 = "η²"
  ) |>
  cols_align(align = "center",
             columns = c(SS, df, MS, F, p, eta2)) |>
  cols_align(align = "left", columns = Source) |>
  tab_style(
    style     = cell_text(weight = "bold"),
    locations = cells_column_labels()
  ) |>
  tab_style(
    style     = cell_fill(color = "#F0F5FA"),
    locations = cells_body(rows = 3)
  ) |>
  sub_missing(missing_text = "") |>
  tab_options(
    table.border.top.style            = "hidden",
    heading.border.bottom.style       = "solid",
    column_labels.border.top.style    = "solid",
    column_labels.border.bottom.style = "solid",
    table.border.bottom.style         = "solid",
    table.font.size                   = px(13),
    heading.title.font.size           = px(14),
    heading.align                     = "left"
  )
```

### Table 5. Tukey HSD Post-Hoc Comparisons

```{r}
tukey_result <- TukeyHSD(aov_result)
tukey_df     <- as.data.frame(tukey_result$level)
tukey_df     <- tibble::rownames_to_column(tukey_df,
                                            "Comparison")

tukey_df |>
  dplyr::mutate(
    diff    = round(diff, 2),
    lwr     = round(lwr, 2),
    upr     = round(upr, 2),
    `p adj` = ifelse(`p adj` < .001, "< .001",
                     round(`p adj`, 3)),
    CI      = paste0("[", lwr, ", ", upr, "]"),
    Sig     = ifelse(as.numeric(
                gsub("< .001", "0", `p adj`)) < .05 |
                `p adj` == "< .001", "✓", "—")
  ) |>
  dplyr::select(Comparison, diff, CI, `p adj`, Sig) |>
  gt() |>
  tab_header(
    title = "Table 5. Tukey HSD Post-Hoc Comparisons"
  ) |>
  tab_source_note(
    source_note = paste0(
      "Note. diff = mean difference; CI = 95% confidence interval; ",
      "p adj = Tukey-adjusted p-value. ✓ = p < .05; — = not significant."
    )
  ) |>
  cols_label(
    Comparison = "Comparison",
    diff       = "Mean Diff (min)",
    CI         = "95% CI",
    `p adj`    = "p (adjusted)",
    Sig        = "Significant"
  ) |>
  cols_align(align = "center",
             columns = c(diff, CI, `p adj`, Sig)) |>
  cols_align(align = "left", columns = Comparison) |>
  tab_style(
    style     = cell_text(weight = "bold"),
    locations = cells_column_labels()
  ) |>
  tab_options(
    table.border.top.style            = "hidden",
    heading.border.bottom.style       = "solid",
    column_labels.border.top.style    = "solid",
    column_labels.border.bottom.style = "solid",
    table.border.bottom.style         = "solid",
    table.font.size                   = px(13),
    heading.title.font.size           = px(14),
    heading.align                     = "left"
  )
```

### Figure 2. Group Means with 95% CI

```{r}
#| fig-width: 6
#| fig-height: 4.5

ems |>
  dplyr::group_by(level) |>
  dplyr::summarise(
    mean = mean(response_time),
    ci_l = t.test(response_time)$conf.int[1],
    ci_u = t.test(response_time)$conf.int[2],
    .groups = "drop"
  ) |>
  ggplot(aes(x = level, y = mean, color = level)) +
  geom_point(size = 4) +
  geom_errorbar(aes(ymin = ci_l, ymax = ci_u),
                width = 0.15, linewidth = 0.8) +
  scale_color_manual(
    values = c("EMT-1" = "#D85A30",
               "EMT-2" = "#1D9E75",
               "EMT-P" = "#185FA5")
  ) +
  annotate("text", x = 1, y = 11.8,
           label = "a", size = 4, fontface = "bold") +
  annotate("text", x = 2, y = 10.4,
           label = "b", size = 4, fontface = "bold") +
  annotate("text", x = 3, y = 9.0,
           label = "c", size = 4, fontface = "bold") +
  labs(
    title   = "Figure 2. Mean Response Time and 95% CI by EMT Level",
    x       = "EMT Certification Level",
    y       = "Response Time (minutes)",
    caption = "Note. Error bars = 95% CI. Different letters (a, b, c) indicate significant pairwise differences (Tukey HSD, p < .05). N = 200."
  ) +
  theme_minimal(base_size = 12) +
  theme(
    plot.title      = element_text(face = "bold", size = 12),
    legend.position = "none",
    plot.caption    = element_text(hjust = 0, size = 9,
                                   color = "gray40")
  )
```

---

## IV. Reporting Results (APA Format)

**ANOVA main effect:**

> A one-way ANOVA revealed a significant effect of EMT certification level on on-scene response time, *F*(2, 197) = 18.43, *p* < .001, η² = .16 (large effect).

**Post-hoc comparisons:**

> Tukey HSD post-hoc comparisons showed that EMT-1 (*M* = 10.4 min) had significantly longer response times than both EMT-2 (*M* = 9.1 min, *p* < .001) and EMT-P (*M* = 7.8 min, *p* < .001). EMT-2 was also significantly slower than EMT-P (*p* = .012).

**Combined report:**

> A one-way ANOVA comparing response times across EMT certification levels was statistically significant, *F*(2, 197) = 18.43, *p* < .001, η² = .16. Tukey HSD post-hoc comparisons revealed significant differences between all three pairwise combinations (*p*s < .05), with response time decreasing progressively with higher certification level (see Figure 2).

---

## Weekly Assignment

**Q1** Compare `stress_vas` across the four districts (`district`). Run a complete analysis: descriptive statistics table, assumption checks, ANOVA summary table, and Tukey HSD post-hoc comparison table. All tables in APA `gt` format.

**Q2** If Levene's test indicates unequal variances, use `oneway.test()` for Welch ANOVA and `games_howell_test()` (rstatix package) for post-hoc comparisons. Report whether the conclusions change.

**Q3** Compute both η² and ω² for Q1. Explain the difference between the two measures and present the results in an APA-format `gt` table.

**Q4 Reflection:** Does a significant ANOVA (p < .05) mean "all groups are different from each other"? A classmate says "a significant F-test means every pair of groups differs." How would you respond? Use the post-hoc results to illustrate your answer.