---
title: "Topic 2：Descriptive Statistics & Epidemiological Measures"
subtitle: "Central Tendency, Dispersion, Epidemiological Rates, R Implementation"
execute:
  warning: false
  message: false
  echo: true
---

## Learning Objectives

1. Understand central tendency and dispersion in descriptive statistics
2. Select the appropriate measure: mean vs median
3. Calculate and interpret epidemiological measures (prevalence, RR, OR)
4. Perform complete descriptive statistics on `ems_main.csv` using R

---

## Statistical Method Map (This Topic)

| X (Predictor) | Y (Outcome) | Method |
|:-------------|:-----------|:-------|
| None (univariate) | **Continuous** | Mean, SD, Median, IQR |
| None (univariate) | **Categorical** | Frequency, %, rates |

::: {.callout-note}
Descriptive statistics do not distinguish X from Y — we are simply describing the distribution of a single variable. The X-Y framework begins in Topic 5 when we examine relationships between two variables.
:::

---

## I. Statistical Theory

### 1.1 Why Do We Need Descriptive Statistics?

`ems_main.csv` contains 200 observations and 27 variables. You cannot read through every individual value — you need to compress these 200 numbers into a small set of meaningful summaries.

Descriptive statistics answer two fundamental questions:

- **Where is the data centered?** → Central tendency (mean, median)
- **How spread out is the data?** → Dispersion (SD, IQR)

### 1.2 Central Tendency

#### Mean

$$\bar{x} = \frac{\sum_{i=1}^{n} x_i}{n}$$

**Critical weakness:** Highly sensitive to extreme values. A single unusually long response time can pull the mean substantially upward.

#### Median

Arrange all values from smallest to largest and take the middle value.

**Advantage:** Not affected by extreme values — reflects the "typical" case.

::: {.callout-important}
**Decision rule:**

- Histogram approximates a bell shape → Normal → Report **Mean ± SD**
- Histogram is skewed or has outliers → Report **Median (IQR)**
- In medical research, response times and length of stay are typically skewed → Median is usually preferred
:::

### 1.3 Dispersion

#### Standard Deviation (SD)

$$s = \sqrt{\frac{\sum_{i=1}^{n}(x_i - \bar{x})^2}{n-1}}$$

The average distance of each observation from the mean. The denominator uses **n-1** (Bessel's correction) to produce an unbiased estimate.

#### Interquartile Range (IQR)

$$IQR = Q_3 - Q_1$$

The spread of the middle 50% of data. Used together with the median when data are skewed.

### 1.4 Epidemiological Measures

#### Prevalence

$$P = \frac{\text{Number of cases}}{\text{Population at risk}}$$

**Cross-sectional concept:** proportion currently having a condition.

#### Relative Risk (RR)

$$RR = \frac{\text{Incidence rate in exposed group}}{\text{Incidence rate in unexposed group}}$$

Used in cohort studies. RR > 1 = exposure increases risk.

#### Odds Ratio (OR)

$$OR = \frac{ad}{bc}$$

Used in case-control studies and logistic regression (Topic 10).

::: {.callout-tip}
**RR vs OR:** When prevalence < 10%, OR ≈ RR. At higher prevalence, OR overestimates RR.
:::

---

## II. R Implementation

### Load packages and data

```{r}
library(tidyverse)
library(gt)
library(psych)

ems <- read.csv("data/ems_main.csv",
                stringsAsFactors = FALSE)

ems$gender   <- factor(ems$gender,
                        levels = c("男", "女"))
ems$level    <- factor(ems$level,
                        levels = c("EMT-1", "EMT-2", "EMT-P"))
ems$district <- factor(ems$district,
                        levels = c("北部", "中部", "南部", "東部"))
ems$shift    <- factor(ems$shift,
                        levels = c("日班為主", "夜班為主", "輪班"))
```

### Table 1. Descriptive Statistics for Continuous Variables

```{r}
desc_raw <- describe(
  ems |> dplyr::select(age, years, monthly_cases,
                        bmi, sbp, sleep_hrs,
                        stress_vas, mbi_ee, response_time)
)

data.frame(
  Variable = c("Age (years)", "Years of service",
               "Monthly calls", "BMI (kg/m²)",
               "Systolic BP (mmHg)", "Sleep hours (h)",
               "Stress VAS", "MBI-EE",
               "Response time (min)"),
  n        = as.integer(desc_raw$n),
  Mean     = round(desc_raw$mean, 2),
  SD       = round(desc_raw$sd, 2),
  Median   = round(desc_raw$median, 2),
  Min      = round(desc_raw$min, 2),
  Max      = round(desc_raw$max, 2),
  Skewness = round(desc_raw$skew, 2)
) |>
  gt() |>
  tab_header(
    title = "Table 1. Descriptive Statistics for Key Continuous Variables"
  ) |>
  tab_spanner(label = "Central Tendency",
              columns = c(Mean, Median)) |>
  tab_spanner(label = "Dispersion",
              columns = c(SD, Min, Max)) |>
  tab_source_note(
    source_note = "Note. N = 200. |Skewness| > 1 indicates notable skew."
  ) |>
  cols_align(align = "center", columns = -Variable) |>
  cols_align(align = "left",   columns = Variable) |>
  tab_style(
    style     = cell_text(weight = "bold"),
    locations = cells_column_labels()
  ) |>
  tab_style(
    style     = cell_text(weight = "bold"),
    locations = cells_column_spanners()
  ) |>
  tab_options(
    table.border.top.style            = "hidden",
    heading.border.bottom.style       = "solid",
    column_labels.border.top.style    = "solid",
    column_labels.border.bottom.style = "solid",
    table.border.bottom.style         = "solid",
    table.font.size                   = px(13),
    heading.title.font.size           = px(14),
    heading.align                     = "left"
  )
```

### Table 2. Response Time by EMT Level

```{r}
ems |>
  dplyr::group_by(level) |>
  dplyr::summarise(
    n      = n(),
    Mean   = round(mean(response_time), 2),
    SD     = round(sd(response_time), 2),
    Median = round(median(response_time), 2),
    Q1     = round(quantile(response_time, 0.25), 2),
    Q3     = round(quantile(response_time, 0.75), 2),
    .groups = "drop"
  ) |>
  dplyr::rename(Level = level) |>
  gt() |>
  tab_header(
    title = "Table 2. Response Time by EMT Certification Level"
  ) |>
  tab_spanner(label = "Central Tendency",
              columns = c(Mean, Median)) |>
  tab_spanner(label = "Quartiles",
              columns = c(Q1, Q3)) |>
  tab_source_note(
    source_note = "Note. Units: minutes. Q1 = 25th percentile; Q3 = 75th percentile."
  ) |>
  cols_align(align = "center", columns = -Level) |>
  cols_align(align = "left",   columns = Level) |>
  tab_style(
    style     = cell_text(weight = "bold"),
    locations = cells_column_labels()
  ) |>
  tab_style(
    style     = cell_text(weight = "bold"),
    locations = cells_column_spanners()
  ) |>
  tab_options(
    table.border.top.style            = "hidden",
    heading.border.bottom.style       = "solid",
    column_labels.border.top.style    = "solid",
    column_labels.border.bottom.style = "solid",
    table.border.bottom.style         = "solid",
    table.font.size                   = px(13),
    heading.title.font.size           = px(14),
    heading.align                     = "left"
  )
```

### Table 3. EMT Level Distribution

```{r}
ems |>
  dplyr::count(level) |>
  dplyr::mutate(
    Percentage = round(n / sum(n) * 100, 1),
    Cumulative = round(cumsum(Percentage), 1)
  ) |>
  dplyr::rename(Level = level) |>
  gt() |>
  tab_header(
    title = "Table 3. Distribution of EMT Certification Levels"
  ) |>
  tab_source_note(
    source_note = "Note. N = 200."
  ) |>
  cols_label(
    Level      = "Level",
    n          = "n",
    Percentage = "Percentage (%)",
    Cumulative = "Cumulative (%)"
  ) |>
  cols_align(align = "center",
             columns = c(n, Percentage, Cumulative)) |>
  cols_align(align = "left", columns = Level) |>
  tab_style(
    style     = cell_text(weight = "bold"),
    locations = cells_column_labels()
  ) |>
  tab_options(
    table.border.top.style            = "hidden",
    heading.border.bottom.style       = "solid",
    column_labels.border.top.style    = "solid",
    column_labels.border.bottom.style = "solid",
    table.border.bottom.style         = "solid",
    table.font.size                   = px(13),
    heading.title.font.size           = px(14),
    heading.align                     = "left"
  )
```

### Figure 1. Distribution of Response Time

```{r}
#| fig-width: 7
#| fig-height: 4.5

ggplot(ems, aes(x = response_time)) +
  geom_histogram(bins = 20,
                 fill = "#85B7EB",
                 color = "white") +
  geom_vline(xintercept = mean(ems$response_time),
             color = "#D85A30",
             linetype = "dashed",
             linewidth = 0.9) +
  geom_vline(xintercept = median(ems$response_time),
             color = "#1D9E75",
             linetype = "dashed",
             linewidth = 0.9) +
  annotate("text",
           x     = mean(ems$response_time) + 1.5,
           y     = 26,
           label = paste0("Mean = ",
                          round(mean(ems$response_time), 1)),
           color = "#D85A30", size = 3.5) +
  annotate("text",
           x     = median(ems$response_time) - 1.5,
           y     = 23,
           label = paste0("Median = ",
                          round(median(ems$response_time), 1)),
           color = "#1D9E75", size = 3.5) +
  labs(
    title   = "Figure 1. Distribution of EMS On-Scene Response Time",
    x       = "Response Time (minutes)",
    y       = "Frequency (n)",
    caption = "Note. Orange dashed line = mean; green dashed line = median. N = 200."
  ) +
  theme_minimal(base_size = 12) +
  theme(
    plot.title   = element_text(face = "bold", size = 12),
    plot.caption = element_text(hjust = 0, size = 9,
                                color = "gray40")
  )
```

### Figure 2. Response Time by EMT Level

```{r}
#| fig-width: 7
#| fig-height: 4.5

ggplot(ems, aes(x = level, y = response_time,
                fill = level)) +
  geom_boxplot(alpha = 0.6,
               outlier.shape = 21,
               outlier.fill  = "white",
               outlier.size  = 2) +
  geom_jitter(width = 0.15, alpha = 0.25, size = 1.2) +
  scale_fill_manual(
    values = c("EMT-1" = "#F5C4B3",
               "EMT-2" = "#9FE1CB",
               "EMT-P" = "#85B7EB")
  ) +
  labs(
    title   = "Figure 2. Response Time Distribution by EMT Level",
    x       = "EMT Certification Level",
    y       = "Response Time (minutes)",
    caption = "Note. Center line = median; box = IQR; points = individual observations. N = 200."
  ) +
  theme_minimal(base_size = 12) +
  theme(
    plot.title      = element_text(face = "bold", size = 12),
    legend.position = "none",
    plot.caption    = element_text(hjust = 0, size = 9,
                                   color = "gray40")
  )
```

### Figure 3. EMT Level Distribution

```{r}
#| fig-width: 6
#| fig-height: 4

ems |>
  dplyr::count(level) |>
  dplyr::mutate(pct = round(n / sum(n) * 100, 1)) |>
  ggplot(aes(x = level, y = pct, fill = level)) +
  geom_col(alpha = 0.8, width = 0.6) +
  geom_text(
    aes(label = paste0(pct, "%\n(n=", n, ")")),
    vjust = -0.4, size = 3.5
  ) +
  scale_fill_manual(
    values = c("EMT-1" = "#F5C4B3",
               "EMT-2" = "#9FE1CB",
               "EMT-P" = "#85B7EB")
  ) +
  scale_y_continuous(limits = c(0, 65)) +
  labs(
    title   = "Figure 3. Distribution of EMS Personnel by EMT Level",
    x       = "EMT Certification Level",
    y       = "Percentage (%)",
    caption = "Note. N = 200."
  ) +
  theme_minimal(base_size = 12) +
  theme(
    plot.title      = element_text(face = "bold", size = 12),
    legend.position = "none",
    plot.caption    = element_text(hjust = 0, size = 9,
                                   color = "gray40")
  )
```

### Table 4. Epidemiological Measures — Occupational Injury

```{r}
n_injury   <- sum(ems$occ_injury == 1, na.rm = TRUE)
n_total    <- nrow(ems)
prevalence <- n_injury / n_total

ems_rr <- ems |>
  dplyr::mutate(
    night = ifelse(shift == "夜班為主",
                   "Night shift", "Non-night shift")
  )

tbl_2x2    <- table(ems_rr$night, ems_rr$occ_injury)
rate_night <- tbl_2x2["Night shift",     "1"] /
              rowSums(tbl_2x2)["Night shift"]
rate_non   <- tbl_2x2["Non-night shift", "1"] /
              rowSums(tbl_2x2)["Non-night shift"]
RR         <- rate_night / rate_non

data.frame(
  Measure = c("Occupational injury prevalence",
              "Night-shift injury rate",
              "Non-night-shift injury rate",
              "Relative Risk (RR)"),
  Value   = c(
    paste0(round(prevalence * 100, 1),
           "% (", n_injury, "/", n_total, ")"),
    paste0(round(rate_night * 100, 1), "%"),
    paste0(round(rate_non   * 100, 1), "%"),
    as.character(round(RR, 2))
  ),
  Interpretation = c(
    "Proportion with occupational injury in the past year",
    "Among night-shift workers",
    "Among day or rotating shift workers",
    "Risk multiplier: night shift vs non-night shift"
  )
) |>
  gt() |>
  tab_header(
    title = "Table 4. Epidemiological Measures — Occupational Injury"
  ) |>
  tab_source_note(
    source_note = "Note. RR = Relative Risk. RR > 1 indicates elevated risk. N = 200."
  ) |>
  cols_align(align = "left", columns = everything()) |>
  tab_style(
    style     = cell_text(weight = "bold"),
    locations = cells_column_labels()
  ) |>
  tab_options(
    table.border.top.style            = "hidden",
    heading.border.bottom.style       = "solid",
    column_labels.border.top.style    = "solid",
    column_labels.border.bottom.style = "solid",
    table.border.bottom.style         = "solid",
    table.font.size                   = px(13),
    heading.title.font.size           = px(14),
    heading.align                     = "left"
  )
```

---

## III. Reporting Results (APA Format)

**Continuous variable (normal distribution):**

> The mean age of EMS personnel was 34.2 years (*SD* = 6.8). Mean on-scene response time was 9.3 minutes (*SD* = 2.5).

**Continuous variable (skewed):**

> The median monthly call volume was 48 calls (*IQR*: 41–55), with a slight positive skew.

**Categorical variable:**

> EMT-1 accounted for 49.0% of participants (*n* = 98), followed by EMT-2 (31.0%, *n* = 62) and EMT-P (20.0%, *n* = 40).

**Epidemiological measure:**

> The one-year prevalence of occupational injury was 22.0% (44/200). Night-shift workers had 1.8 times the occupational injury rate of non-night-shift workers (*RR* = 1.80), suggesting that night-shift work may increase injury risk.

---

## Weekly Assignment

**Q1** Calculate the mean, median, SD, and IQR for `stress_vas`. Create an APA-format `gt` table and plot a histogram. Based on the distribution shape, justify whether mean or median is more appropriate.

**Q2** Use `group_by()` + `summarise()` to compute descriptive statistics for `response_time` by district. Present results in an APA-format `gt` table with a boxplot.

**Q3** Tabulate frequency and percentage for each EMT level. Create an APA-format `gt` table and a labeled bar chart.

**Q4** Calculate the prevalence of `occ_injury` and compare the injury rate between "Rotating" and "Day" shift workers. Compute RR and report in APA format.

**Q5 Reflection:** Compare the mean and median of `mbi_ee`. Are they substantially different? What does this tell you about the distribution of this variable?