---
title: "Topic 5：Correlation Analysis"
subtitle: "Pearson Correlation, Spearman Correlation, Correlation Matrix, APA Reporting"
execute:
  warning: false
  message: false
  echo: true
---

## Learning Objectives

1. Understand the meaning and limitations of correlation coefficients
2. Select between Pearson and Spearman correlation appropriately
3. Test the statistical significance of a correlation coefficient
4. Produce an APA-format correlation matrix table

---

## Statistical Method Map (This Topic)

| X (Predictor) | Y (Outcome) | Method | Assumptions |
|:-------------|:-----------|:-------|:-----------|
| Continuous (normal) | Continuous (normal) | Pearson correlation | Bivariate normality, linear relationship |
| Continuous (skewed) | Continuous (skewed) | Spearman correlation | No distributional assumption, monotonic relationship |
| Ordinal | Ordinal | Spearman correlation | No distributional assumption |

::: {.callout-note}
Correlation is **symmetric** — there is no directional distinction between X and Y.
r(stress, burnout) = r(burnout, stress).
This differs from regression (Topic 9), which has an explicit prediction direction.
:::

---

## I. Statistical Theory

### 1.1 What Does a Correlation Coefficient Measure?

The correlation coefficient r quantifies the **direction and strength of the linear relationship** between two continuous variables:

$$r \in [-1, +1]$$

- r = +1: perfect positive correlation
- r = 0: no linear relationship (does not mean no relationship)
- r = -1: perfect negative correlation

::: {.callout-important}
**r = 0 does not mean the variables are unrelated.**
It only means there is no *linear* relationship. Two variables can have a perfect curvilinear relationship and still yield r = 0. This is why a scatterplot must always be examined before computing r.
:::

### 1.2 Pearson Correlation Coefficient

$$r = \frac{\sum_{i=1}^{n}(x_i - \bar{x})(y_i - \bar{y})}
           {\sqrt{\sum(x_i-\bar{x})^2 \cdot \sum(y_i-\bar{y})^2}}$$

The numerator is the covariance — how much X and Y deviate from their respective means together. The denominator standardizes by the product of their standard deviations.

**Intuition:** When X is above its mean, is Y also above its mean? If yes → positive correlation; if opposite → negative correlation.

**Assumptions:**

1. Both variables are continuous
2. Data approximate a bivariate normal distribution
3. The relationship is linear (confirm with scatterplot)
4. No severe outliers

### 1.3 Spearman Correlation Coefficient

$$r_s = 1 - \frac{6\sum d_i^2}{n(n^2-1)}$$

where $d_i$ is the difference in ranks for each pair.

**Approach:** Replace raw values with ranks, then compute Pearson correlation on the ranks.

**Advantages:**

- No normality assumption required
- Robust to outliers
- Appropriate for ordinal data (Likert scales)
- Detects any monotonic relationship (not limited to linear)

**Selection guide:**

```
Two continuous variables
    ↓
Shapiro-Wilk normality test
    ├── Both p > .05 (normal) → Pearson
    ├── Either p < .05 (non-normal) → Spearman
    └── Severe outliers present → Spearman
```

### 1.4 Effect Size: Interpreting the Magnitude of r

| |r| range | Strength | Cohen (1988) |
|:---------|:--------|:---------|
| 0.10–0.29 | Small effect | Exists but weak |
| 0.30–0.49 | Medium effect | Noticeable but moderate |
| ≥ 0.50 | Large effect | Strong association |

::: {.callout-tip}
In EMS and health science research, r = .30 is already practically meaningful. Do not dismiss a "small" correlation without considering the research context, sample size, and study purpose.
:::

### 1.5 Significance Testing for Correlation

$$t = \frac{r\sqrt{n-2}}{\sqrt{1-r^2}}, \quad df = n-2$$

H₀: ρ = 0 (the population correlation coefficient is zero)

The p-value answers: "If there truly is no linear relationship in the population, how probable is it to observe a correlation this large by chance?"

::: {.callout-important}
**The large-sample trap:**
With n = 1,000, even r = .06 will yield p < .05. Statistical significance does not equal practical importance. Always report both r (effect size) and p (significance).
:::

### 1.6 Correlation ≠ Causation

Observing a positive correlation between stress_vas and mbi_ee (r = .58, p < .001) allows us to say "workers with higher stress tend to have higher emotional exhaustion" — but **not** that "stress causes emotional exhaustion."

Reasons:

1. **Direction is ambiguous:** Does stress cause burnout, or does burnout make people feel more stressed?
2. **Third variables:** Night shifts may independently influence both (confounding)
3. **Chance:** Especially in the context of multiple comparisons

Causal inference requires random assignment (RCT) or appropriate causal study designs.

---

## II. Analysis Workflow

```
Step 1: Plot scatterplot (check linearity and outliers)
    ↓
Step 2: Shapiro-Wilk normality test
    ↓
Step 3: Choose Pearson or Spearman
    ↓
Step 4: Compute r and p-value
    ↓
Step 5: Assess effect size magnitude
    ↓
Step 6: Report in APA format
```

---

## III. R Implementation

### Load packages and data

```{r}
library(tidyverse)
library(gt)
library(corrplot)
# install.packages("corrplot")

ems <- read.csv("data/ems_main.csv",
                stringsAsFactors = FALSE)

ems$gender   <- factor(ems$gender,
                        levels = c("男", "女"))
ems$level    <- factor(ems$level,
                        levels = c("EMT-1", "EMT-2", "EMT-P"))
ems$district <- factor(ems$district,
                        levels = c("北部", "中部", "南部", "東部"))
ems$shift    <- factor(ems$shift,
                        levels = c("日班為主", "夜班為主", "輪班"))
```

### Figure 1. Scatterplot (Always First)

```{r}
#| fig-width: 7
#| fig-height: 5

ggplot(ems, aes(x = stress_vas, y = mbi_ee)) +
  geom_point(alpha = 0.5, color = "#85B7EB", size = 2) +
  geom_smooth(method = "lm", se = TRUE,
              color = "#D85A30", linewidth = 0.9) +
  labs(
    title   = "Figure 1. Scatterplot: Perceived Stress vs Emotional Exhaustion",
    x       = "Perceived Stress (VAS 0–10)",
    y       = "MBI Emotional Exhaustion (0–6)",
    caption = "Note. Red line = linear trend (with 95% CI). N = 200."
  ) +
  theme_minimal(base_size = 12) +
  theme(
    plot.title   = element_text(face = "bold", size = 12),
    plot.caption = element_text(hjust = 0, size = 9,
                                color = "gray40")
  )
```

### Table 1. Shapiro-Wilk Normality Test

```{r}
sw_stress <- shapiro.test(ems$stress_vas)
sw_mbi    <- shapiro.test(ems$mbi_ee)

data.frame(
  Variable = c("Perceived Stress (stress_vas)",
               "Emotional Exhaustion (mbi_ee)"),
  W        = c(round(sw_stress$statistic, 3),
               round(sw_mbi$statistic, 3)),
  p        = c(round(sw_stress$p.value, 3),
               round(sw_mbi$p.value, 3)),
  Decision = c(ifelse(sw_stress$p.value > .05,
                      "Fail to reject normality",
                      "Reject normality"),
               ifelse(sw_mbi$p.value > .05,
                      "Fail to reject normality",
                      "Reject normality"))
) |>
  gt() |>
  tab_header(
    title = "Table 1. Shapiro-Wilk Normality Test"
  ) |>
  tab_source_note(
    source_note = "Note. p > .05 indicates failure to reject the normality assumption. N = 200."
  ) |>
  cols_align(align = "center", columns = c(W, p)) |>
  cols_align(align = "left",
             columns = c(Variable, Decision)) |>
  tab_style(
    style     = cell_text(weight = "bold"),
    locations = cells_column_labels()
  ) |>
  tab_options(
    table.border.top.style            = "hidden",
    heading.border.bottom.style       = "solid",
    column_labels.border.top.style    = "solid",
    column_labels.border.bottom.style = "solid",
    table.border.bottom.style         = "solid",
    table.font.size                   = px(13),
    heading.title.font.size           = px(14),
    heading.align                     = "left"
  )
```

### Table 2. Pearson Correlation

```{r}
pr <- cor.test(ems$stress_vas, ems$mbi_ee,
               method = "pearson")

data.frame(
  Variables  = "Perceived Stress × Emotional Exhaustion",
  Method     = "Pearson",
  r          = round(pr$estimate, 3),
  t          = round(pr$statistic, 3),
  df         = pr$parameter,
  p          = ifelse(pr$p.value < .001,
                      "< .001",
                      round(pr$p.value, 3)),
  CI         = paste0("[",
                 round(pr$conf.int[1], 3), ", ",
                 round(pr$conf.int[2], 3), "]"),
  Effect     = dplyr::case_when(
    abs(pr$estimate) >= .50 ~ "Large",
    abs(pr$estimate) >= .30 ~ "Medium",
    TRUE ~ "Small"
  )
) |>
  gt() |>
  tab_header(
    title = "Table 2. Pearson Correlation: Perceived Stress and Emotional Exhaustion"
  ) |>
  tab_source_note(
    source_note = "Note. N = 200. CI = 95% confidence interval. Effect size criteria: Cohen (1988)."
  ) |>
  cols_label(
    Variables = "Variables", Method = "Method",
    r = "r", t = "t", df = "df",
    p = "p", CI = "95% CI", Effect = "Effect Size"
  ) |>
  cols_align(align = "center",
             columns = c(Method, r, t, df, p, CI, Effect)) |>
  cols_align(align = "left", columns = Variables) |>
  tab_style(
    style     = cell_text(weight = "bold"),
    locations = cells_column_labels()
  ) |>
  tab_options(
    table.border.top.style            = "hidden",
    heading.border.bottom.style       = "solid",
    column_labels.border.top.style    = "solid",
    column_labels.border.bottom.style = "solid",
    table.border.bottom.style         = "solid",
    table.font.size                   = px(13),
    heading.title.font.size           = px(14),
    heading.align                     = "left"
  )
```

### Table 3. Spearman Correlation

```{r}
sr <- cor.test(ems$stress_vas, ems$mbi_ee,
               method = "spearman",
               exact  = FALSE)

data.frame(
  Variables = "Perceived Stress × Emotional Exhaustion",
  Method    = "Spearman",
  rho       = round(sr$estimate, 3),
  S         = round(sr$statistic, 1),
  p         = ifelse(sr$p.value < .001,
                     "< .001",
                     round(sr$p.value, 3)),
  Effect    = dplyr::case_when(
    abs(sr$estimate) >= .50 ~ "Large",
    abs(sr$estimate) >= .30 ~ "Medium",
    TRUE ~ "Small"
  )
) |>
  gt() |>
  tab_header(
    title = "Table 3. Spearman Correlation: Perceived Stress and Emotional Exhaustion"
  ) |>
  tab_source_note(
    source_note = "Note. N = 200. Spearman ρ is appropriate for non-normal or ordinal data."
  ) |>
  cols_label(
    Variables = "Variables", Method = "Method",
    rho = "ρ", S = "S", p = "p", Effect = "Effect Size"
  ) |>
  cols_align(align = "center",
             columns = c(Method, rho, S, p, Effect)) |>
  cols_align(align = "left", columns = Variables) |>
  tab_style(
    style     = cell_text(weight = "bold"),
    locations = cells_column_labels()
  ) |>
  tab_options(
    table.border.top.style            = "hidden",
    heading.border.bottom.style       = "solid",
    column_labels.border.top.style    = "solid",
    column_labels.border.bottom.style = "solid",
    table.border.bottom.style         = "solid",
    table.font.size                   = px(13),
    heading.title.font.size           = px(14),
    heading.align                     = "left"
  )
```

### Table 4. Correlation Matrix (APA Format)

```{r}
cor_vars <- ems |>
  dplyr::select(stress_vas, mbi_ee, mbi_dp, mbi_pa,
                sleep_hrs, monthly_cases, response_time)

cor_mat <- cor(cor_vars,
               method = "pearson",
               use    = "complete.obs")

cor_df <- as.data.frame(round(cor_mat, 2))
cor_df <- tibble::rownames_to_column(cor_df, "Variable")
cor_df$Variable <- c("1. Perceived Stress",
                     "2. Emotional Exhaustion",
                     "3. Depersonalization",
                     "4. Personal Accomplishment",
                     "5. Sleep Hours",
                     "6. Monthly Calls",
                     "7. Response Time")
names(cor_df) <- c("Variable","1","2","3","4","5","6","7")

cor_df |>
  gt() |>
  tab_header(
    title = "Table 4. Correlation Matrix for Key Study Variables"
  ) |>
  tab_source_note(
    source_note = paste0(
      "Note. N = 200. Values are Pearson correlation coefficients r. ",
      "|r| ≥ .50 = large effect; |r| ≥ .30 = medium effect; ",
      "|r| ≥ .10 = small effect (Cohen, 1988)."
    )
  ) |>
  cols_align(align = "center", columns = -Variable) |>
  cols_align(align = "left",   columns = Variable) |>
  tab_style(
    style     = cell_text(weight = "bold"),
    locations = cells_column_labels()
  ) |>
  tab_options(
    table.border.top.style            = "hidden",
    heading.border.bottom.style       = "solid",
    column_labels.border.top.style    = "solid",
    column_labels.border.bottom.style = "solid",
    table.border.bottom.style         = "solid",
    table.font.size                   = px(13),
    heading.title.font.size           = px(14),
    heading.align                     = "left"
  )
```

### Figure 2. Correlation Matrix Heatmap

```{r}
#| fig-width: 7
#| fig-height: 6

rownames(cor_mat) <- c("Perceived Stress",
                        "Emotional Exhaustion",
                        "Depersonalization",
                        "Personal Accomplishment",
                        "Sleep Hours",
                        "Monthly Calls",
                        "Response Time")
colnames(cor_mat) <- rownames(cor_mat)

corrplot(cor_mat,
         method      = "color",
         type        = "upper",
         order       = "hclust",
         addCoef.col = "black",
         tl.col      = "black",
         tl.cex      = 0.80,
         number.cex  = 0.72,
         col         = colorRampPalette(
           c("#D85A30", "white", "#185FA5"))(200),
         mar         = c(0, 0, 2, 0))
title("Figure 2. Correlation Matrix Heatmap",
      cex.main = 1.0, font.main = 2)
```

---

## IV. Reporting Results (APA Format)

**Single correlation:**

> Perceived stress (stress_vas) and emotional exhaustion (mbi_ee) were significantly positively correlated, *r*(198) = .58, *p* < .001, 95% CI [.48, .67], representing a large effect size.

**Correlation matrix (text summarizes key findings; table presents all values):**

> Table 4 presents the correlation matrix for key study variables. Perceived stress was significantly positively correlated with emotional exhaustion (*r* = .58) and depersonalization (*r* = .47), both *p*s < .001. Sleep hours were significantly negatively correlated with emotional exhaustion (*r* = −.34, *p* < .001), suggesting that insufficient sleep is associated with greater occupational burnout.

---

## Weekly Assignment

**Q1** Plot a scatterplot of `sleep_hrs` vs `stress_vas`, run Shapiro-Wilk for both variables, select Pearson or Spearman based on the results (explain your reasoning), and present the result in an APA-format `gt` table.

**Q2** Compute a correlation matrix for the following five variables: `stress_vas`, `mbi_ee`, `sleep_hrs`, `overtime_hours`, `job_sat`. Present in an APA-format `gt` table.

**Q3** Plot a `corrplot` heatmap for the matrix in Q2. Identify the variable pair with the largest absolute correlation and explain its practical meaning in the EMS context.

**Q4 Reflection:** You find that `monthly_cases` and `mbi_ee` are positively correlated (*r* = .28, *p* < .001). A colleague concludes: "EMS workers with more calls experience more burnout — we should reduce their call volume to improve mental health." What is the problem with this reasoning? How would you respond?