---
title: "Topic 10：Binary Logistic Regression"
subtitle: "Logit Transformation, Odds Ratios, ROC Curve, APA Reporting"
execute:
  warning: false
  message: false
  echo: true
---

## Learning Objectives

1. Understand the statistical logic of logistic regression and the logit transformation
2. Correctly interpret odds ratios (OR)
3. Assess model fit and construct ROC curves
4. Produce APA-format result tables and reporting sentences

---

## Statistical Method Map (This Topic)

| X (Predictor) | Y (Outcome) | Method |
|:-------------|:-----------|:-------|
| Continuous / Categorical | **Binary (0/1)** | Binary logistic regression |
| Continuous / Categorical | Multinomial (≥ 3 categories) | Multinomial logistic regression |
| Continuous / Categorical | Ordinal | Ordinal logistic regression |

::: {.callout-note}
**Why not use linear regression when Y is binary?**

Linear regression predictions can fall outside [0, 1], but probabilities must be constrained to this range. Logistic regression solves this problem through the logit transformation.
:::

---

## I. Statistical Theory

### 1.1 From Linear to Logistic Regression

When Y is binary (0/1), the true target is the probability that Y = 1: P(Y=1|X).

Using linear regression directly: P(Y=1) = β₀ + β₁X → predictions may be < 0 or > 1, which is nonsensical.

**Solution: Logit Transformation**

Convert probability to odds, then take the natural log:

$$\text{logit}(p) = \ln\left(\frac{p}{1-p}\right) = \beta_0 + \beta_1 X_1 + \cdots + \beta_k X_k$$

**Inverse transformation (Sigmoid function):**

$$P(Y=1) = \frac{e^{\beta_0 + \beta_1 X}}{1 + e^{\beta_0 + \beta_1 X}} = \frac{1}{1+e^{-(\beta_0 + \beta_1 X)}}$$

This S-shaped curve maps any real number to the interval (0, 1).

### 1.2 Odds Ratio (OR)

**Odds:** The ratio of the probability that an event occurs to the probability that it does not:

$$\text{Odds} = \frac{P(Y=1)}{P(Y=0)} = \frac{p}{1-p}$$

**Odds Ratio (OR):** The multiplicative change in odds for a 1-unit increase in X:

$$OR = e^{\beta_1}$$

**Interpretation:**

| OR Value | Meaning |
|:---------|:--------|
| OR = 1 | X has no effect on Y |
| OR > 1 | Increasing X increases the odds of Y = 1 |
| OR < 1 | Increasing X decreases the odds of Y = 1 |
| OR = 2 | Each 1-unit increase in X doubles the odds |
| OR = 0.5 | Each 1-unit increase in X halves the odds |

::: {.callout-important}
**OR ≠ RR (Relative Risk)!**

OR = 2 does not mean "twice the probability" — it means "twice the odds." When outcome prevalence is low (< 10%), OR ≈ RR. At higher prevalence, OR overestimates RR.
:::

### 1.3 Confidence Intervals and Significance

$$95\% \text{ CI for OR} = e^{\hat{\beta} \pm 1.96 \times SE(\hat{\beta})}$$

**Assessing significance:**
- CI excludes 1 → statistically significant (equivalent to p < .05)
- CI includes 1 → not significant

### 1.4 Model Fit

**Hosmer-Lemeshow Test:**
H₀: Model fits adequately. p > .05 indicates acceptable fit.

**Nagelkerke R²:**
Pseudo-R² analogous to R² in linear regression; reflects the proportion of variance explained (0–1).

**AIC (Akaike Information Criterion):**
Smaller is better; used for model comparison.

### 1.5 ROC Curve and AUC

The ROC (Receiver Operating Characteristic) curve plots Sensitivity (True Positive Rate) against 1-Specificity (False Positive Rate) across all possible classification thresholds.

**AUC (Area Under the Curve):**

| AUC Range | Discriminative Ability |
|:----------|:----------------------|
| 0.50 | No discrimination (random guessing) |
| 0.70–0.79 | Acceptable |
| 0.80–0.89 | Good |
| ≥ 0.90 | Excellent |

---

## II. Analysis Workflow

```
Step 1: Confirm Y is binary (0/1)
    ↓
Step 2: Descriptive statistics (proportions by group)
    ↓
Step 3: Build logistic regression model
    ↓
Step 4: Compute OR and 95% CI
    ↓
Step 5: Model fit (Hosmer-Lemeshow, Nagelkerke R²)
    ↓
Step 6: ROC curve and AUC
    ↓
Step 7: APA-format report
```

---

## III. R Implementation

### Load packages and data

```{r}
library(tidyverse)
library(gt)
library(pROC)
library(ResourceSelection)

ems <- read.csv("data/ems_main.csv",
                stringsAsFactors = FALSE)

ems$gender   <- factor(ems$gender,
                        levels = c("男", "女"))
ems$level    <- factor(ems$level,
                        levels = c("EMT-1", "EMT-2", "EMT-P"))
ems$district <- factor(ems$district,
                        levels = c("北部", "中部", "南部", "東部"))
ems$shift    <- factor(ems$shift,
                        levels = c("日班為主", "夜班為主", "輪班"))
```

**Research question:** Which factors predict whether EMS personnel experience occupational injury (`occ_injury`)?

### Figure 1. Occupational Injury Rate by EMT Level

```{r}
#| fig-width: 7
#| fig-height: 4.5

ems |>
  dplyr::group_by(level) |>
  dplyr::summarise(
    injury_rate = mean(occ_injury) * 100,
    n = n(),
    .groups = "drop"
  ) |>
  ggplot(aes(x = level, y = injury_rate, fill = level)) +
  geom_col(alpha = 0.8, width = 0.6) +
  geom_text(aes(label = paste0(round(injury_rate, 1), "%")),
            vjust = -0.5, size = 3.5) +
  scale_fill_manual(
    values = c("EMT-1" = "#F5C4B3",
               "EMT-2" = "#9FE1CB",
               "EMT-P" = "#85B7EB")
  ) +
  scale_y_continuous(limits = c(0, 40)) +
  labs(
    title   = "Figure 1. Occupational Injury Rate by EMT Certification Level",
    x       = "EMT Certification Level",
    y       = "Occupational Injury Rate (%)",
    caption = "Note. Occupational injury defined as any work-related injury in the past year. N = 200."
  ) +
  theme_minimal(base_size = 12) +
  theme(
    plot.title      = element_text(face = "bold", size = 12),
    legend.position = "none",
    plot.caption    = element_text(hjust = 0, size = 9,
                                   color = "gray40")
  )
```

### Table 1. Outcome Variable Distribution

```{r}
ems |>
  dplyr::count(occ_injury) |>
  dplyr::mutate(
    Status     = c("No occupational injury",
                   "Occupational injury"),
    Percentage = round(n / sum(n) * 100, 1)
  ) |>
  dplyr::select(Status, n, Percentage) |>
  gt() |>
  tab_header(
    title = "Table 1. Distribution of Occupational Injury"
  ) |>
  tab_source_note(
    source_note = "Note. N = 200. Occupational injury = work-related injury in the past year (occ_injury = 1)."
  ) |>
  cols_align(align = "center", columns = c(n, Percentage)) |>
  cols_align(align = "left",   columns = Status) |>
  tab_style(
    style     = cell_text(weight = "bold"),
    locations = cells_column_labels()
  ) |>
  tab_options(
    table.border.top.style            = "hidden",
    heading.border.bottom.style       = "solid",
    column_labels.border.top.style    = "solid",
    column_labels.border.bottom.style = "solid",
    table.border.bottom.style         = "solid",
    table.font.size                   = px(13),
    heading.title.font.size           = px(14),
    heading.align                     = "left"
  )
```

### Run Logistic Regression

```{r}
logit_model <- glm(occ_injury ~ stress_vas +
                     night_shifts + years + sleep_hrs,
                   data   = ems,
                   family = binomial(link = "logit"))

logit_sum <- summary(logit_model)
```

### Table 2. Logistic Regression Results (with OR and 95% CI)

```{r}
coef_df <- as.data.frame(logit_sum$coefficients)
coef_df <- tibble::rownames_to_column(coef_df, "Variable")
ci_vals <- confint(logit_model)

data.frame(
  Variable = c("Intercept", "Perceived Stress",
               "Night Shifts", "Years of Service",
               "Sleep Hours"),
  B        = round(coef_df$Estimate, 3),
  SE       = round(coef_df$`Std. Error`, 3),
  OR       = round(exp(coef_df$Estimate), 3),
  CI_L     = round(exp(ci_vals[, 1]), 3),
  CI_U     = round(exp(ci_vals[, 2]), 3),
  p        = ifelse(coef_df$`Pr(>|z|)` < .001,
                    "< .001",
                    round(coef_df$`Pr(>|z|)`, 3))
) |>
  dplyr::mutate(
    `95% CI` = paste0("[", CI_L, ", ", CI_U, "]")
  ) |>
  dplyr::select(Variable, B, SE, OR, `95% CI`, p) |>
  gt() |>
  tab_header(
    title = "Table 2. Logistic Regression: Predictors of Occupational Injury"
  ) |>
  tab_source_note(
    source_note = paste0(
      "Note. B = logit coefficient; SE = standard error; ",
      "OR = odds ratio (= e^B); CI = 95% confidence interval of OR. ",
      "N = 200. CI excluding 1 indicates statistical significance (p < .05)."
    )
  ) |>
  cols_align(align = "center",
             columns = c(B, SE, OR, `95% CI`, p)) |>
  cols_align(align = "left", columns = Variable) |>
  tab_style(
    style     = cell_text(weight = "bold"),
    locations = cells_column_labels()
  ) |>
  tab_options(
    table.border.top.style            = "hidden",
    heading.border.bottom.style       = "solid",
    column_labels.border.top.style    = "solid",
    column_labels.border.bottom.style = "solid",
    table.border.bottom.style         = "solid",
    table.font.size                   = px(13),
    heading.title.font.size           = px(14),
    heading.align                     = "left"
  )
```

### Table 3. Model Fit Indices

```{r}
null_model  <- glm(occ_injury ~ 1,
                   data   = ems,
                   family = binomial)
nagel_r2    <- 1 - exp((logit_model$deviance -
                         null_model$deviance) / nrow(ems))
nagel_r2_max <- 1 - exp(-null_model$deviance / nrow(ems))
nagel_r2_n  <- nagel_r2 / nagel_r2_max

hl_test <- hoslem.test(ems$occ_injury,
                        fitted(logit_model), g = 10)

data.frame(
  Index          = c("AIC",
                     "Nagelkerke R²",
                     "Hosmer-Lemeshow χ²",
                     "Hosmer-Lemeshow p"),
  Value          = c(round(AIC(logit_model), 2),
                     round(nagel_r2_n, 3),
                     round(hl_test$statistic, 3),
                     round(hl_test$p.value, 3)),
  Interpretation = c("Smaller is better; use for model comparison",
                     "Pseudo-R²; proportion of variance explained (0–1)",
                     "Goodness-of-fit statistic",
                     "p > .05 indicates adequate model fit")
) |>
  gt() |>
  tab_header(
    title = "Table 3. Logistic Regression Model Fit Indices"
  ) |>
  tab_source_note(
    source_note = "Note. N = 200."
  ) |>
  cols_align(align = "center", columns = Value) |>
  cols_align(align = "left",
             columns = c(Index, Interpretation)) |>
  tab_style(
    style     = cell_text(weight = "bold"),
    locations = cells_column_labels()
  ) |>
  tab_options(
    table.border.top.style            = "hidden",
    heading.border.bottom.style       = "solid",
    column_labels.border.top.style    = "solid",
    column_labels.border.bottom.style = "solid",
    table.border.bottom.style         = "solid",
    table.font.size                   = px(13),
    heading.title.font.size           = px(14),
    heading.align                     = "left"
  )
```

### Figure 2. ROC Curve

```{r}
#| fig-width: 6
#| fig-height: 6

roc_obj <- roc(ems$occ_injury,
               fitted(logit_model),
               quiet = TRUE)

auc_val <- auc(roc_obj)

ggroc(roc_obj, color = "#185FA5", linewidth = 1) +
  geom_abline(slope     = 1,
              intercept = 1,
              linetype  = "dashed",
              color     = "gray60") +
  annotate("text", x = 0.35, y = 0.15,
           label    = paste0("AUC = ",
                             round(auc_val, 3)),
           size     = 4.5,
           color    = "#185FA5",
           fontface = "bold") +
  labs(
    title   = "Figure 2. ROC Curve: Occupational Injury Prediction Model",
    x       = "1 − Specificity (False Positive Rate)",
    y       = "Sensitivity (True Positive Rate)",
    caption = "Note. Dashed line = chance performance (AUC = 0.50). N = 200."
  ) +
  theme_minimal(base_size = 12) +
  theme(
    plot.title   = element_text(face = "bold", size = 12),
    plot.caption = element_text(hjust = 0, size = 9,
                                color = "gray40")
  )
```

### Table 4. ROC Curve Summary

```{r}
ci_auc <- ci.auc(roc_obj)

data.frame(
  Index              = c("AUC",
                         "95% CI Lower Bound",
                         "95% CI Upper Bound",
                         "Discriminative Ability"),
  Value              = c(round(auc_val, 3),
                         round(ci_auc[1], 3),
                         round(ci_auc[3], 3),
                         dplyr::case_when(
                           auc_val >= .90 ~ "Excellent",
                           auc_val >= .80 ~ "Good",
                           auc_val >= .70 ~ "Acceptable",
                           TRUE           ~ "Poor"
                         ))
) |>
  gt() |>
  tab_header(
    title = "Table 4. ROC Curve Analysis Summary"
  ) |>
  tab_source_note(
    source_note = paste0(
      "Note. AUC = Area Under the Curve; ",
      "CI = 95% confidence interval (DeLong method). ",
      "AUC ≥ .80 indicates good discriminative ability. N = 200."
    )
  ) |>
  cols_align(align = "center", columns = Value) |>
  cols_align(align = "left",   columns = Index) |>
  tab_style(
    style     = cell_text(weight = "bold"),
    locations = cells_column_labels()
  ) |>
  tab_options(
    table.border.top.style            = "hidden",
    heading.border.bottom.style       = "solid",
    column_labels.border.top.style    = "solid",
    column_labels.border.bottom.style = "solid",
    table.border.bottom.style         = "solid",
    table.font.size                   = px(13),
    heading.title.font.size           = px(14),
    heading.align                     = "left"
  )
```

---

## IV. Reporting Results (APA Format)

**Logistic regression results:**

> Binary logistic regression was used to examine predictors of occupational injury. The overall model was significant (χ²(4) = 18.42, *p* = .001), with Nagelkerke *R*² = .13 and adequate model fit (Hosmer-Lemeshow *p* = .62).

**OR interpretation:**

> Night shift frequency significantly predicted occupational injury (*OR* = 1.08, 95% CI [1.02, 1.14], *p* = .008), indicating that each additional night shift was associated with an 8% increase in the odds of occupational injury. Sleep hours were a protective factor (*OR* = 0.71, 95% CI [0.54, 0.93], *p* = .013), with each additional hour of sleep associated with a 29% reduction in injury odds.

**ROC curve:**

> ROC curve analysis indicated AUC = 0.73 (95% CI [0.64, 0.82]), representing acceptable discriminative ability.

---

## Weekly Assignment

**Q1** Use `stress_vas`, `sleep_hrs`, and `overtime_hours` to predict `occ_injury`. Run binary logistic regression and present an APA-format `gt` coefficient table (including OR and 95% CI).

**Q2** Compute model fit indices (AIC, Nagelkerke R², Hosmer-Lemeshow) and present in an APA-format `gt` table. Assess whether the model fits adequately.

**Q3** Plot the ROC curve and compute AUC. Based on the AUC value, assess the model's discriminative ability.

**Q4 Reflection:** In your model, `stress_vas` has OR = 1.15 (95% CI [0.98, 1.35]), p = .089 — the CI includes 1. A colleague concludes "this means stress is unrelated to occupational injury." Do you agree? How would you interpret this result from both statistical and practical perspectives?