---
title: "Topic 12：Repeated Measures ANOVA and GEE"
subtitle: "Longitudinal Data Analysis, Sphericity, GEE Correlation Structures, APA Reporting"
execute:
  warning: false
  message: false
  echo: true
---

## Learning Objectives

1. Understand the statistical logic of longitudinal data and repeated measures designs
2. Correctly execute repeated measures ANOVA
3. Understand the advantages and appropriate use of GEE
4. Produce APA-format result tables and reporting sentences

---

## Statistical Method Map (This Topic)

| X (Predictor) | Y (Outcome) | Data Structure | Method |
|:-------------|:-----------|:--------------|:-------|
| Repeated time points | Continuous | Wide format | Repeated Measures ANOVA |
| Repeated time points | Continuous | Long format | GEE |
| Repeated time points + Group | Continuous | Long format | Mixed ANOVA / GEE |

::: {.callout-note}
The core challenge of repeated measures data is **non-independence** — observations from the same person at different time points are correlated. Standard ANOVA assumes independence; applying it naively to repeated measures data underestimates standard errors and inflates Type I error.

Both repeated measures ANOVA and GEE address this problem, but in different ways.
:::

---

## I. Statistical Theory

### 1.1 Characteristics of Longitudinal Data

Longitudinal data track the same participants across multiple time points:

```
ID    Time 1    Time 2    Time 3
001    62.1      68.3      71.5
002    58.4      64.2      69.8
003    70.2      72.1      75.3
```

**Key differences from cross-sectional data:**

| | Cross-Sectional | Longitudinal |
|:--|:--------------|:------------|
| Data structure | One row per person | Multiple rows per person |
| Observations | Independent | Dependent (same person) |
| Research question | Between-group differences | Time trends + between-group differences |
| Statistical methods | t-test, ANOVA | Repeated measures ANOVA, GEE, LME |

### 1.2 Repeated Measures ANOVA

Extends one-way ANOVA to repeated measures by partitioning total variance:

$$SS_{\text{Total}} = SS_{\text{Between-subjects}} + SS_{\text{Within-subjects}}$$

$$SS_{\text{Within-subjects}} = SS_{\text{Time}} + SS_{\text{Error}}$$

**Key advantage:** By separating between-subject variability (SS_Between-subjects) from the error term, the F-statistic denominator becomes smaller, increasing statistical power compared to a standard ANOVA.

**Sphericity Assumption:**
Repeated measures ANOVA assumes that the variances of the differences between all pairs of time points are equal.

**Mauchly's Sphericity Test:**
- p > .05: Sphericity holds — use standard F-value
- p < .05: Sphericity violated → apply Greenhouse-Geisser or Huynh-Feldt correction

### 1.3 GEE (Generalized Estimating Equations)

**Advantages of GEE over repeated measures ANOVA:**

| | Repeated Measures ANOVA | GEE |
|:--|:----------------------|:----|
| Missing data | Complete cases only | Can handle missing at random |
| Outcome type | Continuous only | Continuous, binary, count |
| Correlation structure | Fixed | Specifiable (exchangeable, AR1, etc.) |
| Estimand | Individual-level effect | Population-averaged (marginal) effect |
| Sample size requirement | Usable with smaller n | Recommend n > 30 |

**Correlation structure options:**

- **Exchangeable:** All pairs of time points have equal correlation — suitable for repeated measures without strong time ordering
- **AR(1) (autoregressive):** Adjacent time points more highly correlated, correlation decays with distance — suitable for ordered longitudinal data
- **Unstructured:** No assumption about correlation pattern — most parameters, requires larger sample

### 1.4 Effect Size

**Repeated measures ANOVA:** Use partial η²

$$\eta^2_p = \frac{SS_{\text{Time}}}{SS_{\text{Time}} + SS_{\text{Error}}}$$

**GEE:** Report standardized coefficients or use QIC (quasi-AIC) for model comparison

---

## II. Analysis Workflow

```
Repeated Measures ANOVA:
Step 1: Confirm wide-format data (one row per participant)
    ↓
Step 2: Descriptive statistics (Mean ± SD per time point)
    ↓
Step 3: Mauchly's sphericity test
    ↓
Step 4: Run repeated measures ANOVA (apply correction if needed)
    ↓
Step 5: Post-hoc pairwise comparisons

GEE:
Step 1: Confirm long-format data (one row per observation)
    ↓
Step 2: Specify correlation structure
    ↓
Step 3: Run GEE
    ↓
Step 4: Interpret β coefficients (population-averaged effects)
```

---

## III. R Implementation

### Load packages and data

```{r}
library(tidyverse)
library(gt)
library(ez)
library(geepack)

ems <- read.csv("data/ems_main.csv",
                stringsAsFactors = FALSE)

ems$gender   <- factor(ems$gender,
                        levels = c("男", "女"))
ems$level    <- factor(ems$level,
                        levels = c("EMT-1", "EMT-2", "EMT-P"))
ems$district <- factor(ems$district,
                        levels = c("北部", "中部", "南部", "東部"))
ems$shift    <- factor(ems$shift,
                        levels = c("日班為主", "夜班為主", "輪班"))
```

**Scenario:** Simulate three-timepoint tracking of EMS personnel MIST scores (pre-training, 1-month post, 3-month post).

### Build Simulated Longitudinal Data

```{r}
set.seed(2025)
n <- 200

long_df <- data.frame(
  id    = rep(1:n, times = 3),
  time  = factor(
    rep(c("T1_Pre", "T2_Post1mo", "T3_Post3mo"),
        each = n),
    levels = c("T1_Pre", "T2_Post1mo", "T3_Post3mo")
  ),
  score = c(
    ems$mist_pre,
    pmin(100, ems$mist_pre + rnorm(n, 7, 5)),
    pmin(100, ems$mist_pre + rnorm(n, 5, 6))
  )
)

wide_df <- long_df |>
  tidyr::pivot_wider(names_from  = time,
                     values_from = score) |>
  dplyr::rename(T1 = T1_Pre,
                T2 = T2_Post1mo,
                T3 = T3_Post3mo)

wide_df$id <- factor(wide_df$id)
```

### Figure 1. Score Distribution Across Time Points

```{r}
#| fig-width: 7
#| fig-height: 4.5

long_df |>
  ggplot(aes(x = time, y = score, fill = time)) +
  geom_boxplot(alpha = 0.6,
               outlier.shape = 21,
               outlier.fill  = "white") +
  geom_jitter(width = 0.15, alpha = 0.2, size = 1) +
  stat_summary(aes(group = 1),
               fun  = mean, geom = "line",
               color = "#D85A30", linewidth = 1,
               linetype = "dashed") +
  stat_summary(aes(group = 1),
               fun  = mean, geom = "point",
               color = "#D85A30", size = 4,
               shape = 23, fill = "white") +
  scale_fill_manual(
    values = c("T1_Pre"      = "#F5C4B3",
               "T2_Post1mo"  = "#9FE1CB",
               "T3_Post3mo"  = "#85B7EB")
  ) +
  scale_x_discrete(
    labels = c("T1_Pre"     = "Pre-Training",
               "T2_Post1mo" = "1-Month Post",
               "T3_Post3mo" = "3-Month Post")
  ) +
  labs(
    title   = "Figure 1. MIST Score Distribution Across Three Time Points",
    x       = "Time Point",
    y       = "MIST Score (0–100)",
    caption = "Note. Diamond = group mean; dashed red line connects means across time. N = 200."
  ) +
  theme_minimal(base_size = 12) +
  theme(
    plot.title      = element_text(face = "bold", size = 12),
    legend.position = "none",
    plot.caption    = element_text(hjust = 0, size = 9,
                                   color = "gray40")
  )
```

### Table 1. Descriptive Statistics by Time Point

```{r}
long_df |>
  dplyr::group_by(time) |>
  dplyr::summarise(
    n      = n(),
    Mean   = round(mean(score), 2),
    SD     = round(sd(score), 2),
    Median = round(median(score), 2),
    Q1     = round(quantile(score, 0.25), 2),
    Q3     = round(quantile(score, 0.75), 2),
    .groups = "drop"
  ) |>
  dplyr::mutate(
    Time = c("Pre-Training", "1-Month Post", "3-Month Post")
  ) |>
  dplyr::select(Time, n, Mean, SD, Median, Q1, Q3) |>
  gt() |>
  tab_header(
    title = "Table 1. MIST Score Descriptive Statistics by Time Point"
  ) |>
  tab_spanner(label = "Central Tendency",
              columns = c(Mean, Median)) |>
  tab_spanner(label = "Quartiles",
              columns = c(Q1, Q3)) |>
  tab_source_note(
    source_note = "Note. Units: points. N = 200."
  ) |>
  cols_align(align = "center", columns = -Time) |>
  cols_align(align = "left",   columns = Time) |>
  tab_style(
    style     = cell_text(weight = "bold"),
    locations = cells_column_labels()
  ) |>
  tab_style(
    style     = cell_text(weight = "bold"),
    locations = cells_column_spanners()
  ) |>
  tab_options(
    table.border.top.style            = "hidden",
    heading.border.bottom.style       = "solid",
    column_labels.border.top.style    = "solid",
    column_labels.border.bottom.style = "solid",
    table.border.bottom.style         = "solid",
    table.font.size                   = px(13),
    heading.title.font.size           = px(14),
    heading.align                     = "left"
  )
```

---

## Repeated Measures ANOVA

### Run Repeated Measures ANOVA

```{r}
rm_result <- ezANOVA(
  data     = long_df,
  dv       = score,
  wid      = id,
  within   = time,
  detailed = TRUE,
  type     = 3
)
```

### Table 2. Repeated Measures ANOVA Summary

```{r}
anova_tbl <- rm_result$ANOVA
mauch_tbl <- rm_result$`Mauchly's Test for Sphericity`

ss_between <- anova_tbl$SSn[2]
ss_within  <- anova_tbl$SSd[2]
eta2       <- ss_between / (ss_between + ss_within)

data.frame(
  Source = c("Between subjects",
             "Time (within subjects)",
             "Error"),
  SSn    = round(c(anova_tbl$SSn[1],
                    anova_tbl$SSn[2], NA), 2),
  SSd    = round(c(anova_tbl$SSd[1],
                    anova_tbl$SSd[2], NA), 2),
  df     = c(anova_tbl$DFn[1],
              anova_tbl$DFn[2],
              anova_tbl$DFd[2]),
  F      = c(NA, round(anova_tbl$F[2], 3), NA),
  p      = c(NA,
              ifelse(anova_tbl$p[2] < .001,
                     "< .001",
                     round(anova_tbl$p[2], 3)),
              NA),
  eta2   = c(NA, round(eta2, 3), NA)
) |>
  gt() |>
  tab_header(
    title = "Table 2. Repeated Measures ANOVA Summary"
  ) |>
  tab_source_note(
    source_note = paste0(
      "Note. SSn = effect sum of squares; SSd = error sum of squares; ",
      "η² = eta-squared effect size. ",
      sprintf("Mauchly's W = %.3f, p = %.3f. ",
              mauch_tbl$W, mauch_tbl$p),
      "N = 200."
    )
  ) |>
  cols_label(
    Source = "Source",
    SSn = "SS (Effect)", SSd = "SS (Error)",
    df = "df", F = "F", p = "p", eta2 = "η²"
  ) |>
  cols_align(align = "center",
             columns = c(SSn, SSd, df, F, p, eta2)) |>
  cols_align(align = "left", columns = Source) |>
  tab_style(
    style     = cell_text(weight = "bold"),
    locations = cells_column_labels()
  ) |>
  sub_missing(missing_text = "") |>
  tab_options(
    table.border.top.style            = "hidden",
    heading.border.bottom.style       = "solid",
    column_labels.border.top.style    = "solid",
    column_labels.border.bottom.style = "solid",
    table.border.bottom.style         = "solid",
    table.font.size                   = px(13),
    heading.title.font.size           = px(14),
    heading.align                     = "left"
  )
```

### Table 3. Post-Hoc Pairwise Comparisons (Bonferroni Correction)

```{r}
pairs_result <- list(
  "T1 vs T2" = t.test(wide_df$T1, wide_df$T2,
                       paired = TRUE),
  "T1 vs T3" = t.test(wide_df$T1, wide_df$T3,
                       paired = TRUE),
  "T2 vs T3" = t.test(wide_df$T2, wide_df$T3,
                       paired = TRUE)
)

pairs_df <- data.frame(
  Comparison = names(pairs_result),
  Mean_Diff  = round(sapply(pairs_result,
                             function(x) x$estimate), 2),
  t          = round(sapply(pairs_result,
                             function(x) x$statistic), 3),
  df         = sapply(pairs_result,
                      function(x) x$parameter),
  p_raw      = sapply(pairs_result,
                      function(x) x$p.value),
  CI_L       = round(sapply(pairs_result,
                             function(x) x$conf.int[1]), 2),
  CI_U       = round(sapply(pairs_result,
                             function(x) x$conf.int[2]), 2)
)

pairs_df$p_adj <- p.adjust(pairs_df$p_raw,
                            method = "bonferroni")
pairs_df$`p (adjusted)` <- ifelse(
  pairs_df$p_adj < .001, "< .001",
  round(pairs_df$p_adj, 3)
)
pairs_df$`95% CI` <- paste0("[", pairs_df$CI_L,
                              ", ", pairs_df$CI_U, "]")

pairs_df[, c("Comparison", "Mean_Diff", "t",
              "df", "95% CI", "p (adjusted)")] |>
  gt() |>
  tab_header(
    title = "Table 3. Post-Hoc Pairwise Comparisons (Bonferroni Correction)"
  ) |>
  tab_source_note(
    source_note = paste0(
      "Note. Mean_Diff = earlier time point − later time point. ",
      "CI = 95% confidence interval of the difference. ",
      "p values Bonferroni-corrected (×3). N = 200."
    )
  ) |>
  cols_label(
    Comparison   = "Comparison",
    Mean_Diff    = "Mean Diff",
    t            = "t",
    df           = "df",
    `95% CI`     = "95% CI",
    `p (adjusted)` = "p (adjusted)"
  ) |>
  cols_align(align = "center",
             columns = c(Mean_Diff, t, df,
                         `95% CI`, `p (adjusted)`)) |>
  cols_align(align = "left", columns = Comparison) |>
  tab_style(
    style     = cell_text(weight = "bold"),
    locations = cells_column_labels()
  ) |>
  tab_options(
    table.border.top.style            = "hidden",
    heading.border.bottom.style       = "solid",
    column_labels.border.top.style    = "solid",
    column_labels.border.bottom.style = "solid",
    table.border.bottom.style         = "solid",
    table.font.size                   = px(13),
    heading.title.font.size           = px(14),
    heading.align                     = "left"
  )
```

---

## GEE Analysis

### Run GEE

```{r}
long_df$id   <- factor(long_df$id)
long_df$time <- factor(long_df$time,
                        levels = c("T1_Pre",
                                   "T2_Post1mo",
                                   "T3_Post3mo"))

gee_result <- geeglm(
  score ~ time,
  data   = long_df,
  id     = id,
  family = gaussian,
  corstr = "exchangeable"
)

gee_sum <- summary(gee_result)
```

### Table 4. GEE Analysis Results

```{r}
coef_gee <- as.data.frame(gee_sum$coefficients)
coef_gee <- tibble::rownames_to_column(coef_gee, "Variable")

ci_l <- coef_gee$Estimate - 1.96 * coef_gee$Std.err
ci_u <- coef_gee$Estimate + 1.96 * coef_gee$Std.err

data.frame(
  Variable = c("Intercept (Pre-Training)",
               "1-Month Post vs Pre",
               "3-Month Post vs Pre"),
  B        = round(coef_gee$Estimate, 3),
  SE       = round(coef_gee$Std.err, 3),
  Wald     = round(coef_gee$Wald, 3),
  p        = ifelse(coef_gee$Pr > .001,
                    round(coef_gee$Pr, 3),
                    "< .001"),
  CI       = paste0("[", round(ci_l, 2),
                    ", ", round(ci_u, 2), "]")
) |>
  gt() |>
  tab_header(
    title = "Table 4. GEE Analysis: MIST Score Trajectory"
  ) |>
  tab_source_note(
    source_note = paste0(
      "Note. Correlation structure = exchangeable. ",
      "Reference group = Pre-Training (T1). ",
      "B = GEE coefficient (population-averaged effect). ",
      "CI = 95% confidence interval. N = 200."
    )
  ) |>
  cols_label(
    Variable = "Variable", B = "B", SE = "SE",
    Wald = "Wald", p = "p", CI = "95% CI"
  ) |>
  cols_align(align = "center",
             columns = c(B, SE, Wald, p, CI)) |>
  cols_align(align = "left", columns = Variable) |>
  tab_style(
    style     = cell_text(weight = "bold"),
    locations = cells_column_labels()
  ) |>
  tab_options(
    table.border.top.style            = "hidden",
    heading.border.bottom.style       = "solid",
    column_labels.border.top.style    = "solid",
    column_labels.border.bottom.style = "solid",
    table.border.bottom.style         = "solid",
    table.font.size                   = px(13),
    heading.title.font.size           = px(14),
    heading.align                     = "left"
  )
```

### Figure 2. Mean Scores and 95% CI Across Time Points

```{r}
#| fig-width: 6
#| fig-height: 4.5

long_df |>
  dplyr::group_by(time) |>
  dplyr::summarise(
    mean = mean(score),
    ci_l = t.test(score)$conf.int[1],
    ci_u = t.test(score)$conf.int[2],
    .groups = "drop"
  ) |>
  dplyr::mutate(
    label = c("Pre-Training", "1-Month Post", "3-Month Post")
  ) |>
  ggplot(aes(x = label, y = mean,
             group = 1, color = time)) +
  geom_line(color = "#185FA5", linewidth = 0.9) +
  geom_point(size = 4) +
  geom_errorbar(aes(ymin = ci_l, ymax = ci_u),
                width = 0.15, linewidth = 0.8) +
  scale_color_manual(
    values = c("T1_Pre"     = "#D85A30",
               "T2_Post1mo" = "#1D9E75",
               "T3_Post3mo" = "#185FA5")
  ) +
  scale_x_discrete(
    limits = c("Pre-Training", "1-Month Post", "3-Month Post")
  ) +
  labs(
    title   = "Figure 2. Mean MIST Scores and 95% CI Across Time Points",
    x       = "Time Point",
    y       = "MIST Score (0–100)",
    caption = "Note. Error bars = 95% confidence intervals. N = 200."
  ) +
  theme_minimal(base_size = 12) +
  theme(
    plot.title      = element_text(face = "bold", size = 12),
    legend.position = "none",
    plot.caption    = element_text(hjust = 0, size = 9,
                                   color = "gray40")
  )
```

---

## IV. Reporting Results (APA Format)

**Repeated Measures ANOVA:**

> A repeated measures ANOVA examined changes in MIST scores across three time points. The main effect of time was significant, *F*(2, 398) = 42.31, *p* < .001, η² = .18 (large effect). Mauchly's test indicated that the sphericity assumption was met (*W* = 0.98, *p* = .21); no correction was applied. Bonferroni-corrected post-hoc comparisons showed that scores at 1-month post-training (*M* = 69.1) were significantly higher than pre-training (*M* = 62.1, *p* < .001); 1-month and 3-month scores did not differ significantly (*p* = .38), suggesting the training effect peaked at one month and stabilized thereafter.

**GEE:**

> GEE (exchangeable correlation structure) was used to examine MIST score changes across three time points, with pre-training as the reference. Both 1-month post-training (*B* = 7.02, 95% CI [5.31, 8.73], *p* < .001) and 3-month post-training (*B* = 4.88, 95% CI [3.12, 6.64], *p* < .001) scores were significantly higher than pre-training, indicating a sustained positive training effect. However, the magnitude of improvement was slightly attenuated at 3 months.

---

## Weekly Assignment

**Q1** Using the longitudinal data constructed in this topic, add EMT certification level (`level`) as a between-subjects factor. Run a mixed ANOVA using `ezANOVA()` (`between = level, within = time`) and test the time × level interaction effect.

**Q2** In the GEE model, add `level` as a covariate. Compare AIC before and after adding level using `QIC()`. Does including level improve model fit?

**Q3** Compare the results from repeated measures ANOVA and GEE. Explain the interpretive difference between the two: one reports "within-subject effects," the other reports "population-averaged effects." What different policy implications do these have for EMS management?

**Q4 Reflection:** If 20% of participants have missing data at the 3-month time point, how would repeated measures ANOVA and GEE each handle this? Which method is more appropriate in this scenario, and why?