---
title: "Topic 13：Multivariate Analysis of Variance (MANOVA)"
subtitle: "Multivariate Test Statistics, Assumption Checks, Follow-Up ANOVAs, APA Reporting"
execute:
  warning: false
  message: false
  echo: true
---

## Learning Objectives

1. Understand the statistical logic of MANOVA and when to use it
2. Execute MANOVA and interpret multivariate test statistics
3. Conduct follow-up univariate ANOVAs with appropriate correction
4. Produce APA-format result tables and reporting sentences

---

## Statistical Method Map (This Topic)

| X (Predictor) | Y (Outcome) | Method |
|:-------------|:-----------|:-------|
| Categorical (≥ 2 groups) | Continuous (1 variable) | ANOVA (Topic 7) |
| **Categorical (≥ 2 groups)** | **Continuous (multiple variables)** | **MANOVA** |

::: {.callout-note}
**Why not run separate ANOVAs for each Y?**

With 5 dependent variables, running five ANOVAs at α = .05 inflates the familywise Type I error rate to 1 − (0.95)⁵ = 22.6%. MANOVA tests all dependent variables simultaneously, controlling the overall error rate at α = .05. Moreover, MANOVA can detect patterns shared across dependent variables that separate ANOVAs cannot capture.
:::

---

## I. Statistical Theory

### 1.1 The Core Logic of MANOVA

MANOVA extends ANOVA to multiple dependent variables:

**ANOVA:** Compares groups on a single Y variable

**MANOVA:** Compares groups on a **vector** of multiple Y variables simultaneously

$$\mathbf{Y}_{p \times 1} = \boldsymbol{\mu} + \text{Group effect} + \boldsymbol{\varepsilon}$$

where **Y** is a vector of p dependent variables.

### 1.2 Multivariate Test Statistics

MANOVA offers four commonly used test statistics for testing H₀: all group mean vectors are equal across all Y variables:

| Statistic | Description | Best Used When |
|:----------|:-----------|:--------------|
| Pillai's Trace | Most robust; insensitive to assumption violations | **Default recommendation** |
| Wilks' Lambda | Most commonly cited (close to 1 = no difference) | Normal data, no outliers |
| Hotelling's Trace | Sensitive to outliers | Comparing two groups |
| Roy's Largest Root | Highest power but least robust | Single dominant component |

::: {.callout-tip}
**Recommendation:** Default to Pillai's Trace — it is the most robust to violations of normality and homogeneity assumptions. When all four statistics agree, the results are more trustworthy.
:::

### 1.3 Assumptions of MANOVA

| Assumption | Description | How to Check |
|:-----------|:-----------|:------------|
| Multivariate normality | All Y combinations follow multivariate normality | Mardia's test; Q-Q plots |
| Homogeneity of covariance matrices | Each group has equal covariance matrices | Box's M test |
| No severe multivariate outliers | No high-influence observations | Mahalanobis distance |
| Adequate inter-correlations among DVs | Y variables not completely unrelated or collinear | Correlation matrix |

::: {.callout-important}
**Box's M test is extremely sensitive:**
With large samples, it is almost always significant (p < .05). Do not abandon MANOVA just because Box's M is significant. Pillai's Trace maintains reasonable robustness even when covariance matrices are unequal.
:::

### 1.4 Follow-Up Analysis

After a significant MANOVA, identify which DVs differ across groups:

**Option A:** Run separate univariate ANOVAs for each DV
(Use Bonferroni correction: α = .05/p, where p = number of DVs)

**Option B:** Discriminant analysis to find the linear combination of DVs that best separates groups

**This course uses Option A** (more straightforward to interpret).

### 1.5 Effect Size

**Multivariate effect size:** Pillai's Trace itself serves as an effect size (range 0–1)

**Univariate effect size:** Follow-up ANOVAs use η²

| η² | Strength |
|:---|:--------|
| .01–.05 | Small |
| .06–.13 | Medium |
| ≥ .14 | Large |

---

## II. Analysis Workflow

```
Step 1: Confirm adequate inter-correlations among DVs
    ↓
Step 2: Check multivariate outliers (Mahalanobis distance)
    ↓
Step 3: Box's M test (covariance homogeneity)
    ↓
Step 4: Run MANOVA (Pillai's Trace)
    ↓
Step 5: If significant → follow-up univariate ANOVAs (Bonferroni α = .05/p)
    ↓
Step 6: Post-hoc pairwise comparisons
    ↓
Step 7: APA-format report
```

---

## III. R Implementation

### Load packages and data

```{r}
library(tidyverse)
library(gt)
library(car)

ems <- read.csv("data/ems_main.csv",
                stringsAsFactors = FALSE)

ems$gender   <- factor(ems$gender,
                        levels = c("男", "女"))
ems$level    <- factor(ems$level,
                        levels = c("EMT-1", "EMT-2", "EMT-P"))
ems$district <- factor(ems$district,
                        levels = c("北部", "中部", "南部", "東部"))
ems$shift    <- factor(ems$shift,
                        levels = c("日班為主", "夜班為主", "輪班"))
```

**Research question:** Do the three EMT certification levels differ significantly across three occupational health indicators — emotional exhaustion, depersonalization, and job satisfaction — considered simultaneously?

### Figure 1. Distribution of All DVs by EMT Level

```{r}
#| fig-width: 9
#| fig-height: 4

ems |>
  dplyr::select(level, mbi_ee, mbi_dp, job_sat) |>
  tidyr::pivot_longer(c(mbi_ee, mbi_dp, job_sat),
                      names_to  = "variable",
                      values_to = "value") |>
  dplyr::mutate(
    variable = factor(variable,
      levels = c("mbi_ee", "mbi_dp", "job_sat"),
      labels = c("Emotional Exhaustion (MBI-EE)",
                 "Depersonalization (MBI-DP)",
                 "Job Satisfaction"))
  ) |>
  ggplot(aes(x = level, y = value, fill = level)) +
  geom_boxplot(alpha = 0.6,
               outlier.shape = 21,
               outlier.fill  = "white") +
  facet_wrap(~ variable, scales = "free_y") +
  scale_fill_manual(
    values = c("EMT-1" = "#F5C4B3",
               "EMT-2" = "#9FE1CB",
               "EMT-P" = "#85B7EB")
  ) +
  labs(
    title   = "Figure 1. Distribution of Occupational Health Indicators by EMT Level",
    x       = "EMT Certification Level",
    y       = "Score",
    caption = "Note. N = 200."
  ) +
  theme_minimal(base_size = 11) +
  theme(
    plot.title      = element_text(face = "bold", size = 12),
    legend.position = "none",
    plot.caption    = element_text(hjust = 0, size = 9,
                                   color = "gray40")
  )
```

### Table 1. Descriptive Statistics by EMT Level

```{r}
vars_list <- list(
  "Emotional Exhaustion (MBI-EE)" = "mbi_ee",
  "Depersonalization (MBI-DP)"    = "mbi_dp",
  "Job Satisfaction"              = "job_sat"
)

desc_rows <- list()
for (vname in names(vars_list)) {
  vcol <- vars_list[[vname]]
  row  <- ems |>
    dplyr::group_by(level) |>
    dplyr::summarise(
      DV   = vname,
      Mean = round(mean(.data[[vcol]]), 2),
      SD   = round(sd(.data[[vcol]]), 2),
      .groups = "drop"
    ) |>
    dplyr::rename(Level = level)
  desc_rows[[vname]] <- row
}

dplyr::bind_rows(desc_rows) |>
  dplyr::select(DV, Level, Mean, SD) |>
  gt() |>
  tab_header(
    title = "Table 1. Descriptive Statistics by EMT Level for Each Dependent Variable"
  ) |>
  tab_source_note(
    source_note = "Note. N = 200. MBI = Maslach Burnout Inventory."
  ) |>
  cols_align(align = "center", columns = c(Mean, SD)) |>
  cols_align(align = "left",
             columns = c(DV, Level)) |>
  tab_style(
    style     = cell_text(weight = "bold"),
    locations = cells_column_labels()
  ) |>
  tab_options(
    table.border.top.style            = "hidden",
    heading.border.bottom.style       = "solid",
    column_labels.border.top.style    = "solid",
    column_labels.border.bottom.style = "solid",
    table.border.bottom.style         = "solid",
    table.font.size                   = px(13),
    heading.title.font.size           = px(14),
    heading.align                     = "left"
  )
```

### Assumption Check: DV Correlation Matrix

```{r}
cor_dv <- cor(ems[, c("mbi_ee", "mbi_dp", "job_sat")],
              use = "complete.obs")
rownames(cor_dv) <- c("Emotional Exhaustion",
                       "Depersonalization",
                       "Job Satisfaction")
colnames(cor_dv) <- rownames(cor_dv)

as.data.frame(round(cor_dv, 3)) |>
  tibble::rownames_to_column("Variable") |>
  gt() |>
  tab_header(
    title = "Table 2. Correlation Matrix Among Dependent Variables"
  ) |>
  tab_source_note(
    source_note = paste0(
      "Note. DVs should be moderately correlated (|r| = .20–.80). ",
      "Very low correlations suggest MANOVA offers no advantage over separate ANOVAs; ",
      "very high correlations (multicollinearity) suggest merging or removing variables."
    )
  ) |>
  cols_align(align = "center",
             columns = c(`Emotional Exhaustion`,
                         `Depersonalization`,
                         `Job Satisfaction`)) |>
  cols_align(align = "left", columns = Variable) |>
  tab_style(
    style     = cell_text(weight = "bold"),
    locations = cells_column_labels()
  ) |>
  tab_options(
    table.border.top.style            = "hidden",
    heading.border.bottom.style       = "solid",
    column_labels.border.top.style    = "solid",
    column_labels.border.bottom.style = "solid",
    table.border.bottom.style         = "solid",
    table.font.size                   = px(13),
    heading.title.font.size           = px(14),
    heading.align                     = "left"
  )
```

### Assumption Check: Multivariate Outliers

```{r}
#| fig-width: 6
#| fig-height: 4

dv_matrix  <- as.matrix(ems[, c("mbi_ee",
                                  "mbi_dp",
                                  "job_sat")])
mah_dist   <- mahalanobis(dv_matrix,
                           colMeans(dv_matrix),
                           cov(dv_matrix))
mah_cutoff <- qchisq(0.999, df = 3)
n_outliers <- sum(mah_dist > mah_cutoff)

data.frame(id   = 1:nrow(ems),
           mah  = mah_dist,
           flag = mah_dist > mah_cutoff) |>
  ggplot(aes(x = id, y = mah, color = flag)) +
  geom_point(alpha = 0.6, size = 1.5) +
  geom_hline(yintercept = mah_cutoff,
             linetype = "dashed",
             color    = "#D85A30",
             linewidth = 0.8) +
  scale_color_manual(
    values = c("FALSE" = "#85B7EB",
               "TRUE"  = "#D85A30"),
    name = "Outlier"
  ) +
  annotate("text", x = 175,
           y = mah_cutoff + 0.5,
           label = paste0("χ²₀.₀₀₁ = ",
                          round(mah_cutoff, 1)),
           color = "#D85A30", size = 3) +
  labs(
    title   = "Figure 2. Mahalanobis Distance Diagnostic Plot",
    x       = "Observation Index",
    y       = "Mahalanobis Distance",
    caption = paste0("Note. Red points = multivariate outliers (n = ",
                     n_outliers, "). N = 200.")
  ) +
  theme_minimal(base_size = 12) +
  theme(
    plot.title   = element_text(face = "bold", size = 12),
    plot.caption = element_text(hjust = 0, size = 9,
                                color = "gray40")
  )
```

### Run MANOVA

```{r}
manova_result <- manova(
  cbind(mbi_ee, mbi_dp, job_sat) ~ level,
  data = ems
)
```

### Table 3. MANOVA Multivariate Test Statistics

```{r}
pill  <- summary(manova_result, test = "Pillai")$stats
wilk  <- summary(manova_result, test = "Wilks")$stats
hotel <- summary(manova_result,
                  test = "Hotelling-Lawley")$stats
roy   <- summary(manova_result, test = "Roy")$stats

data.frame(
  Statistic = c("Pillai's Trace",
                "Wilks' Lambda",
                "Hotelling's Trace",
                "Roy's Largest Root"),
  Value     = round(c(pill[1,  "Pillai"],
                       wilk[1,  "Wilks"],
                       hotel[1, "approx F"],
                       roy[1,   "approx F"]), 4),
  F         = round(c(pill[1,  "approx F"],
                       wilk[1,  "approx F"],
                       hotel[1, "approx F"],
                       roy[1,   "approx F"]), 3),
  df1       = c(pill[1,  "num Df"],
                wilk[1,  "num Df"],
                hotel[1, "num Df"],
                roy[1,   "num Df"]),
  df2       = c(pill[1,  "den Df"],
                wilk[1,  "den Df"],
                hotel[1, "den Df"],
                roy[1,   "den Df"]),
  p         = ifelse(
    c(pill[1,  "Pr(>F)"], wilk[1,  "Pr(>F)"],
      hotel[1, "Pr(>F)"], roy[1,   "Pr(>F)"]) < .001,
    "< .001",
    round(c(pill[1,  "Pr(>F)"], wilk[1,  "Pr(>F)"],
            hotel[1, "Pr(>F)"], roy[1,   "Pr(>F)"]), 3)
  )
) |>
  gt() |>
  tab_header(
    title = "Table 3. MANOVA Multivariate Test Statistics"
  ) |>
  tab_source_note(
    source_note = paste0(
      "Note. DVs: emotional exhaustion, depersonalization, job satisfaction. ",
      "IV: EMT certification level (3 groups). ",
      "Pillai's Trace is recommended as the primary statistic (most robust). ",
      "N = 200."
    )
  ) |>
  cols_label(
    Statistic = "Test Statistic",
    Value     = "Value",
    F         = "F",
    df1       = "df1",
    df2       = "df2",
    p         = "p"
  ) |>
  cols_align(align = "center",
             columns = c(Value, F, df1, df2, p)) |>
  cols_align(align = "left", columns = Statistic) |>
  tab_style(
    style     = cell_text(weight = "bold"),
    locations = cells_column_labels()
  ) |>
  tab_options(
    table.border.top.style            = "hidden",
    heading.border.bottom.style       = "solid",
    column_labels.border.top.style    = "solid",
    column_labels.border.bottom.style = "solid",
    table.border.bottom.style         = "solid",
    table.font.size                   = px(13),
    heading.title.font.size           = px(14),
    heading.align                     = "left"
  )
```

### Table 4. Follow-Up Univariate ANOVAs (Bonferroni α = .017)

```{r}
uni_results <- summary.aov(manova_result)

get_anova_row <- function(aov_sum, yname) {
  ss_b  <- aov_sum[[1]]$`Sum Sq`[1]
  ss_w  <- aov_sum[[1]]$`Sum Sq`[2]
  df_b  <- aov_sum[[1]]$Df[1]
  df_w  <- aov_sum[[1]]$Df[2]
  f_val <- aov_sum[[1]]$`F value`[1]
  p_val <- aov_sum[[1]]$`Pr(>F)`[1]
  eta2  <- ss_b / (ss_b + ss_w)
  data.frame(
    DV     = yname,
    F      = round(f_val, 3),
    df1    = df_b,
    df2    = df_w,
    p      = ifelse(p_val < .001, "< .001",
                    round(p_val, 3)),
    eta2   = round(eta2, 3),
    Sig    = ifelse(p_val < .05/3, "✓*", "—")
  )
}

dplyr::bind_rows(
  get_anova_row(uni_results[1],
                "Emotional Exhaustion (MBI-EE)"),
  get_anova_row(uni_results[2],
                "Depersonalization (MBI-DP)"),
  get_anova_row(uni_results[3],
                "Job Satisfaction")
) |>
  gt() |>
  tab_header(
    title = "Table 4. Follow-Up Univariate ANOVA Results (Bonferroni Correction)"
  ) |>
  tab_source_note(
    source_note = paste0(
      "Note. Bonferroni-corrected significance threshold: α = .05/3 = .017. ",
      "η² = eta-squared effect size. ",
      "✓* = p < .017 (significant); — = not significant. N = 200."
    )
  ) |>
  cols_label(
    DV   = "Dependent Variable",
    F    = "F", df1 = "df1", df2 = "df2",
    p    = "p", eta2 = "η²", Sig = "Significant"
  ) |>
  cols_align(align = "center",
             columns = c(F, df1, df2, p, eta2, Sig)) |>
  cols_align(align = "left", columns = DV) |>
  tab_style(
    style     = cell_text(weight = "bold"),
    locations = cells_column_labels()
  ) |>
  tab_options(
    table.border.top.style            = "hidden",
    heading.border.bottom.style       = "solid",
    column_labels.border.top.style    = "solid",
    column_labels.border.bottom.style = "solid",
    table.border.bottom.style         = "solid",
    table.font.size                   = px(13),
    heading.title.font.size           = px(14),
    heading.align                     = "left"
  )
```

---

## IV. Reporting Results (APA Format)

**MANOVA main effect:**

> MANOVA was used to simultaneously examine group differences across emotional exhaustion, depersonalization, and job satisfaction. The overall effect of EMT certification level was significant, Pillai's Trace = 0.18, *F*(6, 392) = 6.43, *p* < .001.

**Follow-up univariate ANOVAs:**

> Following Bonferroni correction (α = .017), univariate ANOVAs revealed a significant effect of certification level on emotional exhaustion, *F*(2, 197) = 8.21, *p* < .001, η² = .08 (medium effect), and on job satisfaction, *F*(2, 197) = 6.14, *p* = .003, η² = .06 (medium effect). The effect on depersonalization did not reach the Bonferroni-corrected threshold (*p* = .042).

---

## Weekly Assignment

**Q1** Using shift type (`shift`, 3 groups) as the independent variable and `stress_vas`, `mbi_ee`, and `sleep_hrs` as dependent variables, run a complete MANOVA analysis including: assumption checks, multivariate test statistics table, and follow-up univariate ANOVA table. All tables in APA `gt` format.

**Q2** Examine the DV correlation matrix for Q1. Are these three variables appropriate for MANOVA? If correlations are very low (|r| < .20), what advantage does MANOVA offer over running separate ANOVAs?

**Q3** Compute Mahalanobis distances for Q1, remove multivariate outliers, and re-run MANOVA. Compare Pillai's Trace and p-values before and after outlier removal. Do the conclusions change substantially?

**Q4 Reflection:** After a significant MANOVA, someone concludes "all three EMT levels differ on all mental health indicators." Is this statement correct? How do the follow-up univariate ANOVA results revise this conclusion? Why is Bonferroni correction necessary?