---
title: "Topic 15：Final Review & Practice Exam"
subtitle: "Method Integration, X-Y Decision Table, Exam Preparation"
execute:
  warning: false
  message: false
  echo: true
---

## Learning Objectives

1. Integrate all course methods and rapidly select the correct approach from the X-Y decision table
2. Practice the complete analysis pipeline: assumption check → execute → report
3. Familiarize yourself with the final exam data format and question types
4. Review APA-format reporting standards

---

## I. Full-Course Statistical Method Summary

### X-Y Decision Table

This table is the core map of the entire course. In the final exam, your first step is always to identify the types of X and Y.

```{r}
library(tidyverse)
library(gt)

decision_tbl <- data.frame(
  Topic  = c(
    "Topic 2", "Topic 2",
    "Topic 3", "Topic 4",
    "Topic 5", "Topic 5",
    "Topic 6", "Topic 6",
    "Topic 7", "Topic 8",
    "Topic 8", "Topic 9",
    "Topic 9", "Topic 10",
    "Topic 11", "Topic 12",
    "Topic 12", "Topic 13",
    "Topic 14"
  ),
  X = c(
    "None (univariate)", "None (univariate)",
    "None", "—",
    "Continuous (normal)", "Continuous (skewed)",
    "Categorical (2 groups, independent)",
    "Categorical (2 groups, paired)",
    "Categorical (≥ 3 groups)",
    "Categorical (2 groups)",
    "Categorical (≥ 3 groups)",
    "Continuous", "Continuous / Categorical",
    "Continuous / Categorical", "Continuous",
    "Repeated time points", "Repeated time points",
    "Categorical (≥ 2 groups)", "Latent variable"
  ),
  Y = c(
    "Continuous", "Categorical",
    "Continuous / Categorical", "—",
    "Continuous", "Continuous",
    "Continuous", "Continuous",
    "Continuous", "Continuous (skewed)",
    "Continuous (skewed)", "Continuous",
    "Continuous", "Binary",
    "Continuous", "Continuous",
    "Continuous", "Continuous (multiple)",
    "Latent variable"
  ),
  Method = c(
    "Mean, SD, Median, IQR",
    "Frequency, %, RR, OR",
    "Histogram, Boxplot, Scatterplot (ggplot2)",
    "Normal distribution, CLT, SE, CI",
    "Pearson Correlation",
    "Spearman Correlation",
    "Independent t-test (Welch)",
    "Paired t-test",
    "One-way ANOVA + Tukey HSD",
    "Mann-Whitney U / Wilcoxon",
    "Kruskal-Wallis + Dunn's test",
    "Simple / Multiple Linear Regression",
    "Multiple Regression (categorical predictors)",
    "Binary Logistic Regression",
    "Confounding & Mediation (Bootstrap)",
    "Repeated Measures ANOVA",
    "GEE",
    "MANOVA",
    "SEM (CFA + Path Analysis)"
  ),
  Nonparametric = c(
    "—", "—", "—", "—",
    "Spearman", "—",
    "Mann-Whitney U", "Wilcoxon Signed-Rank",
    "Kruskal-Wallis", "—", "—",
    "—", "—", "—", "—",
    "GEE", "Repeated Measures ANOVA",
    "—", "—"
  )
)

decision_tbl |>
  gt() |>
  tab_header(
    title = "Table 1. Full-Course Statistical Method X-Y Decision Table"
  ) |>
  tab_source_note(
    source_note = paste0(
      "Note. First step: identify the measurement scale of X and Y. ",
      "When normality assumptions are violated, use nonparametric alternatives."
    )
  ) |>
  cols_label(
    Topic        = "Topic",
    X            = "X (Predictor)",
    Y            = "Y (Outcome)",
    Method       = "Statistical Method",
    Nonparametric = "Nonparametric Alt."
  ) |>
  cols_align(align = "center",
             columns = c(Topic, Nonparametric)) |>
  cols_align(align = "left",
             columns = c(X, Y, Method)) |>
  tab_style(
    style     = cell_text(weight = "bold"),
    locations = cells_column_labels()
  ) |>
  tab_row_group(
    label = "Advanced Methods",
    rows  = 15:19
  ) |>
  tab_row_group(
    label = "Inferential Statistics",
    rows  = 7:14
  ) |>
  tab_row_group(
    label = "Descriptive Statistics & Visualization",
    rows  = 1:6
  ) |>
  tab_options(
    table.border.top.style            = "hidden",
    heading.border.bottom.style       = "solid",
    column_labels.border.top.style    = "solid",
    column_labels.border.bottom.style = "solid",
    table.border.bottom.style         = "solid",
    table.font.size                   = px(13),
    heading.align                     = "left",
    row_group.font.weight             = "bold"
  )
```

---

## II. Assumption Checklist

```{r}
assume_tbl <- data.frame(
  Method = c(
    "Independent t-test",
    "Paired t-test",
    "One-way ANOVA",
    "Linear Regression",
    "Logistic Regression",
    "Repeated Measures ANOVA",
    "MANOVA"
  ),
  Assumptions = c(
    "Normality, independence",
    "Normality of difference scores",
    "Normality, homogeneity of variance, independence",
    "Linearity, residual normality, homoscedasticity, independence",
    "Adequate sample size (≥ 10 per group)",
    "Sphericity (Mauchly's test)",
    "Multivariate normality, homogeneity of covariance matrices"
  ),
  How_to_Check = c(
    "Shapiro-Wilk; Levene's test",
    "Shapiro-Wilk (on difference scores)",
    "Shapiro-Wilk; Levene's test",
    "Residual plots; Q-Q plot; VIF",
    "Sample size check; Hosmer-Lemeshow",
    "Mauchly's test",
    "Box's M test; Mahalanobis distance"
  ),
  If_Violated = c(
    "Mann-Whitney U",
    "Wilcoxon Signed-Rank",
    "Kruskal-Wallis",
    "Robust regression; data transformation",
    "Simplify model; increase sample size",
    "Greenhouse-Geisser correction; GEE",
    "Pillai's Trace (most robust)"
  )
)

assume_tbl |>
  gt() |>
  tab_header(
    title = "Table 2. Assumption Checklist by Method"
  ) |>
  tab_source_note(
    source_note = "Note. When n ≥ 30, CLT protects t-tests and ANOVA against violations of normality."
  ) |>
  cols_label(
    Method       = "Method",
    Assumptions  = "Key Assumptions",
    How_to_Check = "How to Check",
    If_Violated  = "If Violated"
  ) |>
  cols_align(align = "left", columns = everything()) |>
  tab_style(
    style     = cell_text(weight = "bold"),
    locations = cells_column_labels()
  ) |>
  tab_options(
    table.border.top.style            = "hidden",
    heading.border.bottom.style       = "solid",
    column_labels.border.top.style    = "solid",
    column_labels.border.bottom.style = "solid",
    table.border.bottom.style         = "solid",
    table.font.size                   = px(13),
    heading.align                     = "left"
  )
```

---

## III. Effect Size Quick Reference

```{r}
effect_tbl <- data.frame(
  Method     = c(
    "t-test",
    "ANOVA",
    "Correlation",
    "Logistic Regression",
    "Mann-Whitney U",
    "Kruskal-Wallis",
    "MANOVA",
    "Linear Regression"
  ),
  Index      = c(
    "Cohen's d",
    "η² (Eta-squared)",
    "r (correlation coefficient)",
    "Nagelkerke R²",
    "r = Z / √N",
    "ε² (Epsilon-squared)",
    "Pillai's Trace",
    "R², Adjusted R²"
  ),
  Small      = c("0.20", ".01", ".10",
                  "—", ".10", ".01", "—", ".02"),
  Medium     = c("0.50", ".06", ".30",
                  "—", ".30", ".06", "—", ".13"),
  Large      = c("0.80", ".14", ".50",
                  "—", ".50", ".14", "—", ".26"),
  R_Code     = c(
    "manual: d = |Δμ| / sp",
    "manual: SSb / SSt",
    "cor.test()$estimate",
    "manual Nagelkerke formula",
    "abs(qnorm(p/2)) / sqrt(N)",
    "manual: (H-k+1)/(N-k)",
    "summary(manova)$stats",
    "summary(lm)$r.squared"
  )
)

effect_tbl |>
  gt() |>
  tab_header(
    title = "Table 3. Effect Size Quick Reference"
  ) |>
  tab_source_note(
    source_note = "Note. Effect size criteria from Cohen (1988). Always report effect size alongside p-values."
  ) |>
  cols_label(
    Method  = "Method",
    Index   = "Effect Size Index",
    Small   = "Small",
    Medium  = "Medium",
    Large   = "Large",
    R_Code  = "R Syntax"
  ) |>
  cols_align(align = "center",
             columns = c(Small, Medium, Large)) |>
  cols_align(align = "left",
             columns = c(Method, Index, R_Code)) |>
  tab_style(
    style     = cell_text(weight = "bold"),
    locations = cells_column_labels()
  ) |>
  tab_options(
    table.border.top.style            = "hidden",
    heading.border.bottom.style       = "solid",
    column_labels.border.top.style    = "solid",
    column_labels.border.bottom.style = "solid",
    table.border.bottom.style         = "solid",
    table.font.size                   = px(13),
    heading.align                     = "left"
  )
```

---

## IV. Practice Exercises

### Load data

```{r}
library(gt)
library(car)
library(rstatix)
library(pROC)
library(ResourceSelection)

ems <- read.csv("data/ems_main.csv",
                stringsAsFactors = FALSE)

ems$gender   <- factor(ems$gender,
                        levels = c("男", "女"))
ems$level    <- factor(ems$level,
                        levels = c("EMT-1", "EMT-2", "EMT-P"))
ems$district <- factor(ems$district,
                        levels = c("北部", "中部", "南部", "東部"))
ems$shift    <- factor(ems$shift,
                        levels = c("日班為主", "夜班為主", "輪班"))
```

### Practice 1: Descriptive Statistics

```{r}
library(psych)

desc_raw <- describe(
  ems |> dplyr::select(age, years, monthly_cases,
                        stress_vas, mbi_ee,
                        response_time)
)

data.frame(
  Variable = c("Age (years)", "Years of service",
               "Monthly calls", "Stress VAS",
               "MBI-EE", "Response time (min)"),
  n        = as.integer(desc_raw$n),
  Mean     = round(desc_raw$mean, 2),
  SD       = round(desc_raw$sd, 2),
  Median   = round(desc_raw$median, 2),
  Skewness = round(desc_raw$skew, 2),
  Recommendation = ifelse(
    abs(desc_raw$skew) > 0.5,
    "Median (IQR)", "Mean (SD)")
) |>
  gt() |>
  tab_header(
    title = "Practice 1. Descriptive Statistics for Key Continuous Variables"
  ) |>
  tab_source_note(
    source_note = "Note. |Skewness| > 0.5 suggests reporting Median (IQR). N = 200."
  ) |>
  cols_align(align = "center",
             columns = c(n, Mean, SD, Median, Skewness)) |>
  cols_align(align = "left",
             columns = c(Variable, Recommendation)) |>
  tab_style(
    style     = cell_text(weight = "bold"),
    locations = cells_column_labels()
  ) |>
  tab_options(
    table.border.top.style            = "hidden",
    heading.border.bottom.style       = "solid",
    column_labels.border.top.style    = "solid",
    column_labels.border.bottom.style = "solid",
    table.border.bottom.style         = "solid",
    table.font.size                   = px(13),
    heading.align                     = "left"
  )
```

### Practice 2a: Independent t-Test

```{r}
t_result <- t.test(stress_vas ~ gender,
                   data      = ems,
                   var.equal = FALSE)

m1 <- mean(ems$stress_vas[ems$gender == "男"])
m2 <- mean(ems$stress_vas[ems$gender == "女"])
s1 <- sd(ems$stress_vas[ems$gender == "男"])
s2 <- sd(ems$stress_vas[ems$gender == "女"])
n1 <- sum(ems$gender == "男")
n2 <- sum(ems$gender == "女")
sp <- sqrt(((n1-1)*s1^2 + (n2-1)*s2^2) / (n1+n2-2))
d_val <- abs((m1 - m2) / sp)

data.frame(
  Comparison = "Male vs Female (Stress VAS)",
  t          = round(t_result$statistic, 3),
  df         = round(t_result$parameter, 1),
  p          = ifelse(t_result$p.value < .001,
                      "< .001",
                      round(t_result$p.value, 3)),
  d          = round(d_val, 3),
  CI         = paste0("[",
                 round(t_result$conf.int[1], 2), ", ",
                 round(t_result$conf.int[2], 2), "]"),
  Effect     = dplyr::case_when(
    d_val >= .80 ~ "Large",
    d_val >= .50 ~ "Medium",
    TRUE         ~ "Small"
  )
) |>
  gt() |>
  tab_header(
    title = "Practice 2a. Independent t-Test: Perceived Stress by Sex"
  ) |>
  tab_source_note(
    source_note = "Note. Welch t-test. d = Cohen's d. CI = 95% CI of mean difference. N = 200."
  ) |>
  cols_label(
    Comparison = "Comparison", t = "t", df = "df",
    p = "p", d = "d", CI = "95% CI", Effect = "Effect Size"
  ) |>
  cols_align(align = "center",
             columns = c(t, df, p, d, CI, Effect)) |>
  cols_align(align = "left", columns = Comparison) |>
  tab_style(
    style     = cell_text(weight = "bold"),
    locations = cells_column_labels()
  ) |>
  tab_options(
    table.border.top.style            = "hidden",
    heading.border.bottom.style       = "solid",
    column_labels.border.top.style    = "solid",
    column_labels.border.bottom.style = "solid",
    table.border.bottom.style         = "solid",
    table.font.size                   = px(13),
    heading.align                     = "left"
  )
```

### Practice 2b: One-Way ANOVA

```{r}
aov_r   <- aov(response_time ~ level, data = ems)
aov_s   <- summary(aov_r)
ss_b    <- aov_s[[1]]$`Sum Sq`[1]
ss_w    <- aov_s[[1]]$`Sum Sq`[2]
eta2    <- ss_b / (ss_b + ss_w)

data.frame(
  Source = c("Between groups (Level)",
             "Within groups (Error)",
             "Total"),
  SS     = round(c(ss_b, ss_w, ss_b + ss_w), 2),
  df     = c(aov_s[[1]]$Df[1],
              aov_s[[1]]$Df[2],
              sum(aov_s[[1]]$Df)),
  MS     = c(round(aov_s[[1]]$`Mean Sq`[1], 2),
              round(aov_s[[1]]$`Mean Sq`[2], 2), NA),
  F      = c(round(aov_s[[1]]$`F value`[1], 3),
              NA, NA),
  p      = c(ifelse(aov_s[[1]]$`Pr(>F)`[1] < .001,
                    "< .001",
                    round(aov_s[[1]]$`Pr(>F)`[1], 3)),
              NA, NA),
  eta2   = c(round(eta2, 3), NA, NA)
) |>
  gt() |>
  tab_header(
    title = "Practice 2b. One-Way ANOVA: Response Time by EMT Level"
  ) |>
  tab_source_note(
    source_note = "Note. η² = eta-squared effect size. N = 200."
  ) |>
  cols_label(
    Source = "Source", SS = "SS", df = "df",
    MS = "MS", F = "F", p = "p", eta2 = "η²"
  ) |>
  cols_align(align = "center",
             columns = c(SS, df, MS, F, p, eta2)) |>
  cols_align(align = "left", columns = Source) |>
  tab_style(
    style     = cell_text(weight = "bold"),
    locations = cells_column_labels()
  ) |>
  sub_missing(missing_text = "") |>
  tab_style(
    style     = cell_fill(color = "#F0F5FA"),
    locations = cells_body(rows = 3)
  ) |>
  tab_options(
    table.border.top.style            = "hidden",
    heading.border.bottom.style       = "solid",
    column_labels.border.top.style    = "solid",
    column_labels.border.bottom.style = "solid",
    table.border.bottom.style         = "solid",
    table.font.size                   = px(13),
    heading.align                     = "left"
  )
```

### Practice 2c: Logistic Regression

```{r}
logit_r <- glm(occ_injury ~ stress_vas + night_shifts,
               data   = ems,
               family = binomial)
logit_s <- summary(logit_r)
ci_v    <- confint(logit_r)
coef_d  <- as.data.frame(logit_s$coefficients)
coef_d  <- tibble::rownames_to_column(coef_d, "Variable")

data.frame(
  Variable = c("Intercept", "Perceived Stress",
               "Night Shifts"),
  B        = round(coef_d$Estimate, 3),
  SE       = round(coef_d$`Std. Error`, 3),
  OR       = round(exp(coef_d$Estimate), 3),
  CI       = paste0("[",
                round(exp(ci_v[, 1]), 3), ", ",
                round(exp(ci_v[, 2]), 3), "]"),
  p        = ifelse(coef_d$`Pr(>|z|)` < .001,
                    "< .001",
                    round(coef_d$`Pr(>|z|)`, 3))
) |>
  gt() |>
  tab_header(
    title = "Practice 2c. Logistic Regression: Predictors of Occupational Injury"
  ) |>
  tab_source_note(
    source_note = "Note. OR = odds ratio; CI = 95% CI of OR. N = 200."
  ) |>
  cols_align(align = "center",
             columns = c(B, SE, OR, CI, p)) |>
  cols_align(align = "left", columns = Variable) |>
  tab_style(
    style     = cell_text(weight = "bold"),
    locations = cells_column_labels()
  ) |>
  tab_options(
    table.border.top.style            = "hidden",
    heading.border.bottom.style       = "solid",
    column_labels.border.top.style    = "solid",
    column_labels.border.bottom.style = "solid",
    table.border.bottom.style         = "solid",
    table.font.size                   = px(13),
    heading.align                     = "left"
  )
```

### Practice 3: APA Reporting Templates

Based on the results above, practice writing complete APA-format reporting sentences using these templates:

**Independent t-test:**

> Male EMS personnel (*M* = X.XX, *SD* = X.XX) [did / did not] differ significantly from female personnel (*M* = X.XX, *SD* = X.XX) in perceived stress, *t*(df) = X.XX, *p* = X.XX, 95% CI [X.XX, X.XX], *d* = X.XX (effect size judgment).

**One-Way ANOVA:**

> A one-way ANOVA showed that response time [did / did not] differ significantly across EMT certification levels, *F*(df₁, df₂) = X.XX, *p* [< .001 / = X.XX], η² = X.XX (effect size judgment).

**Logistic regression:**

> Perceived stress [significantly / did not significantly] predict occupational injury, *OR* = X.XX, 95% CI [X.XX, X.XX], *p* = X.XX, indicating that [interpretation].

---

## V. Final Exam Information

::: {.callout-important}
**Final Coding Exam Format (Topic 16)**

- Duration: 120 minutes
- Dataset: `ems_final.csv` (distributed on exam day)
- Tools: R + RStudio; course website (read-only reference)
- Not permitted: Internet search, AI tools, peer discussion

**Required format for each question:**

1. Justification for method selection (X-Y type identification, 1–2 sentences)
2. Assumption check results
3. Complete executable R code
4. Key statistical output
5. APA-format reporting sentence
6. Practical interpretation (1 sentence linking results to EMS context)
:::

```{r}
exam_tbl <- data.frame(
  Question_Type = c(
    "Descriptive Statistics",
    "Inferential (Comparison)",
    "Inferential (Association)",
    "Regression",
    "Advanced Method"
  ),
  Example_Tasks = c(
    "Produce a descriptive statistics table; assess distribution characteristics",
    "Compare continuous Y across 2 or 3 groups; select appropriate test",
    "Examine the relationship between two continuous variables; choose Pearson or Spearman",
    "Build linear or logistic regression; interpret coefficients",
    "Mediation analysis or repeated measures (as specified)"
  ),
  Weight   = c("15%", "25%", "20%", "25%", "15%"),
  Key_Methods = c(
    "describe(), gt() tables",
    "t.test(), aov(), TukeyHSD()",
    "cor.test(), Pearson / Spearman",
    "lm(), glm(), exp() for OR",
    "mediate(), ezANOVA() / geeglm()"
  )
)

exam_tbl |>
  gt() |>
  tab_header(
    title = "Table 4. Final Exam Question Types and Weighting"
  ) |>
  tab_source_note(
    source_note = paste0(
      "Note. Each question requires: method justification, assumption check, ",
      "R code, output, APA reporting sentence, and practical interpretation."
    )
  ) |>
  cols_label(
    Question_Type = "Question Type",
    Example_Tasks = "Example Tasks",
    Weight        = "Weight",
    Key_Methods   = "Key R Functions"
  ) |>
  cols_align(align = "center", columns = Weight) |>
  cols_align(align = "left",
             columns = c(Question_Type, Example_Tasks,
                         Key_Methods)) |>
  tab_style(
    style     = cell_text(weight = "bold"),
    locations = cells_column_labels()
  ) |>
  tab_options(
    table.border.top.style            = "hidden",
    heading.border.bottom.style       = "solid",
    column_labels.border.top.style    = "solid",
    column_labels.border.bottom.style = "solid",
    table.border.bottom.style         = "solid",
    table.font.size                   = px(13),
    heading.align                     = "left"
  )
```

---

## VI. Pre-Exam Self-Assessment Checklist

::: {.callout-tip}
**Basic Operations**

- [ ] Import CSV data and convert to factor
- [ ] Compute descriptive statistics (overall and by group)
- [ ] Produce APA-format `gt` tables
- [ ] Produce APA-format ggplot2 figures

**Statistical Methods**

- [ ] "Two groups + continuous Y" → immediately think through the t-test pipeline
- [ ] "Three groups + continuous Y" → immediately think ANOVA + Tukey
- [ ] "Binary Y" → immediately think logistic regression + OR interpretation
- [ ] "Continuous X + continuous Y" → correlation or linear regression
- [ ] Know when to use nonparametric alternatives

**Reporting**

- [ ] APA format: *t*(df) = value, *p* = value, *d* = value
- [ ] APA format: *F*(df₁, df₂) = value, *p* = value, η² = value
- [ ] APA format: *OR* = value, 95% CI [lower, upper], *p* = value
- [ ] Know that p < .05 does not equal a large effect
- [ ] Know the correct interpretation of confidence intervals
:::

---

## Practice Assignment (Ungraded — Self-Timed Exam Simulation)

Using `ems_main.csv`, complete the following within 60 minutes:

**P1** Produce a complete participant characteristics table (Table 1) including: all categorical variables (frequency, %) and continuous variables (Mean ± SD or Median (IQR) based on distribution). Present in APA `gt` format.

**P2** Compare `mbi_ee` across three districts (excluding East). Run one-way ANOVA and Tukey post-hoc comparisons. Present full results (table + APA reporting sentence).

**P3** Build a multiple linear regression predicting `mbi_ee` from `stress_vas`, `night_shifts`, `sleep_hrs`, and `years`. Report overall model fit and standardized coefficients. Identify the strongest predictor.

**P4** Run logistic regression predicting `occ_injury` from `stress_vas`, `years`, and `level`. Report OR and 95% CI. Plot a ROC curve and compute AUC.

Time yourself and review your APA reporting sentences against the examples in this topic.