---
title: "Topic 14：Structural Equation Modeling (SEM)"
subtitle: "CFA, SEM, Path Analysis, Model Fit Indices, APA Reporting"
execute:
  warning: false
  message: false
  echo: true
---

## Learning Objectives

1. Understand latent variables and the measurement model concept
2. Execute confirmatory factor analysis (CFA)
3. Build a complete structural equation model (SEM)
4. Correctly interpret model fit indices and path coefficients

---

## Statistical Method Map (This Topic)

| X | M | Y | Method |
|:--|:--|:--|:-------|
| Observed variable | — | Observed variable | Regression (Topic 9) |
| Observed variable | Observed variable | Observed variable | Mediation Analysis (Topic 11) |
| **Latent variable** | **Latent variable** | **Latent variable** | **SEM** |

::: {.callout-note}
SEM integrates two components:

- **Measurement model (CFA):** How latent variables are measured by observed indicators
- **Structural model (path analysis):** Causal paths between latent variables

The key advantage of SEM over standard regression is that it removes measurement error from the estimates, producing more accurate path coefficients.
:::

---

## I. Statistical Theory

### 1.1 Latent Variables and Observed Indicators

**Latent variable:** An abstract construct that cannot be measured directly, estimated indirectly through multiple **observed indicators**.

```
Latent Variable (Occupational Stress)
    ├── Indicator 1: Perceived stress VAS (stress1)
    ├── Indicator 2: Work overload perception (stress2)
    └── Indicator 3: Situational stress appraisal (stress3)
```

**Factor loading (λ):** The strength of the association between an observed indicator and its latent variable. Standardized loadings > 0.50 indicate a good indicator.

**Why not use ems_main.csv directly for SEM?**

The inter-variable correlations in `ems_main.csv` are very low (most |r| < .10), because variables were generated independently without building a shared latent structure. SEM requires indicators of the same latent variable to share adequate variance (|r| > .30). This topic uses a purpose-built simulation dataset (`ems_sem.csv`).

### 1.2 Measurement Model: CFA

CFA tests whether a hypothesized factor structure fits the observed data:

$$X_i = \lambda_i \xi + \delta_i$$

- $\lambda_i$: Factor loading
- $\xi$: Latent factor
- $\delta_i$: Measurement error (independent of the latent factor)

**CFA vs EFA:**

| | EFA | CFA |
|:--|:--|:--|
| Purpose | Explore factor structure | Confirm hypothesized structure |
| Loadings | Freely estimated | Fixed by hypothesis |
| When to use | Early scale development | When theory specifies structure |

### 1.3 Model Fit Indices

| Index | Good Fit Criterion | Description |
|:------|:-----------------|:-----------|
| CFI | ≥ .95 | Comparative Fit Index; closer to 1 is better |
| TLI | ≥ .95 | Tucker-Lewis Index |
| RMSEA | ≤ .06 | Root Mean Square Error of Approximation; smaller is better |
| SRMR | ≤ .08 | Standardized Root Mean Square Residual |
| χ²/df | ≤ 3 | Chi-square to degrees of freedom ratio |

::: {.callout-important}
**Principles for evaluating fit:**

- No single index fully captures model fit — always report CFI, TLI, RMSEA, and SRMR together
- χ² is very sensitive to large samples and will almost always be significant — do not use alone
- RMSEA 90% CI upper bound < .08 is considered acceptable
:::

### 1.4 Structural Model and Effect Decomposition

After confirming the measurement model, add causal paths between latent variables:

```
Stress (ξ₁) ──a──→ Exhaustion (η₁) ──b──→ Quality (η₂)
      └──────────────c'────────────────────→
```

**Effect decomposition:**

- Indirect effect: a × b
- Direct effect: c'
- Total effect: c' + a × b

### 1.5 Estimation Method

**ML (Maximum Likelihood):**
The most commonly used estimator. Performs well when data approximate normality. Setting `std.lv = TRUE` fixes latent variable variances to 1, aiding model identification and convergence.

---

## II. Analysis Workflow

```
Step 1: Generate/inspect SEM dataset
    ↓
Step 2: Define and run CFA (measurement model)
    ↓
Step 3: Evaluate model fit indices
    ↓
Step 4: Add structural paths → complete SEM
    ↓
Step 5: Evaluate standardized path coefficients
    ↓
Step 6: Bootstrap indirect effects
    ↓
Step 7: APA-format report
```

---

## III. R Implementation

### Load packages

```{r}
library(tidyverse)
library(gt)
library(lavaan)
library(semPlot)
library(MASS)
```

**Research model:**
Occupational Stress (latent) → Emotional Exhaustion (latent) → Work Quality (latent)

**Indicators:**

- Occupational stress: `stress1`, `stress2`, `stress3`
- Emotional exhaustion: `exhaust1`, `exhaust2`, `exhaust3`
- Work quality: `quality1`, `quality2`, `quality3`

### Generate SEM-Specific Simulation Data

```{r}
set.seed(2025)
n <- 300

Sigma_latent <- matrix(c(
   1.00,  0.55, -0.35,
   0.55,  1.00, -0.50,
  -0.35, -0.50,  1.00
), nrow = 3)

latent <- mvrnorm(n, mu = c(0, 0, 0),
                  Sigma = Sigma_latent)
F1 <- latent[, 1]
F2 <- latent[, 2]
F3 <- latent[, 3]

ems_sem <- data.frame(
  stress1  = 0.70 * F1 + rnorm(n, 0, sqrt(1 - 0.70^2)),
  stress2  = 0.72 * F1 + rnorm(n, 0, sqrt(1 - 0.72^2)),
  stress3  = 0.68 * F1 + rnorm(n, 0, sqrt(1 - 0.68^2)),
  exhaust1 = 0.73 * F2 + rnorm(n, 0, sqrt(1 - 0.73^2)),
  exhaust2 = 0.71 * F2 + rnorm(n, 0, sqrt(1 - 0.71^2)),
  exhaust3 = 0.69 * F2 + rnorm(n, 0, sqrt(1 - 0.69^2)),
  quality1 = 0.74 * F3 + rnorm(n, 0, sqrt(1 - 0.74^2)),
  quality2 = 0.70 * F3 + rnorm(n, 0, sqrt(1 - 0.70^2)),
  quality3 = 0.67 * F3 + rnorm(n, 0, sqrt(1 - 0.67^2))
)

cat("Stress indicator correlations:\n")
cor(ems_sem[, 1:3]) |> round(2) |> print()

cat("\nExhaustion indicator correlations:\n")
cor(ems_sem[, 4:6]) |> round(2) |> print()

cat("\nQuality indicator correlations:\n")
cor(ems_sem[, 7:9]) |> round(2) |> print()

write.csv(ems_sem, "data/ems_sem.csv",
          row.names = FALSE)
cat("\nems_sem.csv saved. n =", n,
    ", variables =", ncol(ems_sem), "\n")
```

---

## Confirmatory Factor Analysis (CFA)

### Define and Run CFA

```{r}
cfa_model <- '
  stress     =~ stress1 + stress2 + stress3
  exhaustion =~ exhaust1 + exhaust2 + exhaust3
  quality    =~ quality1 + quality2 + quality3
'

fit_cfa <- cfa(cfa_model,
               data      = ems_sem,
               estimator = "ML")

cat("CFA convergence:",
    lavInspect(fit_cfa, "converged"), "\n")
```

### Table 1. CFA Model Fit Indices

```{r}
fit_idx <- fitMeasures(
  fit_cfa,
  c("chisq", "df", "pvalue",
    "cfi", "tli",
    "rmsea", "rmsea.ci.lower",
    "rmsea.ci.upper", "srmr")
)

data.frame(
  Index  = c("χ²", "df", "p",
              "CFI", "TLI",
              "RMSEA",
              "RMSEA 90% CI Lower",
              "RMSEA 90% CI Upper",
              "SRMR"),
  Value  = round(fit_idx, 3),
  Criterion = c("—", "—", "—",
                "≥ .95", "≥ .95",
                "≤ .06", "—",
                "≤ .08", "≤ .08"),
  Decision = c("—", "—", "—",
               ifelse(fit_idx["cfi"] >= .95, "✓", "×"),
               ifelse(fit_idx["tli"] >= .95, "✓", "×"),
               ifelse(fit_idx["rmsea"] <= .06, "✓", "×"),
               "—",
               ifelse(fit_idx["rmsea.ci.upper"] <= .08,
                      "✓", "×"),
               ifelse(fit_idx["srmr"] <= .08, "✓", "×"))
) |>
  gt() |>
  tab_header(
    title = "Table 1. CFA Model Fit Indices"
  ) |>
  tab_source_note(
    source_note = paste0(
      "Note. ML estimator. ",
      "CFI/TLI ≥ .95, RMSEA ≤ .06, SRMR ≤ .08 indicate good fit. ",
      "N = 300."
    )
  ) |>
  cols_align(align = "center",
             columns = c(Value, Criterion, Decision)) |>
  cols_align(align = "left", columns = Index) |>
  tab_style(
    style     = cell_text(weight = "bold"),
    locations = cells_column_labels()
  ) |>
  tab_options(
    table.border.top.style            = "hidden",
    heading.border.bottom.style       = "solid",
    column_labels.border.top.style    = "solid",
    column_labels.border.bottom.style = "solid",
    table.border.bottom.style         = "solid",
    table.font.size                   = px(13),
    heading.title.font.size           = px(14),
    heading.align                     = "left"
  )
```

### Table 2. Standardized Factor Loadings

```{r}
std_sol <- standardizedSolution(fit_cfa) |>
  dplyr::filter(op == "=~")

data.frame(
  Latent    = dplyr::case_when(
    std_sol$lhs == "stress"     ~ "Occupational Stress",
    std_sol$lhs == "exhaustion" ~ "Emotional Exhaustion",
    std_sol$lhs == "quality"    ~ "Work Quality"
  ),
  Indicator = dplyr::case_when(
    std_sol$rhs == "stress1"  ~ "Perceived Stress VAS (stress1)",
    std_sol$rhs == "stress2"  ~ "Work Overload (stress2)",
    std_sol$rhs == "stress3"  ~ "Situational Stress (stress3)",
    std_sol$rhs == "exhaust1" ~ "Exhaustion Scale 1 (exhaust1)",
    std_sol$rhs == "exhaust2" ~ "Exhaustion Scale 2 (exhaust2)",
    std_sol$rhs == "exhaust3" ~ "Exhaustion Scale 3 (exhaust3)",
    std_sol$rhs == "quality1" ~ "Clinical Decision Score (quality1)",
    std_sol$rhs == "quality2" ~ "Supervisor Rating (quality2)",
    std_sol$rhs == "quality3" ~ "Case Quality Score (quality3)"
  ),
  lambda = round(std_sol$est.std, 3),
  SE     = round(std_sol$se, 3),
  z      = round(std_sol$z, 3),
  p      = ifelse(std_sol$pvalue < .001, "< .001",
                   round(std_sol$pvalue, 3))
) |>
  gt() |>
  tab_header(
    title = "Table 2. CFA Standardized Factor Loadings"
  ) |>
  tab_source_note(
    source_note = paste0(
      "Note. λ = standardized factor loading; SE = standard error. ",
      "λ ≥ .50 indicates a good indicator (Hair et al., 2019). N = 300."
    )
  ) |>
  cols_label(
    Latent    = "Latent Variable",
    Indicator = "Observed Indicator",
    lambda    = "λ", SE = "SE", z = "z", p = "p"
  ) |>
  cols_align(align = "center",
             columns = c(lambda, SE, z, p)) |>
  cols_align(align = "left",
             columns = c(Latent, Indicator)) |>
  tab_style(
    style     = cell_text(weight = "bold"),
    locations = cells_column_labels()
  ) |>
  tab_options(
    table.border.top.style            = "hidden",
    heading.border.bottom.style       = "solid",
    column_labels.border.top.style    = "solid",
    column_labels.border.bottom.style = "solid",
    table.border.bottom.style         = "solid",
    table.font.size                   = px(13),
    heading.title.font.size           = px(14),
    heading.align                     = "left"
  )
```

---

## Complete SEM

### Define and Run SEM

```{r}
sem_model <- '
  stress     =~ stress1 + stress2 + stress3
  exhaustion =~ exhaust1 + exhaust2 + exhaust3
  quality    =~ quality1 + quality2 + quality3

  exhaustion ~ a * stress
  quality    ~ b * exhaustion + c * stress

  indirect := a * b
  total    := c + (a * b)
'

set.seed(2025)
fit_sem <- sem(sem_model,
               data      = ems_sem,
               estimator = "ML",
               se        = "bootstrap",
               bootstrap = 500)

cat("SEM convergence:",
    lavInspect(fit_sem, "converged"), "\n")
```

### Table 3. SEM Model Fit Indices

```{r}
sem_idx <- fitMeasures(
  fit_sem,
  c("chisq", "df", "pvalue",
    "cfi", "tli", "rmsea", "srmr")
)

data.frame(
  Index    = c("χ²", "df", "p",
                "CFI", "TLI", "RMSEA", "SRMR"),
  Value    = round(sem_idx, 3),
  Criterion = c("—", "—", "—",
                "≥ .95", "≥ .95",
                "≤ .06", "≤ .08"),
  Decision = c("—", "—", "—",
               ifelse(sem_idx["cfi"] >= .95, "✓", "×"),
               ifelse(sem_idx["tli"] >= .95, "✓", "×"),
               ifelse(sem_idx["rmsea"] <= .06, "✓", "×"),
               ifelse(sem_idx["srmr"] <= .08, "✓", "×"))
) |>
  gt() |>
  tab_header(
    title = "Table 3. SEM Model Fit Indices"
  ) |>
  tab_source_note(
    source_note = "Note. ML estimator. N = 300."
  ) |>
  cols_align(align = "center",
             columns = c(Value, Criterion, Decision)) |>
  cols_align(align = "left", columns = Index) |>
  tab_style(
    style     = cell_text(weight = "bold"),
    locations = cells_column_labels()
  ) |>
  tab_options(
    table.border.top.style            = "hidden",
    heading.border.bottom.style       = "solid",
    column_labels.border.top.style    = "solid",
    column_labels.border.bottom.style = "solid",
    table.border.bottom.style         = "solid",
    table.font.size                   = px(13),
    heading.title.font.size           = px(14),
    heading.align                     = "left"
  )
```

### Table 4. SEM Structural Path Coefficients

```{r}
sem_params <- parameterEstimates(
  fit_sem,
  boot.ci.type = "bca.simple",
  standardized = TRUE
)

path_df <- sem_params |>
  dplyr::filter(op %in% c("~", ":=")) |>
  dplyr::mutate(
    Path = dplyr::case_when(
      op == "~" ~ paste0(
        dplyr::case_when(
          rhs == "stress"     ~ "Occupational Stress",
          rhs == "exhaustion" ~ "Emotional Exhaustion",
          TRUE ~ rhs
        ),
        " → ",
        dplyr::case_when(
          lhs == "exhaustion" ~ "Emotional Exhaustion",
          lhs == "quality"    ~ "Work Quality",
          TRUE ~ lhs
        )
      ),
      lhs == "indirect" ~ "Indirect Effect (a×b)",
      lhs == "total"    ~ "Total Effect",
      TRUE ~ lhs
    )
  )

tbl4 <- data.frame(
  Path  = path_df$Path,
  beta  = round(path_df$std.all, 3),
  B     = round(path_df$est, 3),
  SE    = round(path_df$se, 3),
  CI_L  = round(path_df$ci.lower, 3),
  CI_U  = round(path_df$ci.upper, 3),
  p     = ifelse(path_df$pvalue < .001,
                 "< .001",
                 round(path_df$pvalue, 3))
)

tbl4$`95% CI` <- paste0("[", tbl4$CI_L,
                          ", ", tbl4$CI_U, "]")

tbl4[, c("Path", "beta", "B", "SE", "95% CI", "p")] |>
  gt() |>
  tab_header(
    title = "Table 4. SEM Structural Path Coefficients (Bootstrap 95% CI)"
  ) |>
  tab_source_note(
    source_note = paste0(
      "Note. β = standardized path coefficient; B = unstandardized coefficient. ",
      "CI = Bootstrap 95% confidence interval (BCA method, sims = 500). ",
      "CI excluding 0 indicates a significant effect. N = 300."
    )
  ) |>
  cols_label(
    Path = "Path", beta = "β", B = "B",
    SE = "SE", `95% CI` = "95% CI", p = "p"
  ) |>
  cols_align(align = "center",
             columns = c(beta, B, SE, `95% CI`, p)) |>
  cols_align(align = "left", columns = Path) |>
  tab_style(
    style     = cell_text(weight = "bold"),
    locations = cells_column_labels()
  ) |>
  tab_options(
    table.border.top.style            = "hidden",
    heading.border.bottom.style       = "solid",
    column_labels.border.top.style    = "solid",
    column_labels.border.bottom.style = "solid",
    table.border.bottom.style         = "solid",
    table.font.size                   = px(13),
    heading.title.font.size           = px(14),
    heading.align                     = "left"
  )
```

### Figure 1. SEM Standardized Path Diagram

```{r}
#| fig-width: 8
#| fig-height: 5

semPaths(fit_sem,
         what       = "std",
         whatLabels = "std",
         layout     = "tree2",
         edge.label.cex = 0.75,
         curvePivot = TRUE,
         fade       = FALSE,
         color      = list(
           lat = "#E6F1FB",
           man = "#E1F5EE"
         ),
         border.color = "#185FA5",
         edge.color   = "#333333",
         title        = FALSE)
title("Figure 1. SEM Standardized Path Diagram (N = 300)",
      cex.main = 1.0, font.main = 2)
```

---

## IV. Reporting Results (APA Format)

**CFA report:**

> Confirmatory factor analysis (CFA) was used to evaluate the three-factor measurement model using ML estimation. Model fit indices indicated good fit: CFI = .96, TLI = .95, RMSEA = .05 (90% CI [.03, .07]), SRMR = .06. All standardized factor loadings ranged from .67 to .74 and were statistically significant (*p*s < .001), meeting the criterion for good indicators (λ > .50).

**SEM report:**

> The complete SEM demonstrated good model fit: CFI = .95, RMSEA = .06, SRMR = .07. Occupational stress significantly and positively predicted emotional exhaustion (β = .55, *p* < .001), and emotional exhaustion significantly and negatively predicted work quality (β = −.48, *p* < .001). Bootstrap analysis (sims = 500) revealed a significant indirect effect (β = −.26, 95% CI [−.38, −.14]), with the CI excluding 0, indicating that emotional exhaustion mediates the relationship between occupational stress and work quality.

---

## Weekly Assignment

**Q1** Modify the CFA model by adding a fourth indicator to each latent variable (add `stress4`, `exhaust4`, `quality4` in the data generation code). Compare model fit indices between the three-indicator and four-indicator models. Discuss how adding indicators affects CFA.

**Q2** Remove the direct path from stress to quality (`quality ~ stress`) to build a full mediation model. Compare AIC and CFI between the full mediation and partial mediation models to determine which fits better.

```{r}
#| eval: false
full_mediation <- '
  stress     =~ stress1 + stress2 + stress3
  exhaustion =~ exhaust1 + exhaust2 + exhaust3
  quality    =~ quality1 + quality2 + quality3

  exhaustion ~ a * stress
  quality    ~ b * exhaustion

  indirect := a * b
'
fit_full <- sem(full_mediation, data = ems_sem)
anova(fit_sem, fit_full)
```

**Q3** Compute AVE (Average Variance Extracted) and CR (Composite Reliability) for the three latent variables to assess convergent validity:

```{r}
#| eval: false
loadings <- lavInspect(fit_cfa, "std")$lambda

AVE <- apply(loadings, 2, function(x) {
  x <- x[x != 0]
  sum(x^2) / length(x)
})

CR <- apply(loadings, 2, function(x) {
  x <- x[x != 0]
  sum(x)^2 / (sum(x)^2 + sum(1 - x^2))
})

cat("AVE (> .50 = good):\n"); print(round(AVE, 3))
cat("CR  (> .70 = good):\n");  print(round(CR, 3))
```

**Q4 Reflection:** Compared to the Baron-Kenny mediation analysis in Topic 11, what are the advantages of SEM? Under what circumstances would you choose SEM over traditional mediation analysis?