---
title: "Topic 4：Probability Distributions & Sampling Theory"
subtitle: "Normal Distribution, Central Limit Theorem, Standard Error, Confidence Intervals"
execute:
  warning: false
  message: false
  echo: true
---

## Learning Objectives

1. Understand the properties of the normal distribution and the 68-95-99.7 rule
2. Grasp the meaning and practical implications of the Central Limit Theorem (CLT)
3. Distinguish between standard deviation (SD) and standard error (SE)
4. Correctly construct and interpret confidence intervals

---

## Statistical Method Map (This Topic)

::: {.callout-note}
This topic provides the **theoretical foundation** for all inferential statistics. Every method from Topic 5 onward — t-tests, ANOVA, regression — relies on the concepts introduced here.
:::

| Concept | Description | First Applied |
|:--------|:-----------|:-------------|
| Normal distribution | Assumption underlying t-tests | Topic 6 |
| Standard error | Denominator of the t-statistic | Topic 6 |
| Confidence interval | Reporting standard for all inferential results | Topic 6 onward |
| Central Limit Theorem | Why large samples don't require normality | Topic 6 |

---

## I. Statistical Theory

### 1.1 The Role of Probability Distributions

A probability distribution describes the probability associated with every possible value of a random variable. Without a known distribution, we cannot compute p-values or construct confidence intervals.

The core logic of statistical inference is:

```
Observe a sample statistic (e.g., sample mean)
  ↓
Assuming H₀ is true, this statistic follows a known distribution
  ↓
Calculate the probability of observing this value or something more extreme
  ↓
This probability is the p-value
```

Choosing the wrong distribution means using the wrong test.

### 1.2 The Normal Distribution

$$X \sim N(\mu, \sigma^2)$$

The normal distribution is the most important in statistics — not simply because many variables are approximately normal, but because the **Central Limit Theorem** makes it the foundation of inferential statistics.

**Properties:**

- Symmetric, centered at μ
- Mean = Median = Mode
- Fully determined by μ (location) and σ (spread)
- Total area under the curve = 1

**The 68-95-99.7 Rule:**

| Range | Coverage |
|:------|:--------|
| μ ± 1σ | ≈ 68.3% |
| μ ± 2σ | ≈ 95.4% |
| μ ± 3σ | ≈ 99.7% |

::: {.callout-tip}
**EMS Application:**
If EMS response time ~ N(9.3, 2.5²), then approximately 95% of responses fall within 9.3 ± 2×2.5 = [4.3, 14.3] minutes. Responses exceeding 14.3 minutes represent only ~2.5% of calls and could serve as a performance alert threshold.
:::

### 1.3 The Standard Normal Distribution (Z)

$$Z = \frac{X - \mu}{\sigma} \sim N(0, 1)$$

Standardizing any normal distribution (subtract mean, divide by SD) yields a standard normal with mean = 0, SD = 1.

**Interpretation of Z-scores:** How many standard deviations is this observation from the mean?

- Z = 1.96 → right-tail area = 2.5% (two-tailed α = .05 critical value)
- Z = 1.64 → right-tail area = 5% (one-tailed α = .05 critical value)

### 1.4 The t-Distribution

When the population SD σ is unknown — which is almost always the case — replacing it with the sample SD s introduces additional uncertainty. The resulting statistic follows a **t-distribution**:

$$T = \frac{\bar{X} - \mu}{s / \sqrt{n}} \sim t_{(n-1)}$$

**Properties of the t-distribution:**

- Heavier tails than the standard normal
- Degrees of freedom df = n - 1
- As df increases → approaches the standard normal
- At df ≥ 30, t and Z are nearly indistinguishable

::: {.callout-important}
**Why t instead of Z?**

Using s instead of σ introduces extra uncertainty. The heavier tails of the t-distribution reflect this uncertainty — it demands more extreme evidence before rejecting H₀, which is especially important in small samples.
:::

### 1.5 The Central Limit Theorem (CLT)

::: {.callout-important}
**One of the most important theorems in statistics.**
:::

**Statement:** When drawing random samples of size n from any population (regardless of distribution shape), as n increases, the **sampling distribution of the sample mean** approaches a normal distribution:

$$\bar{X} \sim N\left(\mu, \frac{\sigma^2}{n}\right)$$

**Three critical points:**

1. **"Any population":** The original data do not need to be normally distributed
2. **"Sample mean":** It is the distribution of means that becomes normal — not the raw data
3. **"As n increases":** In practice, n ≥ 30 is generally sufficient

**Practical implication:**

```
Raw data may be skewed (e.g., response times, income)
       ↓
As long as n ≥ 30
       ↓
The sampling distribution of the mean approximates normality
       ↓
t-tests remain valid
```

### 1.6 SD vs SE — The Most Commonly Confused Pair

| | Standard Deviation (SD) | Standard Error (SE) |
|:--|:----------------------|:-------------------|
| **Describes** | Spread of individual observations | Spread of sample means across repeated samples |
| **Formula** | $s = \sqrt{\frac{\sum(x_i-\bar{x})^2}{n-1}}$ | $SE = \frac{s}{\sqrt{n}}$ |
| **Purpose** | Describe data variability | Inference, confidence intervals |
| **As n increases** | Does not shrink (reflects true variability) | Shrinks (estimate becomes more precise) |
| **Report when** | Descriptive statistics (Mean ± SD) | Inferential statistics (CI) |

::: {.callout-tip}
**Memory aid:**
SD asks "how different are individuals from each other?"
SE asks "if I repeated this sampling, how much would the mean jump around?"
:::

### 1.7 Confidence Intervals (CI)

$$CI = \bar{x} \pm t_{(\alpha/2,\, df)} \times SE$$

**Correct interpretation of a 95% CI:**

> "If we repeated this sampling procedure 100 times and computed a CI each time, approximately 95 of those intervals would contain the true population mean μ."

**Common misinterpretations:**

❌ "There is a 95% probability that μ falls in this interval."
→ Wrong. μ is fixed — it either is or isn't in any given interval. Probability doesn't apply to a fixed parameter.

❌ "95% of the data falls in this interval."
→ Wrong. That describes ±2 SD, not the CI.

::: {.callout-important}
**What CI width tells you:**

- Narrower CI → more precise estimate (larger n, smaller variability)
- Wider CI → more uncertainty (smaller n, larger variability)
- CI excluding 0 (for a difference) → statistically significant
:::

---

## II. R Implementation

### Load packages and data

```{r}
library(tidyverse)
library(gt)

ems <- read.csv("data/ems_main.csv",
                stringsAsFactors = FALSE)

ems$level <- factor(ems$level,
                    levels = c("EMT-1", "EMT-2", "EMT-P"))
```

### Figure 1. Normal Distribution and the 68-95-99.7 Rule

```{r}
#| fig-width: 7
#| fig-height: 4.5

mu <- 9.3; sigma <- 2.5
x  <- seq(mu - 4*sigma, mu + 4*sigma, length.out = 500)
df_norm <- data.frame(x = x, y = dnorm(x, mu, sigma))

ggplot(df_norm, aes(x, y)) +
  geom_area(data = subset(df_norm,
              x >= mu - 3*sigma & x <= mu + 3*sigma),
            fill = "#CBE0F5", alpha = 0.6) +
  geom_area(data = subset(df_norm,
              x >= mu - 2*sigma & x <= mu + 2*sigma),
            fill = "#85B7EB", alpha = 0.6) +
  geom_area(data = subset(df_norm,
              x >= mu - sigma  & x <= mu + sigma),
            fill = "#378ADD", alpha = 0.6) +
  geom_line(linewidth = 1, color = "#0C447C") +
  geom_vline(xintercept = mu,
             linetype = "dashed",
             color = "#D85A30", linewidth = 0.8) +
  annotate("text", x = mu, y = max(df_norm$y) * 1.05,
           label = paste0("μ = ", mu),
           color = "#D85A30", size = 3.5) +
  annotate("text", x = mu, y = max(df_norm$y) * 0.55,
           label = "68.3%", color = "white",
           size = 3.5, fontface = "bold") +
  annotate("text", x = mu + 1.6*sigma,
           y = max(df_norm$y) * 0.20,
           label = "95.4%", color = "#185FA5", size = 3.2) +
  annotate("text", x = mu + 2.6*sigma,
           y = max(df_norm$y) * 0.05,
           label = "99.7%", color = "#185FA5", size = 3.0) +
  labs(
    title   = "Figure 1. Normal Distribution and the 68-95-99.7 Rule",
    x       = "Response Time (minutes)",
    y       = "Probability Density",
    caption = "Note. Dark blue = μ±1σ (68.3%); medium blue = μ±2σ (95.4%); light blue = μ±3σ (99.7%)."
  ) +
  theme_minimal(base_size = 12) +
  theme(
    plot.title   = element_text(face = "bold", size = 12),
    plot.caption = element_text(hjust = 0, size = 9,
                                color = "gray40")
  )
```

### Figure 2. t-Distribution vs Standard Normal

```{r}
#| fig-width: 7
#| fig-height: 4.5

x <- seq(-4, 4, length.out = 500)
df_t <- bind_rows(
  data.frame(x = x, y = dnorm(x),
             dist = "Standard Normal (Z)"),
  data.frame(x = x, y = dt(x, df = 5),
             dist = "t-distribution (df = 5)"),
  data.frame(x = x, y = dt(x, df = 30),
             dist = "t-distribution (df = 30)")
)

ggplot(df_t, aes(x, y, color = dist, linetype = dist)) +
  geom_line(linewidth = 0.9) +
  scale_color_manual(
    values = c("Standard Normal (Z)"      = "#0C447C",
               "t-distribution (df = 5)"  = "#D85A30",
               "t-distribution (df = 30)" = "#1D9E75"),
    name = "Distribution"
  ) +
  scale_linetype_manual(
    values = c("Standard Normal (Z)"      = "solid",
               "t-distribution (df = 5)"  = "dashed",
               "t-distribution (df = 30)" = "dotdash"),
    name = "Distribution"
  ) +
  labs(
    title   = "Figure 2. Comparison of t-Distribution and Standard Normal",
    x       = "Standardized Value",
    y       = "Probability Density",
    caption = "Note. Smaller df produces heavier tails. At df ≥ 30, t approximates the standard normal."
  ) +
  theme_minimal(base_size = 12) +
  theme(
    plot.title      = element_text(face = "bold", size = 12),
    plot.caption    = element_text(hjust = 0, size = 9,
                                   color = "gray40"),
    legend.position = "bottom"
  )
```

### Figure 3. Central Limit Theorem Demonstration

```{r}
#| fig-width: 8
#| fig-height: 5

set.seed(2025)

simulate_clt <- function(n_sample, n_sim = 1000) {
  means <- replicate(n_sim,
                     mean(rexp(n_sample, rate = 0.1)))
  data.frame(mean = means,
             n    = paste0("n = ", n_sample))
}

clt_df <- bind_rows(
  simulate_clt(5),
  simulate_clt(10),
  simulate_clt(30),
  simulate_clt(100)
)
clt_df$n <- factor(clt_df$n,
  levels = c("n = 5","n = 10","n = 30","n = 100"))

ggplot(clt_df, aes(x = mean)) +
  geom_histogram(aes(y = after_stat(density)),
                 bins = 40, fill = "#85B7EB",
                 color = "white", alpha = 0.8) +
  geom_density(color = "#D85A30", linewidth = 0.8) +
  facet_wrap(~ n, scales = "free") +
  labs(
    title    = "Figure 3. Central Limit Theorem Demonstration",
    subtitle = "Population is exponential (right-skewed); sample means approach normality as n increases",
    x        = "Sample Mean",
    y        = "Density",
    caption  = "Note. Each panel shows 1,000 simulated sample means. Red curve = kernel density estimate."
  ) +
  theme_minimal(base_size = 12) +
  theme(
    plot.title    = element_text(face = "bold", size = 12),
    plot.subtitle = element_text(color = "gray40", size = 10),
    plot.caption  = element_text(hjust = 0, size = 9,
                                 color = "gray40")
  )
```

### Computing Confidence Intervals

```{r}
n    <- nrow(ems)
xbar <- mean(ems$response_time)
s    <- sd(ems$response_time)
se   <- s / sqrt(n)
t_cv <- qt(0.975, df = n - 1)

ci_lower <- xbar - t_cv * se
ci_upper <- xbar + t_cv * se

cat("Sample mean:          ", round(xbar, 2), "minutes\n")
cat("Standard deviation:   ", round(s, 2), "\n")
cat("Standard error:       ", round(se, 3), "\n")
cat("t critical value:     ", round(t_cv, 3), "\n")
cat("95% CI:               [", round(ci_lower, 2),
    ",", round(ci_upper, 2), "]\n")
```

### Table 1. Mean Response Time and 95% CI by EMT Level

```{r}
ems |>
  dplyr::group_by(level) |>
  dplyr::summarise(
    n    = n(),
    Mean = round(mean(response_time), 2),
    SD   = round(sd(response_time), 2),
    SE   = round(sd(response_time) / sqrt(n()), 3),
    CI_L = round(t.test(response_time)$conf.int[1], 2),
    CI_U = round(t.test(response_time)$conf.int[2], 2),
    .groups = "drop"
  ) |>
  dplyr::mutate(`95% CI` = paste0("[", CI_L, ", ", CI_U, "]")) |>
  dplyr::select(level, n, Mean, SD, SE, `95% CI`) |>
  dplyr::rename(Level = level) |>
  gt() |>
  tab_header(
    title = "Table 1. Response Time: Mean and 95% Confidence Interval by EMT Level"
  ) |>
  tab_source_note(
    source_note = "Note. Units: minutes. SE = standard error; CI = confidence interval."
  ) |>
  cols_align(align = "center", columns = -Level) |>
  cols_align(align = "left",   columns = Level) |>
  tab_style(
    style     = cell_text(weight = "bold"),
    locations = cells_column_labels()
  ) |>
  tab_options(
    table.border.top.style            = "hidden",
    heading.border.bottom.style       = "solid",
    column_labels.border.top.style    = "solid",
    column_labels.border.bottom.style = "solid",
    table.border.bottom.style         = "solid",
    table.font.size                   = px(13),
    heading.title.font.size           = px(14),
    heading.align                     = "left"
  )
```

### Figure 4. Mean and 95% CI by EMT Level

```{r}
#| fig-width: 6
#| fig-height: 4.5

ems |>
  dplyr::group_by(level) |>
  dplyr::summarise(
    mean = mean(response_time),
    ci_l = t.test(response_time)$conf.int[1],
    ci_u = t.test(response_time)$conf.int[2],
    .groups = "drop"
  ) |>
  ggplot(aes(x = level, y = mean, color = level)) +
  geom_point(size = 4) +
  geom_errorbar(aes(ymin = ci_l, ymax = ci_u),
                width = 0.15, linewidth = 0.8) +
  scale_color_manual(
    values = c("EMT-1" = "#D85A30",
               "EMT-2" = "#1D9E75",
               "EMT-P" = "#185FA5")
  ) +
  labs(
    title   = "Figure 4. Mean Response Time and 95% CI by EMT Level",
    x       = "EMT Certification Level",
    y       = "Response Time (minutes)",
    caption = "Note. Points = sample means; error bars = 95% confidence intervals. N = 200."
  ) +
  theme_minimal(base_size = 12) +
  theme(
    plot.title      = element_text(face = "bold", size = 12),
    legend.position = "none",
    plot.caption    = element_text(hjust = 0, size = 9,
                                   color = "gray40")
  )
```

---

## III. Reporting Results (APA Format)

**Descriptive statistics with CI:**

> The mean on-scene response time was 9.31 minutes (*SD* = 2.48, *SE* = 0.18, 95% CI [8.97, 9.65]).

**Interpreting a confidence interval:**

> If this sampling procedure were repeated, approximately 95% of the resulting confidence intervals would contain the true population mean response time. The 95% CI for the current sample is [8.97, 9.65] minutes.

**Comparing CIs across groups:**

> As shown in Figure 4, the 95% CI for EMT-P ([6.8, 8.2] min) does not overlap with that of EMT-1 ([9.7, 11.0] min), providing preliminary evidence of a significant group difference (formal testing in Topic 7 ANOVA).

---

## Weekly Assignment

**Q1** Manually calculate the 95% CI for `stress_vas` using the formula, then verify using `t.test()`. Confirm the two results match and write an APA-format reporting sentence.

**Q2** Create a mean + 95% CI point plot (like Figure 4) for `monthly_cases` by district (`district`). Accompany with an APA-format `gt` table.

**Q3** Run the CLT demonstration code, but replace the population with `rbinom(n, size = 10, prob = 0.3)`. Observe how the distribution of sample means changes at n = 5, 30, and 100. Describe what you observe.

**Q4 Reflection:** Study A reports a 95% CI of [2.1, 8.4]; Study B reports [4.8, 5.7]. Both have p < .05. Which study provides a more precise estimate? Why? What does this imply for how you read the literature?