---
title: "Topic 1：R Setup & Data Import"
subtitle: "Installation, RStudio Interface, Basic Operations, Importing ems_main.csv"
execute:
  warning: false
  message: false
  echo: true
---

## Learning Objectives

By the end of this topic, you should be able to:

1.  Install R and RStudio successfully
2.  Navigate the four RStudio panels
3.  Execute basic R syntax
4.  Import `ems_main.csv` and inspect the data structure

------------------------------------------------------------------------

## 1. Installing R and RStudio

::: callout-important
**Installation order matters: install R first, then RStudio.** RStudio is the interface; R is the engine.
:::

### Install R

1.  Go to <https://cran.r-project.org/>
2.  Select your operating system (Windows / macOS)
3.  Download the latest version (currently R 4.4.x) and install using all default settings

### Install RStudio

1.  Go to <https://posit.co/download/rstudio-desktop/>
2.  Download the free RStudio Desktop version
3.  Install using all default settings

Once installed, **only open RStudio** — you do not need to open R separately.

------------------------------------------------------------------------

## 2. The RStudio Interface

When RStudio opens, you will see four panels:

| Panel        | Name              | Purpose                          |
|:-------------|:------------------|:---------------------------------|
| Top-left     | **Script Editor** | Write and save your code         |
| Bottom-left  | **Console**       | Execute commands and view output |
| Top-right    | **Environment**   | View loaded objects and datasets |
| Bottom-right | **Files / Plots** | File management and plot output  |

::: callout-tip
**Best practice:** Always write code in the Script Editor (top-left), not directly in the Console. This way your analysis is saved, reproducible, and easy to revise.
:::

------------------------------------------------------------------------

## 3. Your First R Commands

Paste the following into the Script Editor, select it, and press `Ctrl + Enter` (Mac: `Cmd + Enter`) to run:

```{r}
# This is a comment — R ignores everything after #
1 + 1

# Assign a value to a variable
x <- 10
y <- 3
x + y
x * y
x / y

# Text (character string)
name <- "EMS Statistics"
print(name)
```

::: callout-note
`<-` is R's assignment operator — it stores the value on the right into the variable on the left. You can also use `=`, but `<-` is the R convention.

Keyboard shortcut: `Alt + -` (Mac: `Option + -`)
:::

------------------------------------------------------------------------

## 4. Installing and Loading Packages

R's power comes from packages. The packages used in this course need to be **installed once**, then **loaded at the start of each session**.

```{r}
#| eval: false

# Install packages (run once per computer)
install.packages("tidyverse")    # Data wrangling and visualization
install.packages("gt")           # APA-format tables
install.packages("psych")        # Descriptive statistics
install.packages("car")          # Levene's test and regression diagnostics
install.packages("effectsize")   # Effect size calculation
install.packages("rstatix")      # Nonparametric tests and post-hoc comparisons
install.packages("pROC")         # ROC curves
install.packages("ResourceSelection") # Hosmer-Lemeshow test
install.packages("mediation")    # Mediation analysis
install.packages("ez")           # Repeated measures ANOVA
install.packages("geepack")      # GEE
install.packages("lavaan")       # SEM
install.packages("semPlot")      # SEM path diagrams
install.packages("lm.beta")      # Standardized regression coefficients
install.packages("corrplot")     # Correlation matrix heatmap
```

```{r}
#| eval: false

# Load packages (run at the start of every R session)
library(tidyverse)
library(gt)
```

::: callout-important
`install.packages()` = download and install (once only)

`library()` = activate for use (every session)

This is the most common source of confusion for beginners.
:::

------------------------------------------------------------------------

## 5. Data Types in R

R has several basic data types that correspond to variable types in statistics:

```{r}
# Numeric (continuous variable)
age           <- 34
response_time <- 8.5

# Character (raw input for categorical variables)
gender <- "Male"
level  <- "EMT-P"

# Factor (correct format for categorical variables in analysis)
level_f <- factor(level, levels = c("EMT-1", "EMT-2", "EMT-P"))

# Logical
is_senior <- age > 40
print(is_senior)   # FALSE

# Check data type
class(age)      # "numeric"
class(gender)   # "character"
class(level_f)  # "factor"
```

::: callout-tip
**Always check before analysis:** Categorical variables (gender, EMT level, district) must be stored as `factor` in R. If left as character strings, many statistical functions will either throw an error or produce incorrect results.
:::

------------------------------------------------------------------------

## 6. Importing the Main Dataset

This course uses a single dataset throughout the entire semester: `ems_main.csv`.

### Steps

1.  Place `ems_main.csv` in a `data/` subfolder inside your project directory
2.  Run the following code:

```{r}
#| eval: false

# Import CSV
ems <- read.csv("data/ems_main.csv",
                fileEncoding = "UTF-8",
                stringsAsFactors = FALSE)

# Check dimensions (should be 200 rows × 27 columns)
dim(ems)

# View first six rows
head(ems)

# Inspect data structure
str(ems)

# Basic summary statistics
summary(ems)
```

### Convert categorical variables to factors

```{r}
#| eval: false

# Do this immediately after importing — required for correct analysis
ems$gender   <- factor(ems$gender,
                        levels = c("Male", "Female"))
ems$level    <- factor(ems$level,
                        levels = c("EMT-1", "EMT-2", "EMT-P"))
ems$district <- factor(ems$district,
                        levels = c("North", "Central",
                                   "South", "East"))
ems$shift    <- factor(ems$shift,
                        levels = c("Day", "Night", "Rotating"))

# Confirm conversion
str(ems)
levels(ems$level)   # Should show: "EMT-1" "EMT-2" "EMT-P"
```

------------------------------------------------------------------------

## 7. Dataset Variable Dictionary

| Variable | Description | Type | Unit / Notes |
|:---|:---|:---|:---|
| `id` | Personnel ID | Character | EMS001–EMS200 |
| `gender` | Sex | Categorical | Male / Female |
| `age` | Age | Continuous | Years |
| `level` | EMT certification level | Categorical | EMT-1 / EMT-2 / EMT-P |
| `years` | Years of service | Continuous | Years |
| `district` | Service district | Categorical | North / Central / South / East |
| `shift` | Shift type | Categorical | Day / Night / Rotating |
| `monthly_cases` | Monthly call volume | Continuous | Calls |
| `night_shifts` | Monthly night shifts | Continuous | Shifts |
| `overtime_hours` | Monthly overtime | Continuous | Hours |
| `bmi` | Body Mass Index | Continuous | kg/m² |
| `sbp` | Systolic blood pressure | Continuous | mmHg |
| `dbp` | Diastolic blood pressure | Continuous | mmHg |
| `sleep_hrs` | Average daily sleep | Continuous | Hours |
| `sleep_qual` | Sleep quality | Continuous | PSQI 0–21; higher = worse |
| `stress_vas` | Subjective stress | Continuous | VAS 0–10 |
| `mbi_ee` | MBI Emotional Exhaustion | Continuous | 0–6 |
| `mbi_dp` | MBI Depersonalization | Continuous | 0–6 |
| `mbi_pa` | MBI Personal Accomplishment | Continuous | 0–6 |
| `job_sat` | Job satisfaction | Continuous | 1–5 |
| `mist_pre` | Pre-training MIST score | Continuous | 0–100 |
| `mist_post` | Post-training MIST score | Continuous | 0–100 |
| `supervisor_score` | Supervisor rating | Continuous | 0–100 |
| `response_time` | On-scene response time | Continuous | Minutes |
| `case_severity` | Case severity | Categorical | 1=Minor / 2=Moderate / 3=Severe / 4=OHCA |
| `ohca_survival` | OHCA survival status | Binary | 0=Death / 1=Survival (OHCA cases only) |
| `occ_injury` | Occupational injury (past year) | Binary | 0=No / 1=Yes |

------------------------------------------------------------------------

## Weekly Assignment

Using `ems_main.csv`, complete the following tasks and submit your R code with output:

**Q1** Run `str(ems)` and describe: what data type is each categorical variable *before* converting to factor?

**Q2** Run `summary(ems)` and find the following values: - Minimum, maximum, and median of `response_time` - Mean of `mbi_ee`

**Q3** Use `table()` to tabulate `level` and `district` separately:

```{r}
#| eval: false
table(ems$level)
table(ems$district)
```

**Q4 Reflection:** The `ohca_survival` variable contains many missing values (NA). Why? Under what circumstances does this variable have meaningful values?
