---
title: "Topic 3：Data Visualization"
subtitle: "ggplot2 Layer Logic, Choosing the Right Chart, APA-Format Output"
execute:
  warning: false
  message: false
  echo: true
---

## Learning Objectives

1. Understand the layer-based logic of ggplot2
2. Select the correct chart type based on variable type
3. Produce publication-ready figures following APA guidelines
4. Use visualization to support and interpret descriptive statistics

---

## Statistical Method Map (This Topic)

| X (Predictor) | Y (Outcome) | Appropriate Chart |
|:-------------|:-----------|:-----------------|
| None | Continuous | Histogram, density plot, boxplot |
| None | Categorical | Bar chart |
| Categorical | Continuous | Grouped boxplot, grouped bar chart |
| Continuous | Continuous | Scatterplot, line plot |

::: {.callout-important}
**Choosing the wrong chart type is one of the most common errors:**

- Distribution of a continuous variable → histogram (not a bar chart)
- Distribution of a categorical variable → bar chart (not a histogram)
- Relationship between two continuous variables → scatterplot (not a line chart)
:::

---

## I. Statistical Theory

### 1.1 Why Is Visualization Part of Statistics?

Anscombe's Quartet is one of the most famous demonstrations in statistics: four datasets share nearly identical means, standard deviations, and correlation coefficients — yet their scatterplots look completely different.

**Numbers can mislead. Graphs rarely do.**

Always pair descriptive statistics with visualization:

- Identical means do not imply identical distributions
- A significant p-value does not imply a large effect
- A high correlation coefficient does not imply a linear relationship

### 1.2 The ggplot2 Layer System

ggplot2 builds plots by stacking layers — like adding successive sheets on a drawing:

```
ggplot(data, aes(mapping))     ← Canvas + axes
  + geom_xxx()                 ← Geometric shape (points, lines, bars)
  + scale_xxx()                ← Axis and color settings
  + labs()                     ← Title and axis labels
  + theme_xxx()                ← Overall visual style
```

Each `+` adds one layer. The canvas (`ggplot()`) must always come first.

### 1.3 Understanding aes()

`aes()` stands for aesthetic mapping — it connects data columns to visual properties:

| aes argument | Visual element |
|:------------|:--------------|
| `x =` | Horizontal axis |
| `y =` | Vertical axis |
| `fill =` | Fill color |
| `color =` | Border or point color |
| `shape =` | Point shape |
| `size =` | Point or line size |

::: {.callout-tip}
**Key rule:**

- Color varies by data → place inside `aes()`
- Fixed color for all elements → place outside `aes()`

```r
# Color varies by level (inside aes)
aes(x = response_time, fill = level)

# All bars the same blue (outside aes)
geom_histogram(fill = "#85B7EB")
```
:::

### 1.4 Choosing the Right Chart

```
What do you want to show?
  ├── Distribution of a single variable
  │     ├── Continuous → Histogram / Density plot
  │     └── Categorical → Bar chart
  ├── Relationship between two variables
  │     ├── Continuous × Continuous → Scatterplot
  │     ├── Categorical × Continuous → Grouped boxplot
  │     └── Categorical × Categorical → Stacked bar chart
  └── Trend over time
        └── Line chart
```

### 1.5 APA Figure Guidelines

::: {.callout-important}
**APA 7th Edition — Key Rules:**

- Figure title appears **above** the figure, in bold: "Figure X. Description"
- Note appears **below** the figure: "*Note.* Text."
- No 3D effects
- Use colorblind-friendly palettes
- Figures must remain interpretable in grayscale
:::

---

## II. R Implementation

### Load packages and data

```{r}
library(tidyverse)
library(gt)

ems <- read.csv("data/ems_main.csv",
                stringsAsFactors = FALSE)

ems$gender   <- factor(ems$gender,
                        levels = c("男", "女"))
ems$level    <- factor(ems$level,
                        levels = c("EMT-1", "EMT-2", "EMT-P"))
ems$district <- factor(ems$district,
                        levels = c("北部", "中部", "南部", "東部"))
ems$shift    <- factor(ems$shift,
                        levels = c("日班為主", "夜班為主", "輪班"))
```

### Figure 1. Histogram — Distribution of a Continuous Variable

```{r}
#| fig-width: 7
#| fig-height: 4.5

ggplot(ems, aes(x = response_time)) +
  geom_histogram(bins = 20,
                 fill = "#85B7EB",
                 color = "white") +
  geom_vline(xintercept = mean(ems$response_time),
             color = "#D85A30",
             linetype = "dashed",
             linewidth = 0.9) +
  geom_vline(xintercept = median(ems$response_time),
             color = "#1D9E75",
             linetype = "dashed",
             linewidth = 0.9) +
  annotate("text",
           x     = mean(ems$response_time) + 1.5,
           y     = 26,
           label = paste0("Mean = ",
                          round(mean(ems$response_time), 1)),
           color = "#D85A30", size = 3.5) +
  annotate("text",
           x     = median(ems$response_time) - 1.5,
           y     = 23,
           label = paste0("Median = ",
                          round(median(ems$response_time), 1)),
           color = "#1D9E75", size = 3.5) +
  labs(
    title   = "Figure 1. Distribution of EMS On-Scene Response Time",
    x       = "Response Time (minutes)",
    y       = "Frequency (n)",
    caption = "Note. Orange dashed line = mean; green dashed line = median. N = 200."
  ) +
  theme_minimal(base_size = 12) +
  theme(
    plot.title   = element_text(face = "bold", size = 12),
    plot.caption = element_text(hjust = 0, size = 9,
                                color = "gray40")
  )
```

### Figure 2. Density Plot — Comparing Distributions Across Groups

```{r}
#| fig-width: 7
#| fig-height: 4.5

ggplot(ems, aes(x = response_time, fill = level)) +
  geom_density(alpha = 0.5) +
  scale_fill_manual(
    values = c("EMT-1" = "#F5C4B3",
               "EMT-2" = "#9FE1CB",
               "EMT-P" = "#85B7EB"),
    name = "EMT Level"
  ) +
  labs(
    title   = "Figure 2. Response Time Density by EMT Certification Level",
    x       = "Response Time (minutes)",
    y       = "Density",
    caption = "Note. Shaded areas represent the distribution shape for each certification level. N = 200."
  ) +
  theme_minimal(base_size = 12) +
  theme(
    plot.title   = element_text(face = "bold", size = 12),
    plot.caption = element_text(hjust = 0, size = 9,
                                color = "gray40")
  )
```

### Figure 3. Boxplot — Group Comparison

```{r}
#| fig-width: 7
#| fig-height: 4.5

ggplot(ems, aes(x = level, y = response_time,
                fill = level)) +
  geom_boxplot(alpha = 0.6,
               outlier.shape = 21,
               outlier.fill  = "white",
               outlier.size  = 2) +
  geom_jitter(width = 0.15, alpha = 0.25, size = 1.2) +
  scale_fill_manual(
    values = c("EMT-1" = "#F5C4B3",
               "EMT-2" = "#9FE1CB",
               "EMT-P" = "#85B7EB")
  ) +
  labs(
    title   = "Figure 3. Response Time by EMT Certification Level",
    x       = "EMT Certification Level",
    y       = "Response Time (minutes)",
    caption = "Note. Center line = median; box = IQR; points = individual observations. N = 200."
  ) +
  theme_minimal(base_size = 12) +
  theme(
    plot.title      = element_text(face = "bold", size = 12),
    legend.position = "none",
    plot.caption    = element_text(hjust = 0, size = 9,
                                   color = "gray40")
  )
```

### Figure 4. Bar Chart — Categorical Variable Distribution

```{r}
#| fig-width: 6
#| fig-height: 4

ems |>
  dplyr::count(level) |>
  dplyr::mutate(pct = round(n / sum(n) * 100, 1)) |>
  ggplot(aes(x = level, y = pct, fill = level)) +
  geom_col(alpha = 0.8, width = 0.6) +
  geom_text(
    aes(label = paste0(pct, "%\n(n=", n, ")")),
    vjust = -0.4, size = 3.5
  ) +
  scale_fill_manual(
    values = c("EMT-1" = "#F5C4B3",
               "EMT-2" = "#9FE1CB",
               "EMT-P" = "#85B7EB")
  ) +
  scale_y_continuous(limits = c(0, 65)) +
  labs(
    title   = "Figure 4. Distribution of EMS Personnel by EMT Level",
    x       = "EMT Certification Level",
    y       = "Percentage (%)",
    caption = "Note. N = 200."
  ) +
  theme_minimal(base_size = 12) +
  theme(
    plot.title      = element_text(face = "bold", size = 12),
    legend.position = "none",
    plot.caption    = element_text(hjust = 0, size = 9,
                                   color = "gray40")
  )
```

### Figure 5. Scatterplot — Relationship Between Two Continuous Variables

```{r}
#| fig-width: 7
#| fig-height: 5

ggplot(ems, aes(x = stress_vas, y = mbi_ee,
                color = level)) +
  geom_point(alpha = 0.5, size = 1.8) +
  geom_smooth(method = "lm", se = TRUE,
              color = "#333333",
              linewidth = 0.8,
              aes(group = 1)) +
  scale_color_manual(
    values = c("EMT-1" = "#D85A30",
               "EMT-2" = "#1D9E75",
               "EMT-P" = "#185FA5"),
    name = "EMT Level"
  ) +
  labs(
    title   = "Figure 5. Relationship Between Perceived Stress and Emotional Exhaustion",
    x       = "Perceived Stress (VAS 0–10)",
    y       = "MBI Emotional Exhaustion (0–6)",
    caption = "Note. Black trend line represents the overall linear fit (with 95% CI). N = 200."
  ) +
  theme_minimal(base_size = 12) +
  theme(
    plot.title   = element_text(face = "bold", size = 12),
    plot.caption = element_text(hjust = 0, size = 9,
                                color = "gray40")
  )
```

### Table 1. Descriptive Statistics by EMT Level (to accompany Figure 3)

```{r}
ems |>
  dplyr::group_by(level) |>
  dplyr::summarise(
    n      = n(),
    Mean   = round(mean(response_time), 2),
    SD     = round(sd(response_time), 2),
    Median = round(median(response_time), 2),
    Q1     = round(quantile(response_time, 0.25), 2),
    Q3     = round(quantile(response_time, 0.75), 2),
    .groups = "drop"
  ) |>
  dplyr::rename(Level = level) |>
  gt() |>
  tab_header(
    title = "Table 1. Response Time Descriptive Statistics by EMT Level"
  ) |>
  tab_spanner(label = "Central Tendency",
              columns = c(Mean, Median)) |>
  tab_spanner(label = "Quartiles",
              columns = c(Q1, Q3)) |>
  tab_source_note(
    source_note = "Note. Units: minutes. Q1 = 25th percentile; Q3 = 75th percentile."
  ) |>
  cols_align(align = "center", columns = -Level) |>
  cols_align(align = "left",   columns = Level) |>
  tab_style(
    style     = cell_text(weight = "bold"),
    locations = cells_column_labels()
  ) |>
  tab_style(
    style     = cell_text(weight = "bold"),
    locations = cells_column_spanners()
  ) |>
  tab_options(
    table.border.top.style            = "hidden",
    heading.border.bottom.style       = "solid",
    column_labels.border.top.style    = "solid",
    column_labels.border.bottom.style = "solid",
    table.border.bottom.style         = "solid",
    table.font.size                   = px(13),
    heading.title.font.size           = px(14),
    heading.align                     = "left"
  )
```

### Saving High-Resolution Figures

```{r}
#| eval: false

# Create output folder
dir.create("figures", showWarnings = FALSE)

# Save at 300 dpi for publication submission
ggsave("figures/fig3_boxplot.png",
       width = 7, height = 4.5, dpi = 300)

# TIFF format (required by some journals)
ggsave("figures/fig3_boxplot.tiff",
       width = 7, height = 4.5, dpi = 300,
       compression = "lzw")
```

---

## III. Reporting Results (APA Format)

**Referencing a figure:**

> Figure 2 shows that EMT-P personnel have a notably left-shifted response time distribution (*M* = 7.8 min) compared to EMT-1 (*M* = 10.4 min), with only partial overlap between the two distributions, suggesting that certification level may influence response efficiency (see Topic 7 ANOVA for formal testing).

**Referencing a table alongside a figure:**

> As shown in Table 1, the median response time decreased progressively from EMT-1 (9.4 min) to EMT-2 (8.1 min) to EMT-P (6.9 min), indicating a clear gradient across certification levels.

---

## Weekly Assignment

**Q1** Plot a histogram of `stress_vas`, marking both the mean and median with dashed lines. Describe the shape of the distribution (normal? skewed?). Follow APA formatting.

**Q2** Create a grouped boxplot comparing `monthly_cases` across the four districts (`district`). Accompany it with an APA-format `gt` descriptive statistics table.

**Q3** Create a bar chart showing the proportion of workers with occupational injury (`occ_injury`) within each shift type (`shift`). Label each bar with the percentage.

**Q4** Create a scatterplot exploring the relationship between `sleep_hrs` and `mbi_ee`, with a linear trend line. Color points by `gender`.

**Q5 Reflection:** In Figure 2, the density plot shows that the EMT-P response time distribution overlaps partially with EMT-1. What does this imply for the inference that "certification level affects response time"? How will Topic 7 (ANOVA) formally address this question?