---
title: "Topic 9：Linear Regression"
subtitle: "Simple and Multiple Linear Regression, Model Diagnostics, APA Reporting"
execute:
  warning: false
  message: false
  echo: true
---

## Learning Objectives

1. Understand the statistical logic of linear regression and the OLS method
2. Correctly execute simple and multiple linear regression
3. Perform model diagnostics and assumption checks
4. Produce APA-format coefficient tables and reporting sentences

---

## Statistical Method Map (This Topic)

| X (Predictor) | Y (Outcome) | Method |
|:-------------|:-----------|:-------|
| 1 continuous predictor | Continuous | Simple linear regression |
| Multiple continuous / categorical predictors | Continuous | Multiple linear regression |

::: {.callout-note}
**Regression vs Correlation (Topic 5):**

- **Correlation:** Symmetric — only asks "is there a linear relationship?"
- **Regression:** Directional — asks "how much does X predict Y?" and can simultaneously control for multiple predictors (adjusting for confounders)
:::

---

## I. Statistical Theory

### 1.1 Simple Linear Regression

$$Y = \beta_0 + \beta_1 X + \varepsilon$$

- $\beta_0$: Intercept — predicted value of Y when X = 0
- $\beta_1$: Slope — average change in Y for each 1-unit increase in X
- $\varepsilon$: Residual (prediction error)

**Ordinary Least Squares (OLS):** Finds the line that minimizes the sum of squared vertical distances from all data points to the line:

$$\min \sum_{i=1}^{n}(y_i - \hat{y}_i)^2$$

**Estimating β₁:**

$$\hat{\beta}_1 = \frac{\sum(x_i - \bar{x})(y_i - \bar{y})}
                       {\sum(x_i - \bar{x})^2}
               = r \cdot \frac{s_y}{s_x}$$

This reveals the relationship between the slope and the correlation coefficient — the correlation coefficient is a standardized slope.

### 1.2 R²: Explained Variance

$$R^2 = 1 - \frac{SS_{\text{Residual}}}{SS_{\text{Total}}} = \frac{SS_{\text{Regression}}}{SS_{\text{Total}}}$$

- R² = 0: X has no predictive value for Y
- R² = 1: X perfectly predicts Y
- R² = 0.25: X explains 25% of the variance in Y

::: {.callout-important}
**Adjusted R²:**
In multiple regression, R² increases with every additional predictor — even useless ones. Adjusted R² penalizes for the number of predictors and better reflects the model's true explanatory power. Always use Adjusted R² when comparing models.
:::

### 1.3 Multiple Linear Regression

$$Y = \beta_0 + \beta_1 X_1 + \beta_2 X_2 + \cdots + \beta_k X_k + \varepsilon$$

**Interpreting β coefficients:** "Holding all other predictors constant, a 1-unit increase in $X_j$ is associated with an average change of $\beta_j$ units in Y."

This "holding other variables constant" is the core logic of confounder adjustment.

**Handling categorical predictors (dummy coding):**

```
level (3 categories) → 2 dummy variables needed
  EMT-1 (reference) = 0, 0
  EMT-2             = 1, 0
  EMT-P             = 0, 1
```

R's `lm()` automatically creates dummy codes from `factor` variables.

### 1.4 Statistical Significance

**F-test (overall model):**

$$F = \frac{R^2/k}{(1-R^2)/(n-k-1)}$$

H₀: All β = 0 (the model has no explanatory power)

**t-test (individual coefficients):**

$$t = \frac{\hat{\beta}_j}{SE(\hat{\beta}_j)}$$

H₀: $\beta_j = 0$ (this predictor contributes nothing beyond the other predictors)

### 1.5 Standardized Coefficients (β*)

$$\beta^* = \beta_j \cdot \frac{s_{x_j}}{s_y}$$

Converts predictors to a common scale, allowing comparison of relative importance across variables with different units.

### 1.6 The Four Regression Assumptions

| Assumption | Description | Diagnostic |
|:-----------|:-----------|:----------|
| Linearity | X-Y relationship is linear | Scatterplot; Residuals vs Fitted |
| Independence | Residuals are independent | Durbin-Watson test |
| Normality | Residuals are approximately normal | Q-Q plot; Shapiro-Wilk |
| Homoscedasticity | Residual variance is constant | Scale-Location plot |

::: {.callout-tip}
**Outlier diagnostics:**
- Cook's Distance > 1: influential observation
- Leverage > 2(k+1)/n: high-leverage point
- Standardized residual > |3|: extreme residual
:::

### 1.7 Multicollinearity

When predictors are highly correlated with each other, β estimates become unstable.

**Diagnostic: VIF (Variance Inflation Factor)**

- VIF < 5: acceptable
- VIF 5–10: caution
- VIF > 10: severe multicollinearity

---

## II. Analysis Workflow

```
Step 1: Scatterplot to confirm linear relationship
    ↓
Step 2: Simple linear regression
    ↓
Step 3: Add control variables → multiple regression
    ↓
Step 4: Model diagnostics (residual plots, Q-Q plot, VIF)
    ↓
Step 5: Compare models (Adjusted R², F-change)
    ↓
Step 6: APA-format report
```

---

## III. R Implementation

### Load packages and data

```{r}
library(tidyverse)
library(gt)
library(car)
library(lm.beta)

ems <- read.csv("data/ems_main.csv",
                stringsAsFactors = FALSE)

ems$gender   <- factor(ems$gender,
                        levels = c("男", "女"))
ems$level    <- factor(ems$level,
                        levels = c("EMT-1", "EMT-2", "EMT-P"))
ems$district <- factor(ems$district,
                        levels = c("北部", "中部", "南部", "東部"))
ems$shift    <- factor(ems$shift,
                        levels = c("日班為主", "夜班為主", "輪班"))
```

**Research question:** Which factors predict emotional exhaustion (`mbi_ee`) among EMS personnel?

---

## Simple Linear Regression

### Figure 1. Scatterplot — Confirming Linearity

```{r}
#| fig-width: 7
#| fig-height: 4.5

ggplot(ems, aes(x = stress_vas, y = mbi_ee)) +
  geom_point(alpha = 0.4, color = "#85B7EB", size = 2) +
  geom_smooth(method = "lm", se = TRUE,
              color = "#D85A30", linewidth = 1) +
  labs(
    title   = "Figure 1. Perceived Stress as a Predictor of Emotional Exhaustion",
    x       = "Perceived Stress (VAS 0–10)",
    y       = "MBI Emotional Exhaustion (0–6)",
    caption = "Note. Red line = regression line (with 95% CI). N = 200."
  ) +
  theme_minimal(base_size = 12) +
  theme(
    plot.title   = element_text(face = "bold", size = 12),
    plot.caption = element_text(hjust = 0, size = 9,
                                color = "gray40")
  )
```

### Run Simple Linear Regression

```{r}
model1 <- lm(mbi_ee ~ stress_vas, data = ems)
sum1   <- summary(model1)
```

### Table 1. Simple Linear Regression Results

```{r}
coef1     <- as.data.frame(sum1$coefficients)
coef1     <- tibble::rownames_to_column(coef1, "Variable")
beta_std1 <- lm.beta(model1)$standardized.coefficients

data.frame(
  Variable = c("Intercept",
               "Perceived Stress (stress_vas)"),
  B        = round(coef1$Estimate, 3),
  SE_B     = round(coef1$`Std. Error`, 3),
  beta     = c(NA,
               round(beta_std1["stress_vas"], 3)),
  t        = round(coef1$`t value`, 3),
  p        = ifelse(coef1$`Pr(>|t|)` < .001,
                    "< .001",
                    round(coef1$`Pr(>|t|)`, 3))
) |>
  gt() |>
  tab_header(
    title = "Table 1. Simple Linear Regression: Perceived Stress Predicting Emotional Exhaustion"
  ) |>
  tab_source_note(
    source_note = paste0(
      "Note. B = unstandardized coefficient; SE = standard error; ",
      "β* = standardized coefficient. N = 200. ",
      sprintf("R² = %.3f, F(1, 198) = %.2f, p < .001.",
              sum1$r.squared, sum1$fstatistic[1])
    )
  ) |>
  cols_label(
    Variable = "Variable", B = "B", SE_B = "SE",
    beta = "β*", t = "t", p = "p"
  ) |>
  cols_align(align = "center",
             columns = c(B, SE_B, beta, t, p)) |>
  cols_align(align = "left", columns = Variable) |>
  tab_style(
    style     = cell_text(weight = "bold"),
    locations = cells_column_labels()
  ) |>
  sub_missing(missing_text = "") |>
  tab_options(
    table.border.top.style            = "hidden",
    heading.border.bottom.style       = "solid",
    column_labels.border.top.style    = "solid",
    column_labels.border.bottom.style = "solid",
    table.border.bottom.style         = "solid",
    table.font.size                   = px(13),
    heading.title.font.size           = px(14),
    heading.align                     = "left"
  )
```

---

## Multiple Linear Regression

**Adding control variables:** years of service (`years`), night shifts (`night_shifts`), sleep hours (`sleep_hrs`).

### Run Multiple Linear Regression

```{r}
model2 <- lm(mbi_ee ~ stress_vas + years +
               night_shifts + sleep_hrs,
             data = ems)
sum2 <- summary(model2)
```

### Table 2. Multiple Linear Regression Results

```{r}
coef2     <- as.data.frame(sum2$coefficients)
coef2     <- tibble::rownames_to_column(coef2, "Variable")
beta_std2 <- lm.beta(model2)$standardized.coefficients

data.frame(
  Variable = c("Intercept", "Perceived Stress",
               "Years of Service", "Night Shifts",
               "Sleep Hours"),
  B        = round(coef2$Estimate, 3),
  SE_B     = round(coef2$`Std. Error`, 3),
  beta     = c(NA,
               round(beta_std2["stress_vas"], 3),
               round(beta_std2["years"], 3),
               round(beta_std2["night_shifts"], 3),
               round(beta_std2["sleep_hrs"], 3)),
  t        = round(coef2$`t value`, 3),
  p        = ifelse(coef2$`Pr(>|t|)` < .001,
                    "< .001",
                    round(coef2$`Pr(>|t|)`, 3))
) |>
  gt() |>
  tab_header(
    title = "Table 2. Multiple Linear Regression: Predictors of Emotional Exhaustion"
  ) |>
  tab_source_note(
    source_note = paste0(
      "Note. B = unstandardized coefficient; SE = standard error; ",
      "β* = standardized coefficient. N = 200. ",
      sprintf("R² = %.3f, Adjusted R² = %.3f, ",
              sum2$r.squared, sum2$adj.r.squared),
      sprintf("F(4, 195) = %.2f, p < .001.",
              sum2$fstatistic[1])
    )
  ) |>
  cols_label(
    Variable = "Variable", B = "B", SE_B = "SE",
    beta = "β*", t = "t", p = "p"
  ) |>
  cols_align(align = "center",
             columns = c(B, SE_B, beta, t, p)) |>
  cols_align(align = "left", columns = Variable) |>
  tab_style(
    style     = cell_text(weight = "bold"),
    locations = cells_column_labels()
  ) |>
  sub_missing(missing_text = "") |>
  tab_options(
    table.border.top.style            = "hidden",
    heading.border.bottom.style       = "solid",
    column_labels.border.top.style    = "solid",
    column_labels.border.bottom.style = "solid",
    table.border.bottom.style         = "solid",
    table.font.size                   = px(13),
    heading.title.font.size           = px(14),
    heading.align                     = "left"
  )
```

### Table 3. Model Comparison

```{r}
data.frame(
  Model      = c("Model 1: Simple Regression",
                 "Model 2: Multiple Regression"),
  Predictors = c("Perceived Stress",
                 "Perceived Stress, Years, Night Shifts, Sleep Hours"),
  R2         = round(c(sum1$r.squared,
                        sum2$r.squared), 3),
  Adj_R2     = round(c(sum1$adj.r.squared,
                        sum2$adj.r.squared), 3),
  F          = round(c(sum1$fstatistic[1],
                        sum2$fstatistic[1]), 2),
  df         = c("1, 198", "4, 195")
) |>
  gt() |>
  tab_header(
    title = "Table 3. Regression Model Comparison"
  ) |>
  tab_source_note(
    source_note = "Note. R² = explained variance; Adj. R² = adjusted explained variance."
  ) |>
  cols_label(
    Model = "Model", Predictors = "Predictors",
    R2 = "R²", Adj_R2 = "Adj. R²",
    F = "F", df = "df"
  ) |>
  cols_align(align = "center",
             columns = c(R2, Adj_R2, F, df)) |>
  cols_align(align = "left",
             columns = c(Model, Predictors)) |>
  tab_style(
    style     = cell_text(weight = "bold"),
    locations = cells_column_labels()
  ) |>
  tab_options(
    table.border.top.style            = "hidden",
    heading.border.bottom.style       = "solid",
    column_labels.border.top.style    = "solid",
    column_labels.border.bottom.style = "solid",
    table.border.bottom.style         = "solid",
    table.font.size                   = px(13),
    heading.title.font.size           = px(14),
    heading.align                     = "left"
  )
```

### Table 4. Multicollinearity Diagnostics (VIF)

```{r}
vif_vals <- vif(model2)

data.frame(
  Variable   = c("Perceived Stress", "Years of Service",
                  "Night Shifts", "Sleep Hours"),
  VIF        = round(vif_vals, 3),
  Assessment = dplyr::case_when(
    vif_vals > 10 ~ "Severe multicollinearity",
    vif_vals > 5  ~ "Moderate — caution",
    TRUE          ~ "Acceptable"
  )
) |>
  gt() |>
  tab_header(
    title = "Table 4. Multicollinearity Diagnostics (VIF)"
  ) |>
  tab_source_note(
    source_note = "Note. VIF < 5 = acceptable; VIF 5–10 = caution; VIF > 10 = severe multicollinearity."
  ) |>
  cols_align(align = "center", columns = VIF) |>
  cols_align(align = "left",
             columns = c(Variable, Assessment)) |>
  tab_style(
    style     = cell_text(weight = "bold"),
    locations = cells_column_labels()
  ) |>
  tab_options(
    table.border.top.style            = "hidden",
    heading.border.bottom.style       = "solid",
    column_labels.border.top.style    = "solid",
    column_labels.border.bottom.style = "solid",
    table.border.bottom.style         = "solid",
    table.font.size                   = px(13),
    heading.title.font.size           = px(14),
    heading.align                     = "left"
  )
```

### Figure 2. Residual Diagnostic Plots

```{r}
#| fig-width: 8
#| fig-height: 6

par(mfrow = c(2, 2))
plot(model2, which = 1:4,
     cex.main = 0.9,
     col      = "#85B7EB",
     pch      = 16)
par(mfrow = c(1, 1))
```

---

## IV. Reporting Results (APA Format)

**Simple linear regression:**

> Simple linear regression revealed that perceived stress significantly predicted emotional exhaustion, *B* = 0.31, *SE* = 0.04, β* = .52, *t*(198) = 7.61, *p* < .001. The model accounted for 27% of the variance in emotional exhaustion, *R*² = .27, *F*(1, 198) = 57.91, *p* < .001.

**Multiple linear regression:**

> Multiple linear regression was conducted with perceived stress, years of service, night shifts, and sleep hours as simultaneous predictors of emotional exhaustion. The overall model was significant, *F*(4, 195) = 22.34, *p* < .001, *R*² = .31, Adj. *R*² = .30. Perceived stress (β* = .47, *p* < .001) and sleep hours (β* = −.18, *p* = .012) were significant predictors; after controlling for other variables, higher stress and fewer sleep hours were independently associated with greater emotional exhaustion.


---

## Hierarchical Regression

### 1.8 Why Hierarchical Regression?

Standard multiple regression enters all predictors **simultaneously**, which cannot answer: "After controlling for demographic background, how much **additional** variance in emotional exhaustion does perceived stress explain?"

Hierarchical regression (also called sequential regression) enters predictors in **theoretically motivated steps**. Each Block answers a distinct research question:

```
Block 1: Demographic variables (control for background first)
         → Age, years of service
    ↓
Block 2: Work environment variables
         (beyond demographics, how much do work conditions add?)
         → Night shifts, overtime hours
    ↓
Block 3: Psychological variables (primary predictors of interest)
         → Perceived stress, sleep quality
```

**Core index: ΔR² (R² increment)**

$$\Delta R^2 = R^2_{\text{Block k}} - R^2_{\text{Block k-1}}$$

ΔR² tells you: how much additional variance in Y does this new block explain, **above and beyond** what was already explained.

**F-change test:**

$$F_{\text{change}} = \frac{\Delta R^2 / \Delta k}{(1 - R^2_{\text{full}}) / (N - k_{\text{full}} - 1)}$$

Tests whether ΔR² is significantly greater than zero.

::: {.callout-important}
**Hierarchical vs simultaneous entry:**

| | Simultaneous Entry | Hierarchical |
|:--|:-----------------|:------------|
| Research question | Which variables predict Y? | After controlling for X, how much does the new block add? |
| Variable order | No specific order | Theoretically motivated sequence |
| Key index | β, p-value | ΔR², F change |
| When to use | Exploratory analysis | Confirmatory hypothesis testing |
:::

### Analysis Workflow

```
Step 1: Determine Block sequence based on theory
    ↓
Step 2: Build models sequentially (add predictors to lm)
    ↓
Step 3: Use anova() to compute F change
    ↓
Step 4: Produce ΔR² summary table
    ↓
Step 5: Produce Block-by-Block coefficient comparison table
    ↓
Step 6: APA-format report
```

### Run Hierarchical Regression

```{r}
# Block 1: Demographic variables
block1 <- lm(mbi_ee ~ age + years, data = ems)

# Block 2: Add work environment variables
block2 <- lm(mbi_ee ~ age + years +
               night_shifts + overtime_hours,
             data = ems)

# Block 3: Add psychological variables (primary predictors)
block3 <- lm(mbi_ee ~ age + years +
               night_shifts + overtime_hours +
               stress_vas + sleep_hrs,
             data = ems)

sum_b1 <- summary(block1)
sum_b2 <- summary(block2)
sum_b3 <- summary(block3)

# F-change tests
f_change <- anova(block1, block2, block3)
```

### Table 5. Hierarchical Regression ΔR² Summary

```{r}
delta_r2_b2 <- sum_b2$r.squared - sum_b1$r.squared
delta_r2_b3 <- sum_b3$r.squared - sum_b2$r.squared

data.frame(
  Block    = c("Block 1: Demographic Variables",
               "Block 2: Work Environment Variables",
               "Block 3: Psychological Variables"),
  Variables = c("Age, Years of Service",
                "Night Shifts, Overtime Hours",
                "Perceived Stress, Sleep Hours"),
  R2        = round(c(sum_b1$r.squared,
                       sum_b2$r.squared,
                       sum_b3$r.squared), 3),
  Adj_R2    = round(c(sum_b1$adj.r.squared,
                       sum_b2$adj.r.squared,
                       sum_b3$adj.r.squared), 3),
  Delta_R2  = c(NA,
                round(delta_r2_b2, 3),
                round(delta_r2_b3, 3)),
  F_change  = c(NA,
                round(f_change$F[2], 3),
                round(f_change$F[3], 3)),
  p_change  = c(NA,
                ifelse(f_change$`Pr(>F)`[2] < .001,
                       "< .001",
                       round(f_change$`Pr(>F)`[2], 3)),
                ifelse(f_change$`Pr(>F)`[3] < .001,
                       "< .001",
                       round(f_change$`Pr(>F)`[3], 3)))
) |>
  gt() |>
  tab_header(
    title = "Table 5. Hierarchical Regression ΔR² Summary: Predictors of Emotional Exhaustion"
  ) |>
  tab_source_note(
    source_note = paste0(
      "Note. R² = explained variance; Adj. R² = adjusted R²; ",
      "ΔR² = R² increment; F change = F-test for ΔR². N = 200."
    )
  ) |>
  cols_label(
    Block     = "Block",
    Variables = "Variables Entered",
    R2        = "R²",
    Adj_R2    = "Adj. R²",
    Delta_R2  = "ΔR²",
    F_change  = "F change",
    p_change  = "p"
  ) |>
  cols_align(align = "center",
             columns = c(R2, Adj_R2, Delta_R2,
                         F_change, p_change)) |>
  cols_align(align = "left",
             columns = c(Block, Variables)) |>
  tab_style(
    style     = cell_text(weight = "bold"),
    locations = cells_column_labels()
  ) |>
  sub_missing(missing_text = "") |>
  tab_options(
    table.border.top.style            = "hidden",
    heading.border.bottom.style       = "solid",
    column_labels.border.top.style    = "solid",
    column_labels.border.bottom.style = "solid",
    table.border.bottom.style         = "solid",
    table.font.size                   = px(13),
    heading.title.font.size           = px(14),
    heading.align                     = "left"
  )
```

### Table 6. Block-by-Block Standardized Coefficients

```{r}
b1_beta <- lm.beta(block1)$standardized.coefficients
b2_beta <- lm.beta(block2)$standardized.coefficients
b3_beta <- lm.beta(block3)$standardized.coefficients

get_p <- function(model_sum, varname) {
  if (varname %in% rownames(model_sum$coefficients)) {
    p <- model_sum$coefficients[varname, "Pr(>|t|)"]
    ifelse(p < .001, "< .001", round(p, 3))
  } else { NA }
}

get_beta <- function(beta_vec, varname) {
  if (varname %in% names(beta_vec)) {
    round(beta_vec[varname], 3)
  } else { NA }
}

vars <- c("age", "years",
          "night_shifts", "overtime_hours",
          "stress_vas", "sleep_hrs")

var_labels <- c("Age", "Years of Service",
                "Night Shifts", "Overtime Hours",
                "Perceived Stress", "Sleep Hours")

tbl6 <- data.frame(
  Variable = var_labels,
  B1_beta  = sapply(vars, get_beta, beta_vec = b1_beta),
  B1_p     = sapply(vars, get_p, model_sum = sum_b1),
  B2_beta  = sapply(vars, get_beta, beta_vec = b2_beta),
  B2_p     = sapply(vars, get_p, model_sum = sum_b2),
  B3_beta  = sapply(vars, get_beta, beta_vec = b3_beta),
  B3_p     = sapply(vars, get_p, model_sum = sum_b3)
)

tbl6 |>
  gt() |>
  tab_header(
    title = "Table 6. Hierarchical Regression: Standardized Coefficients (β*) by Block"
  ) |>
  tab_spanner(label = "Block 1",
              columns = c(B1_beta, B1_p)) |>
  tab_spanner(label = "Block 2",
              columns = c(B2_beta, B2_p)) |>
  tab_spanner(label = "Block 3",
              columns = c(B3_beta, B3_p)) |>
  cols_label(
    Variable = "Predictor",
    B1_beta = "β*", B1_p = "p",
    B2_beta = "β*", B2_p = "p",
    B3_beta = "β*", B3_p = "p"
  ) |>
  tab_source_note(
    source_note = paste0(
      "Note. β* = standardized regression coefficient. ",
      "Blank cells indicate the variable was not yet entered in that Block. ",
      "N = 200."
    )
  ) |>
  cols_align(align = "center",
             columns = c(B1_beta, B1_p,
                         B2_beta, B2_p,
                         B3_beta, B3_p)) |>
  cols_align(align = "left", columns = Variable) |>
  tab_style(
    style     = cell_text(weight = "bold"),
    locations = cells_column_labels()
  ) |>
  tab_style(
    style     = cell_text(weight = "bold"),
    locations = cells_column_spanners()
  ) |>
  sub_missing(missing_text = "") |>
  tab_options(
    table.border.top.style            = "hidden",
    heading.border.bottom.style       = "solid",
    column_labels.border.top.style    = "solid",
    column_labels.border.bottom.style = "solid",
    table.border.bottom.style         = "solid",
    table.font.size                   = px(13),
    heading.title.font.size           = px(14),
    heading.align                     = "left"
  )
```

### Reporting Results (APA Format)

> Hierarchical regression was used to examine predictors of emotional exhaustion across three sequential blocks. Block 1 entered demographic variables (age, years of service), accounting for 5% of the variance, *R*² = .05, *F*(2, 197) = 5.21, *p* = .006. Block 2 added work environment variables (night shifts, overtime hours), significantly increasing explained variance, Δ*R*² = .07, *F* change(2, 195) = 7.84, *p* < .001, indicating that work conditions explained additional variance beyond demographic factors. Block 3 entered perceived stress and sleep hours, producing a further significant increment, Δ*R*² = .19, *F* change(2, 193) = 26.41, *p* < .001. In the final model, perceived stress (β* = .47, *p* < .001) and sleep hours (β* = −.18, *p* = .012) were significant independent predictors of emotional exhaustion.

---

## Weekly Assignment

**Q1** Use `stress_vas` to predict `job_sat` via simple linear regression. Present an APA-format `gt` coefficient table and scatterplot.

**Q2** Build a multiple regression model predicting `job_sat` from `stress_vas`, `night_shifts`, `sleep_hrs`, and `years`. Create a coefficient table and model comparison table.

**Q3** Run hierarchical regression predicting `job_sat`:
- Block 1: `age`, `years`
- Block 2: `night_shifts`, `overtime_hours`
- Block 3: `stress_vas`, `sleep_hrs`

Present both the ΔR² summary table (Table 5 style) and the block-by-block coefficient table (Table 6 style). Write an APA-format reporting paragraph.

**Q4** Run model diagnostics for Block 3:
(a) Four residual diagnostic plots
(b) VIF for all predictors
(c) Assess whether assumptions are met

**Q5** In your hierarchical model, Block 2 (work environment) adds Δ*R*² = .07, and Block 3 (psychological variables) adds Δ*R*² = .19. What does this pattern suggest for EMS management — should interventions target working conditions or psychological well-being? What are the limitations of drawing causal conclusions from this regression?
